---
title: "Inter-Annotator Agreement"
format:
  html:
    toc: true
knitr:
  opts_chunk: 
    collapse: false
editor: visual
---

## Packages

```{r}
#| output: false
#| message: false

library(tidyverse)
library(readxl)
library(tidycomm)
library(irr)
library(psy)
library(writexl)
library(ggbeeswarm)

```

## List files

```{r}
#| output: false
#| message: false

f_stefan <- list.files("../Coders/Sheets for Stefan, Sveta and Michael/", pattern = "^SH", full.names = T)
f_sveta <- list.files("../Coders/Sheets for Stefan, Sveta and Michael/", pattern = "^SK", full.names = T)
f_michael <- list.files("../Coders/Sheets for Stefan, Sveta and Michael/", pattern = "^MP", full.names = T)
f_f <- list.files("../Coders/Sheets for coders/", pattern = "^F", full.names = T) # coder F


# dataframe with coder (= crowdsourced participant), annotator (= person who rated the crowdsourced participant's answer), task number. Each spreadsheet contains data from multiple coders, but the rating of the answers was conducted by one individual.
files <- tibble ( file = c(f_stefan, f_sveta, f_michael, f_f))
files <- separate_wider_delim(files, file, delim = "TASK", names = c("file_rest", "task"), cols_remove = F)
files <- separate_wider_delim(files, task, "_", names = c("task", "coders"), too_many = "merge")
files$annotator <- gsub(".*//", "", files$file_rest)
files <- select(files, -file_rest)

# When exporting from Googledocs, semicolons in the file name get lost. As they are in some cases vital for reconstructing the coder IDs from the file name, we add a column with the original filenames.
orig_fn1 <- readLines("helpers/filenames_stefan_sveta_michael.txt")
orig_fn2 <- readLines("helpers/filenames_coders.txt")

orig_fn <- tibble(orig_file = c(orig_fn1, orig_fn2))
orig_fn <- separate_wider_delim(orig_fn, orig_file, delim = "TASK", names = c("file_rest", "task"), cols_remove = F)
orig_fn <- separate_wider_delim(orig_fn, task, ":", names = c("task", "coders"), too_many = "merge")
colnames(orig_fn)[1] <- "annotator"
colnames(orig_fn)[3] <- "coders_orig"

# remove whitespace from task numbers and annotators
orig_fn$task <- trimws(orig_fn$task)
files$task <- trimws(files$task)

orig_fn$annotator <- trimws(orig_fn$annotator)
files$annotator <- trimws(files$annotator)

# join both
files <- left_join(files, orig_fn, by = c("task", "annotator"))


# find overlapping task numbers (ignore warning for now)
files01 <- files %>% arrange(as.numeric(task)) %>% select(task, file) %>% pivot_wider(names_from = task, values_from = file)

# find all with multiple coders
files01 <- files01[,which(sapply(1:length(files01), function(i) length(unlist(files01[,i])))>1)]


```

## Function for reading in Excel sheets

This function reads in the Excel sheet and then returns a dataframe with either the vocalizations or the ape gestures (where two sheets with apes are available in one workbook, both are combined).

```{r}

# function for adding participant prefix to column name
add_participant_prefix <- function(x) {
  x <- unlist(strsplit(x, split = " "))
  x <- paste0(x[2], "_", x[1])
  return(x)
}

# function for replacing non-numeric ratings

replace_non_numerics <- function(df) {
  
  # find relevant columns
  find_ratings <- grep("Rating", colnames(df))

for(i in 1:length(find_ratings)) {
  if(!is.numeric(unlist(df[,find_ratings[i]]))) {
    df[,find_ratings[i]] <- suppressWarnings(as.numeric(unlist(df[,find_ratings[i]])))
  }
}
  
  return(df)
}




read_excel_custom <- function(path, type = c("voc", "apes")) {
  
  # list of sheets to determine how many
  # vocalization and/or ape sheets there are
  my_sheets <- excel_sheets(path)
  
  
  # get coders from origoinal filename
  coder_ids <- files[which(files$file==path),]$coders_orig
  coder_ids <- unlist(strsplit(trimws(coder_ids), "; ?|, ?| "))
  
  
  if(any(nchar(coder_ids)<6)) {
    small_id <- which(nchar(coder_ids) < 6)
    coder_ids <- coder_ids[-small_id]
  }
  

  
  #coder_ids <- gsub(".xlsx", "", unlist(strsplit(coder_ids, "TASK.*?_"))[2])
  # unlist(strsplit(trimws(coder_ids), "; ?|, ?| +|(?<=v)_(?! )", perl = T))
  #coder_ids <- unlist(strsplit(trimws(coder_ids), ", ?| +|(?<=v)_(?! )", perl = T))
  
  # coder_ids <- unlist(strsplit(trimws(coder_ids), ", ?| +|(?<=v)_", perl = T))
  
  if(type == "voc") {
    voc <- read_excel(path, sheet = "Vocalizations")
    
    # replace non-numeric items
    voc <- replace_non_numerics(voc)
    
    # remove additional columns if there are any
    remove_cols <- grep("\\.\\.\\.", colnames(voc))
    
    if(length(remove_cols) > 0) {
        voc <- voc[,-remove_cols]
      }
    
    # replace column names for better pivoting
    colnames(voc) <- gsub("Participant ", "Answer P", colnames(voc))
    
 for(i in 1:length(colnames(voc))) {
   if(grepl("P.$", colnames(voc)[i])) {
     colnames(voc)[i] <- add_participant_prefix(colnames(voc)[i])
   }
 }
    
    voc <- voc %>% pivot_longer(cols = c(3:length(voc)), names_to = c("Participant", ".value"), names_sep = "_")
 
    
    # add Coder IDs
    n_c <- length(levels(factor(voc$Participant))) # number of coders
    voc <- left_join(voc, tibble(Participant = paste0("P", 1:n_c),
           Coder_ID = coder_ids[1:n_c]))
    
    
    
    # check if there is another vocalizations sheet
    # (if there are two vocalizations sheets, there are no
    # ape sheets! This is good, otherwise it would mess up the coder IDs)
    
    if(any(grepl("Vocalizations \\(2\\)", my_sheets))) {
      
      
        voc2 <- read_excel(path, sheet = "Vocalizations (2)")
        
        # remove additional columns if there are any
    remove_cols <- grep("\\.\\.\\.", colnames(voc2))
    
    if(length(remove_cols) > 0) {
        voc2 <- voc2[,-remove_cols]
      }
        
            # replace non-numeric items
    voc2 <- replace_non_numerics(voc2)
        
    
    # replace column names for better pivoting
    colnames(voc2) <- gsub("Participant ", "Answer P", colnames(voc2))
    
 for(i in 1:length(colnames(voc2))) {
   if(grepl("P.$", colnames(voc2)[i])) {
     colnames(voc2)[i] <- add_participant_prefix(colnames(voc2)[i])
   }
 }
    
    voc2 <- voc2 %>% pivot_longer(cols = c(3:length(voc2)), names_to = c("Participant", ".value"), names_sep = "_")
 
    
    # add Coder IDs
    n_c <- length(levels(factor(voc2$Participant))) # number of coders
    
    # remove coder IDs from Vocalizations tab
    remove_n <- length(levels(factor(voc$Coder_ID)))
    coder_ids_voc2 <- coder_ids[(remove_n+1):length(coder_ids)]
    
    # join with coder IDs
    voc2 <- left_join(voc2, tibble(Participant = paste0("P", 1:n_c),
           Coder_ID = coder_ids_voc2[1:n_c]))
    
    # combine with voc
    voc <- rbind(voc, voc2)
    
    }
    
    # remove Participant number to avoid confusion
    voc <- select(voc, -Participant)
    return(voc)
    
    
  } else {
    
    if(any(grepl("Ape 2", my_sheets))) {
      apes01 <- read_excel(path, sheet = "Ape 1")
      
          # replace non-numeric items
    apes01 <- replace_non_numerics(apes01)
      
      
      # remove additional columns if there are any
      remove_cols <- grep("\\.\\.\\.", colnames(apes01))
      
      if(length(remove_cols) > 0) {
        apes01 <- apes01[,-remove_cols]
      }
      
      # make sure column names are consistent
    colnames(apes01) <- gsub("p4", "P4", colnames(apes01))
      
      # change column names for better pivoting
      colnames(apes01) <- gsub("Participant ", "Answer P", colnames(apes01))
      
      for(i in 1:length(colnames(apes01))) {
        if(grepl("P.$", colnames(apes01)[i])) {
          colnames(apes01)[i] <- add_participant_prefix(colnames(apes01)[i])
        }
      }
      
    # pivot to longer format
      apes01 <- apes01 %>% pivot_longer(cols = c(3:length(apes01)), names_to = c("Participant", ".value"), names_sep="_")
      
      # add participant ID: 
      # to do so, first check how many participants
      # there are in the vocalizations spreadsheet
      # and remove those from the coder_id vector
      
      remove_n <- length(grep("Participant", colnames(read_excel(path, sheet = "Vocalizations"))))
      
      # these are the coder IDs for the apes:
      coder_ids_apes <- coder_ids[(remove_n+1) : length(coder_ids)]
      
      # number of participants
      n_c <- length(levels(factor(apes01$Participant)))
      
      # add coder ID
      apes01 <- left_join(apes01, tibble(Participant = paste0("P", c(1:n_c)),
             Coder_ID = coder_ids_apes[1:n_c]))
      
      # remove those for the next iteration
      coder_ids_apes2 <- coder_ids_apes[(n_c+1):length(coder_ids_apes)]
      
      # read second apes sheet
      apes02 <- read_excel(path, sheet = "Ape 2")
      
          # replace non-numeric items
    apes02 <- replace_non_numerics(apes02)
      
              # make sure that ratings are numeric
        apes02 <- apes02 %>% suppressWarnings(mutate(across(starts_with("Rating"), as.numeric)))
      
            # remove additional columns if there are any
      remove_cols <- grep("\\.\\.\\.", colnames(apes02))
      
      if(length(remove_cols) > 0) {
        apes02 <- apes02[,-remove_cols]
      }
      
      # make sure column names are consistent
    colnames(apes02) <- gsub("p4", "P4", colnames(apes02))
      
      # change column names for better pivoting
      colnames(apes02) <- gsub("Participant ", "Answer P", colnames(apes02))
      
      for(i in 1:length(colnames(apes02))) {
        if(grepl("P.$", colnames(apes02)[i])) {
          colnames(apes02)[i] <- add_participant_prefix(colnames(apes02)[i])
        }
      }
      
    # pivot to longer format
      apes02 <- apes02 %>% pivot_longer(cols = c(3:length(apes02)), names_to = c("Participant", ".value"), names_sep="_")
      
      
      
      # add participant ID: 
      # number of participants
      n_c <- length(levels(factor(apes02$Participant)))
      
      apes02 <- left_join(apes02, tibble(Participant = paste0("P", c(1:n_c)),
             Coder_ID = coder_ids_apes2))
      
      
      apes <- rbind(apes01, apes02)
      
      # remove Participant to avoid confusion
      apes <-select(apes, -Participant)
      
      # return result
      return(apes)
      
      
    } else {
      apes <- read_excel(path, sheet = "Ape 1")
      
          # replace non-numeric items
    apes <- replace_non_numerics(apes)
    
      
      # remove additional columns if there are any
      remove_cols <- grep("\\.\\.\\.", colnames(apes))
      if(length(remove_cols) > 0) {
        apes <- apes[,-remove_cols]
      }
      
      # make sure column names are consistent
    colnames(apes) <- gsub("p4", "P4", colnames(apes))
      
      # change column names for better pivoting
      colnames(apes) <- gsub("Participant ", "Answer P", colnames(apes))
      
      for(i in 1:length(colnames(apes))) {
        if(grepl("P.$", colnames(apes)[i])) {
          colnames(apes)[i] <- add_participant_prefix(colnames(apes)[i])
        }
      }
      
    # pivot to longer format
      apes <- apes %>% pivot_longer(cols = c(3:length(apes)), names_to = c("Participant", ".value"), names_sep="_")
      
      # add participant ID: 
      # to do so, first check how many participants
      # there are in the vocalizations spreadsheet
      # and remove those from the coder_id vector
      remove_n <- length(grep("Participant", colnames(read_excel(path, sheet = "Vocalizations"))))
      
      # these are the coder IDs for the apes:
      coder_ids_apes <- coder_ids[(remove_n+1) : length(coder_ids)]
      
      # number of participants
      n_c <- length(levels(factor(apes$Participant)))
      
       # add coder ID
      apes <- left_join(apes, tibble(Participant = paste0("P", c(1:n_c)),
             Coder_ID = coder_ids_apes[1:n_c]))
      
      # remove Participant to avoid confusion
      apes <-select(apes, -Participant)
      
      # return result
      return(apes)
      
      
    }
    
    
  }
  
}



# read in all Excel sheets 

for(i in 1:nrow(files)) {
  
  my_excel_sheets <- excel_sheets(files$file[i])
  
  if(any(grepl("Ape", my_excel_sheets))) {
    cur <- read_excel_custom(files$file[i], "apes")
    
    if(!exists("all_sheets")) {
      all_sheets <- cur
    } else {
      all_sheets <- rbind(all_sheets, cur)
    }
    
    
  }
  
  
  
  
  
  
  print(my_excel_sheets)
  
}


# writexl::write_xlsx(all_sheets, "helpers/all_apes.xlsx")

# same for vocalizations

for(i in 1:nrow(files)) {
  
  my_excel_sheets <- excel_sheets(files$file[i])
  
  if(any(grepl("Vocalization", my_excel_sheets))) {
    cur <- read_excel_custom(files$file[i], "voc")
    
    if(!exists("all_sheets2")) {
      all_sheets2 <- cur
    } else {
      all_sheets2 <- rbind(all_sheets2, cur)
    }
    
    
  }
  
  
  
  
  
  
  print(my_excel_sheets)
  
}


# writexl::write_xlsx(all_sheets2, "helpers/all_vocs.xlsx")



```

## Finding mismatches - vocalizations

```{r}

for(j in 1:length(files01)) {
  # matching files
cur_files <- unlist(files01[,j])


# vocalizations
cur_vocs <- lapply(cur_files, read_excel_custom, "voc")

for(i in 1:length(cur_vocs)) {
  
  colnames(cur_vocs[[i]])[which(colnames(cur_vocs[[i]]) == "Rating")] <- paste0("Rating", i)
  
}

cur_vocs01 <- cur_vocs[[1]]

if(length(cur_vocs)>1) {
  cur_vocs01$Rating2 <- cur_vocs[[2]]$Rating2
  cur_vocs01$match <- ifelse(cur_vocs01$Rating1 == cur_vocs01$Rating2, TRUE, FALSE)
}

if(length(cur_vocs)>2) {
  cur_vocs01$Rating3 <- cur_vocs[[3]]$Rating3
  cur_vocs01$match <- ifelse(cur_vocs01$Rating1 == cur_vocs01$Rating2 & cur_vocs01$Rating1 == cur_vocs01$Rating3, TRUE, FALSE)
}

# remove Coder IDs and add Task ID instead
cur_vocs01 <- cur_vocs01 %>% select(-`Trial id`, -Coder_ID)
cur_vocs01 <- mutate(cur_vocs01, Task_ID = colnames(files01)[j])

writexl::write_xlsx(filter(cur_vocs01, !match), paste0("mismatches/mismatches_voc", colnames(files01)[j], ".xlsx"))

print(j)
}






```

## 

## Finding mismatches - apes

```{r}

for(j in 1:length(files01)) {
  
  
  # matching files
cur_files <- unlist(files01[,j])



# check if apes are in the sheet
if(any(grepl("Ape", excel_sheets(cur_files[1])))) {
  
  # apes
cur_vocs <- lapply(cur_files, read_excel_custom, "apes")

for(i in 1:length(cur_vocs)) {
  
  colnames(cur_vocs[[i]])[which(colnames(cur_vocs[[i]]) == "Rating")] <- paste0("Rating", i)
  
}

cur_vocs01 <- cur_vocs[[1]]

if(length(cur_vocs)>1) {
  cur_vocs01$Rating2 <- cur_vocs[[2]]$Rating2
  cur_vocs01$match <- ifelse(cur_vocs01$Rating1 == cur_vocs01$Rating2, TRUE, FALSE)
}

if(length(cur_vocs)>2) {
  cur_vocs01$Rating3 <- cur_vocs[[3]]$Rating3
  cur_vocs01$match <- ifelse(cur_vocs01$Rating1 == cur_vocs01$Rating2 & cur_vocs01$Rating1 == cur_vocs01$Rating3, TRUE, FALSE)
}

# remove Coder IDs and add Task ID instead
cur_vocs01 <- cur_vocs01 %>% select(-matches("Trial [Ii]d"), -Coder_ID)
cur_vocs01 <- mutate(cur_vocs01, Task_ID = colnames(files01)[j])

writexl::write_xlsx(filter(cur_vocs01, !match), paste0("mismatches/mismatches_apes", colnames(files01)[j], ".xlsx"))
  
}



print(j)
}




```

## 

## Combine all mismatches

```{r}

f <- list.files("mismatches/", full.names = TRUE)
f_voc <- grep("apes", f, invert = T, value = T)
f_apes  <- grep("apes", f, invert = F, value = T)



mismatches_apes <- do.call(bind_rows, lapply(1:length(f_apes), function(i) mutate(read_xlsx(f_apes[i]), file = gsub(".*/", "", f_apes[i]))))

mismatches_voc <- do.call(bind_rows, lapply(1:length(f_voc), function(i) mutate(read_xlsx(f_voc[i]), file = gsub(".*/", "", f_voc[i]))))


```

## Calculate IAA - vocalizations

```{r}

all_iaa <- tibble(
  Task = rep(NA, ncol(files01)),
  file1 = character(ncol(files01)),
  file2 = character(ncol(files01)),
  file3 = character(ncol(files01)),
  Annotator1 = character(ncol(files01)),
  Annotator2 = character(ncol(files01)),
  Annotator3 = character(ncol(files01)),
  IAA = numeric(ncol(files01))
)


for(j in 1:length(files01)) {
  
  # list of current files
  cur_files <- unlist(files01[,j])
  
  # current task
all_iaa$Task[j] <- colnames(files01[,j])
all_iaa[j,(2:(2+(length(cur_files))-1))] <- tibble(files = cur_files) %>% t()

# get annotator codes
cur_annotators <- sapply(1:length(cur_files), function(i) unname(gsub(" .*", "", gsub(".*//", "", cur_files[i]))))

# add annotator codes to sheet
all_iaa[j,(5:(5+(length(cur_annotators))-1))] <- tibble(annotators = cur_annotators) %>% t() %>% unname()
  
  # read items
cur_sheets <- lapply(cur_files, read_excel_custom, type = "voc")

# use the first sheet as master sheet
# and add the ratings from the others
cur_master_sheet <- cur_sheets[[1]]

if(length(cur_sheets) == 3) {
  iaa_cur <- bind_cols(bind_cols(cur_master_sheet,
          select(cur_sheets[[2]], matches("[Rr]ating")), .name_repair = "unique"), select(cur_sheets[[3]], matches("[Rr]ating")), .name_repair = "unique") %>% select(matches("[Rr]ating")) %>% as.matrix() %>% t() %>% kripp.alpha(method = "ordinal")
} else {
  iaa_cur <- bind_cols(cur_master_sheet,
          select(cur_sheets[[2]], matches("[Rr]ating")), .name_repair = "unique") %>% select(matches("[Rr]ating")) %>% as.matrix() %>% t() %>% kripp.alpha(method = "ordinal")
}

all_iaa$IAA[j] <- iaa_cur$value

  
}

# write_xlsx(all_iaa, "iaa/all_iaa_vocalizations.xlsx")


```

## IAA Apes

```{r}

all_iaa <- tibble(
  Task = rep(NA, ncol(files01)),
  file1 = character(ncol(files01)),
  file2 = character(ncol(files01)),
  file3 = character(ncol(files01)),
  Annotator1 = character(ncol(files01)),
  Annotator2 = character(ncol(files01)),
  Annotator3 = character(ncol(files01)),
  IAA = numeric(ncol(files01))
)


for(j in 1:length(files01)) {
  
  # list of current files
  cur_files <- unlist(files01[,j])
  
  # check if there are ape sheets
  my_sheets <- excel_sheets(cur_files[1])
  
  if(any(grepl("Ape", my_sheets))) {
    # current task
all_iaa$Task[j] <- colnames(files01[,j])
all_iaa[j,(2:(2+(length(cur_files))-1))] <- tibble(files = cur_files) %>% t()

# get annotator codes
cur_annotators <- sapply(1:length(cur_files), function(i) unname(gsub(" .*", "", gsub(".*//", "", cur_files[i]))))

# add annotator codes to sheet
all_iaa[j,(5:(5+(length(cur_annotators))-1))] <- tibble(annotators = cur_annotators) %>% t() %>% unname()
  
  # read items
cur_sheets <- lapply(cur_files, read_excel_custom, type = "apes")

# use the first sheet as master sheet
# and add the ratings from the others
cur_master_sheet <- cur_sheets[[1]]

if(length(cur_sheets) == 3) {
  iaa_cur <- bind_cols(bind_cols(cur_master_sheet,
          select(cur_sheets[[2]], matches("[Rr]ating")), .name_repair = "unique"), select(cur_sheets[[3]], matches("[Rr]ating")), .name_repair = "unique") %>% select(matches("[Rr]ating")) %>% as.matrix() %>% t() %>% kripp.alpha(method = "ordinal")
} else {
  iaa_cur <- bind_cols(cur_master_sheet,
          select(cur_sheets[[2]], matches("[Rr]ating")), .name_repair = "unique") %>% select(matches("[Rr]ating")) %>% as.matrix() %>% t() %>% kripp.alpha(method = "ordinal")
}

all_iaa$IAA[j] <- iaa_cur$value
  }
  
  

  
}

# write_xlsx(all_iaa, "iaa/all_iaa_apes.xlsx")

```

## Visualize IAA distribution

```{r}

iaa_apes <- read_xlsx("iaa/all_iaa_apes.xlsx")
iaa_vocs <- read_xlsx("iaa/all_iaa_vocalizations.xlsx")

# replace annotator acronyms by Rater_number
replace_annotator_acronyms <- function(x) {
  x <- case_when(x == "SK" ~ "Rater 1",
          x == "MP" ~ "Rater 4",
          x == "SH" ~ "Rater 2",
          x == "F" ~ "Rater 3")
  
  return(x)
}

iaa_apes <- iaa_apes %>% mutate(across(starts_with("Annotator"), replace_annotator_acronyms))

iaa_vocs <- iaa_vocs %>% mutate(across(starts_with("Annotator"), replace_annotator_acronyms))

# remove empty rows
iaa_apes <- iaa_apes %>% filter(!is.na(Task))

# visualize

tibble(type = c(rep("Apes", nrow(iaa_apes)), rep("Vocalizations", nrow(iaa_vocs))), IAA = c(iaa_apes$IAA, iaa_vocs$IAA)) %>% ggplot(aes(x = type, y = IAA)) + geom_beeswarm(col = "darkblue") + geom_boxplot(width = .5, fill = "yellow", alpha = .1) + xlab("Type") + 
  theme(axis.text = element_text(size = 18), 
        axis.title = element_text(size = 18)) +
  ggtitle("Inter-Annotator Agreement") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5,size=19))

# ggsave("iaa_both.png", width = 8, height = 8)

```
