---
title: "Experiments in language evolution: inferences from closed- vs open-ended semantic space paradigms"
subtitle: "Preparation"
author: "analysis by Aleksandra Ćwiek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_document:
    theme: readable
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '3'
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What this document does

**Purpose.** This notebook loads raw data (vocalizations/gestures
metadata + ConceptNet similarities), cleans and harmonizes variable
names/keys, and produces a single analysis-ready table used by
`2_modeling.Rmd`.\

**Outputs.** (a) a cleaned CSV (`final_df.csv`) written to `data/`; (b)
sanity-check summaries and quick plots saved under
`plots/1_preparation/`.\

**Reproducibility.** All paths are relative, a fixed random seed is set,
and package versions are recorded at the end.

# Data preparation

## Source setup

```{r source setup, echo = TRUE, message=FALSE, warning = FALSE}
########## folders ##########
library(here)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

parentfolder <- dirname(getwd())

data          <- paste0(parentfolder, '/data/')
models        <- paste0(parentfolder, '/models/')
plots         <- paste0(parentfolder, '/plots/')
scripts       <- paste0(parentfolder, '/scripts/')

#################### packages ####################
# Core tidyverse for wrangling + plotting; janitor/skimr for quick structure checks; patchwork for combining plots
library(tidyverse)
library(janitor)
library(skimr)
library(knitr)
library(cowplot)
library(ggdist)
library(readxl)
library(lubridate)
library(patchwork)
library(hexbin)
library(PRROC)

# Keep plots/readability consistent
theme_set(theme_minimal(base_size = 12) + panel_border())

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

**Auto-saving figures.** Each plot with `fig.cap` is copied to
`plots/<Rmd-name>/Figure_XX.(png|tif)`. This guarantees stable filenames
for papers, corresponding to figure numbers from the html.

```{r echo = TRUE, message=FALSE, warning = FALSE}
# --- figure & table captions ---
outputFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to"); if (is.null(outputFormat)) outputFormat <- ""
capTabNo <- 1L; capFigNo <- 1L
.last_fig_no <- NA_integer_

capTab <- function(x){
  if (outputFormat == 'html'){
    x <- paste0("**Table ", capTabNo, ".** ", x)
    capTabNo <<- capTabNo + 1L
  }
  x
}

capFig <- function(x, show_R_version=FALSE, show_package_versions=NULL, is_map=FALSE){
  if (outputFormat == 'html'){
    assign(".last_fig_no", capFigNo, envir = .GlobalEnv) # remember current number
    x <- paste0("**Figure ", capFigNo, ".** ", x)
    capFigNo <<- capFigNo + 1L
  }
  x
}

# --- plot hook to auto-save ---
.default_plot_hook <- knitr::knit_hooks$get("plot")
.has_magick <- requireNamespace("magick", quietly = TRUE)

knitr::knit_hooks$set(plot = function(x, options){
  res <- if (is.function(.default_plot_hook)) .default_plot_hook(x, options) else NULL

  if (!is.null(options$fig.cap)) {
    fig_no <- get(".last_fig_no", envir = .GlobalEnv, inherits = FALSE)
    if (!is.na(fig_no)) {
      fig_no_str <- sprintf("%02d", fig_no)

      # subfolder = Rmd filename without extension
      rmd_name <- tools::file_path_sans_ext(basename(knitr::current_input()))
      target_dir <- file.path(plots, rmd_name)
      if (!dir.exists(target_dir)) dir.create(target_dir, recursive = TRUE)

      png_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".png"))
      tif_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".tif"))

      file.copy(x, png_out, overwrite = TRUE)

      if (.has_magick) {
        img <- magick::image_read(png_out)
        magick::image_write(img, path = tif_out, format = "tiff", compression = "lzw")
      } else {
        file.copy(png_out, tif_out, overwrite = TRUE)
      }
    }
  }

  res
})

```

## Load in data frames

-   `vocs.xlsx` — trial-level responses (free text), one row per
    participant×stimulus.

-   `all_metadata.xlsx` — stimulus metadata (concept, modality, etc.).

-   `cosine_sim.csv` — ConceptNet cosine between response head and
    target concept; first row is a header row from export and is
    promoted to column names here.

**Harmonization.** We `clean_names()` so joins are predictable; cosine
is coerced to numeric after header fix.

```{r echo=TRUE, message=FALSE, warning=FALSE}
vocs <- read_excel(paste0(data, "vocs.xlsx"))  %>% clean_names()
meta <- read_excel(paste0(data, "all_metadata.xlsx"))  %>% clean_names()

cn <- read_delim(paste0(data, "cosine_sim.csv"), delim = ';') %>% clean_names()
colnames(cn) <- as.character(cn[1,])
cn <- cn[-1,]
cn$cosine_similarity <- as.numeric(cn$cosine_similarity)
```

## Clean and prep data frame

**Cleaning & merges.** We standardize IDs (participant,
stimulus_concept), derive analysis variables, and left-join ConceptNet
scores on the response–target pair. We keep both the raw 4-point coder
scale and a binary “exact/near-miss vs miss” to support multiple
analysis lenses. Where coders disagreed, final labels reflect post-hoc
adjudication.

```{r echo=TRUE, message=FALSE, warning=FALSE}
## ---------- 1) Normalize vocs & cn ----------
vocs_keys <- vocs %>%
  mutate(
    right_answer_key = str_squish(tolower(right_answer)),
    answer_key       = str_squish(tolower(answer))
  )

cn_keys <- cn %>%
  rename(right_answer = Right_answer,
         answer       = Answer,
         cosine_similarity = cosine_similarity) %>%
  mutate(
    right_answer_key = str_squish(tolower(right_answer)),
    answer_key       = str_squish(tolower(answer))
  )

## ---------- 2) Fix final_decision ----------
vocs_fixed <- vocs_keys %>%
  mutate(
    final_decision    = ifelse(final_decision == 31, 2, final_decision),
    judged_similarity = final_decision,
    binary_guess      = ifelse(final_decision == 3, 1, 0)
  )

## ---------- 3) Clean native_language ----------
meta_lang <- meta %>%
  mutate(
    native_language = native_language %>%
      str_squish() %>%
      str_replace_all("[[:punct:]]+$", "") %>%
      str_trim(),
    nl_low = tolower(native_language),
    native_language = case_when(
      str_detect(nl_low, "english") ~ "English",
      nl_low %in% c("enlgish") ~ "English",
      str_detect(nl_low, "spanish") ~ "Spanish",
      str_detect(nl_low, "portugu") ~ "Portuguese",
      nl_low %in% c("portuguesa", "pirtuguese", "portugal") ~ "Portuguese",
      nl_low %in% c("italiano") | str_detect(nl_low, "^italian") ~ "Italian",
      str_detect(nl_low, "hebrew") ~ "Hebrew",
      str_detect(nl_low, "oromo|ormo") ~ "Oromo",
      str_detect(nl_low, "zulu") ~ "Zulu",
      TRUE ~ str_to_title(nl_low)
    )
  )

## ---------- 4) Time & coverage filters (session-level) ----------
per_part <- meta_lang %>%
  group_by(rec_session_id) %>%
  summarise(
    completed_all = all(completed, na.rm = TRUE),
    n_rows = n(),
    start_min = suppressWarnings(min(start_time, na.rm = TRUE)),
    end_max   = suppressWarnings(max(end_time,   na.rm = TRUE)),
    duration_min = as.numeric(difftime(end_max, start_min, units = "mins")),
    .groups = "drop"
  )

# Relative coverage vs. max rows recorded in meta for any participant
max_rows <- max(per_part$n_rows, na.rm = TRUE)

per_part <- per_part %>%
  mutate(
    pct_trials = n_rows / max_rows,
    reason_excl_session = case_when(
      !completed_all ~ "Not completed",
      pct_trials < 0.90 ~ "<90% trials",
      duration_min < 2 ~ "Too short (<2m)",
      duration_min > 100 ~ "Too long (>100m)",
      TRUE ~ NA_character_
    )
  )

## ---------- 5) Per-trial IQR outlier detection (within participant) ----------
# We attempt trial-level durations from meta (one row per trial is ideal).
# If only 1 row per participant, this step safely does nothing.
meta_trials <- meta_lang %>%
  select(rec_session_id, trial_id, start_time, end_time) %>%
  mutate(trial_dur_min = as.numeric(difftime(end_time, start_time, units = "mins"))) %>%
  filter(!is.na(trial_dur_min))

per_trial_outliers <- meta_trials %>%
  group_by(rec_session_id) %>%
  mutate(
    n_trials_avail = n(),
    Q1 = quantile(trial_dur_min, 0.25, na.rm = TRUE),
    Q3 = quantile(trial_dur_min, 0.75, na.rm = TRUE),
    IQRv = IQR(trial_dur_min, na.rm = TRUE),
    low_f = Q1 - 1.5 * IQRv,
    high_f = Q3 + 1.5 * IQRv,
    is_outlier = trial_dur_min < low_f | trial_dur_min > high_f
  ) %>%
  summarise(
    n_trials_avail = first(n_trials_avail),
    outlier_share = mean(is_outlier, na.rm = TRUE),
    reason_excl_trial = case_when(
      n_trials_avail >= 5 & outlier_share > 0.20 ~ "Per-trial time outliers >20% (IQR)",
      TRUE ~ NA_character_
    ),
    .groups = "drop"
  )

# Merge session reasons + per-trial reasons
per_part <- per_part %>%
  left_join(per_trial_outliers, by = "rec_session_id") %>%
  mutate(
    reason_exclusion = case_when(
      !is.na(reason_excl_session) & !is.na(reason_excl_trial) ~ paste(reason_excl_session, reason_excl_trial, sep = "; "),
      !is.na(reason_excl_session) ~ reason_excl_session,
      !is.na(reason_excl_trial) ~ reason_excl_trial,
      TRUE ~ NA_character_
    )
  ) %>%
  select(rec_session_id, n_rows, pct_trials, duration_min, outlier_share, reason_exclusion)

# Keep & exclude IDs
keep_ids <- per_part %>%
  filter(is.na(reason_exclusion)) %>%
  transmute(rec_session_id_chr = as.character(rec_session_id))

exclusion_report <- per_part %>%
  filter(!is.na(reason_exclusion)) %>%
  transmute(
    participant = as.character(rec_session_id),
    reason_exclusion,
    n_rows, pct_trials, duration_min, outlier_share
  )

exclusion_summary <- exclusion_report %>%
  separate_rows(reason_exclusion, sep = ";\\s*") %>%
  count(reason_exclusion, sort = TRUE)

## ---------- 6) Participant IDs from coder_id + corrections ----------
vocs_fixed <- vocs_fixed %>%
  mutate(
    participant = str_extract(coder_id, "\\d+"),
    participant = as.character(participant),
    participant = case_when(
      participant == "113978"   ~ "1113978",
      participant == "114003"   ~ "1114003",
      participant == "11134822" ~ "1113482",
      participant == "11112900" ~ "1112900",
      participant == "11129907" ~ "1112907",
      participant == "11129917" ~ "1112917",
      participant == "11129918" ~ "1112918",
      TRUE ~ participant
    )
  )

## ---------- 7) Merge cosine ----------
vocs_cn <- vocs_fixed %>%
  left_join(cn_keys %>% select(right_answer_key, answer_key, cosine_similarity),
            by = c("right_answer_key", "answer_key"))

## ---------- 8) Further cleanup meta ----------

meta_lang <- meta_lang %>%
  rename(environment = environment_choice) %>%
  mutate(
    environment = case_when(
      environment == 1 ~ "home",
      environment == 2 ~ "transport",
      environment == 3 ~ "another public space",
      environment == 4 ~ "other",
      TRUE ~ NA_character_
    ),
    audio_device = case_when(
      audio_device == 1 ~ "speaker",
      audio_device == 2 ~ "headphones",
      audio_device == 3 ~ "other",
      TRUE ~ NA_character_
    ),
    text_device = case_when(
      text_device == 1 ~ "computer",
      text_device == 2 ~ "phone",
      text_device == 3 ~ "other",
      TRUE ~ NA_character_
    )
  )


## ---------- 9) Final df (filter to kept participants first) ----------
df <- vocs_cn %>%
  semi_join(keep_ids, by = c("participant" = "rec_session_id_chr")) %>%
  left_join(
    meta_lang %>%
      mutate(rec_session_id_chr = as.character(rec_session_id)) %>%
      select(
        rec_session_id_chr,
        native_language,
        text_device,
        audio_device,
        environment,
        hearing_disabilities,
        toughness,
        quantity
      ),
    by = c("participant" = "rec_session_id_chr")
  ) %>%
  transmute(
    participant,
    native_language,
    hearing_disabilities,
    environment,
    audio_device,
    text_device,
    toughness,
    quantity_guessed = quantity,
    trial_id,
    stimulus_concept = right_answer,
    guessed_concept  = answer,
    binary_guess,
    judged_similarity,
    cosine_similarity
  )

## ---------- 10) Cleanup environment ----------
rm(vocs_keys, cn_keys, vocs_fixed, meta_lang, meta_trials, per_trial_outliers,
   per_part, max_rows, keep_ids, vocs_cn)

## ---------- 11) Outputs ----------
# df                 -> final dataset
# exclusion_report   -> detailed rows (participant-level) with reason(s)
# exclusion_summary  -> counts per reason
# check which rows are NA in cosine_similarity
# df %>% filter(is.na(cosine_similarity)) %>% nrow()
```

Check final df.

```{r echo=TRUE, message=FALSE, warning=FALSE}
skim(df)
```

Rescale cosine similarity (from −1–1 to 0–1).

```{r echo=TRUE, message=FALSE, warning=FALSE}
df <- df %>%
  mutate(
    cosine_rescaled = (cosine_similarity + 1)/2,          # -1..1 → 0..1
    ord_norm        = judged_similarity/3,                 # 0..3 → 0..1
    ord_f           = factor(judged_similarity, levels = 0:3),
    bin_f           = factor(binary_guess, levels = c(0,1))
  )

# For binning cosine
df_bins <- df %>%
  filter(!is.na(cosine_rescaled)) %>%
  mutate(cos_decile = ntile(cosine_rescaled, 10))
```

# Descriptive statistics

First sense of means and sd.

```{r echo=TRUE, message=FALSE, warning=FALSE}
df %>%
  summarise(
    n = n(),
    binary_mean = mean(binary_guess),
    binary_sd   = sd(binary_guess),
    ord_mean    = mean(judged_similarity),
    ord_sd      = sd(judged_similarity),
    cos_org_mean    = mean(cosine_similarity, na.rm=TRUE),
    cos_org_sd      = sd(cosine_similarity, na.rm=TRUE),
    cos_res_mean    = mean(cosine_rescaled, na.rm=TRUE),
    cos_res_sd      = sd(cosine_rescaled, na.rm=TRUE)
  )
```

Per-concept.

```{r}
df %>%
  group_by(stimulus_concept) %>%
  summarise(
    n = n(),
    mean_binary = mean(binary_guess),
    mean_judged = mean(judged_similarity),
    mean_cos_org = mean(cosine_similarity, na.rm = TRUE),
    mean_cos_res = mean(cosine_rescaled, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_binary))
```

Cross-tab of binary vs ordinal.

```{r echo=TRUE, message=FALSE, warning=FALSE}
table(df$binary_guess, df$judged_similarity)
```

## Add descriptive statistics on
participant,
    native_language,
    hearing_disabilities,
    environment,
    audio_device,
    text_device,
    toughness,
    quantity_guessed = quantity,

# Visual exploration

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Seed for jitter
set.seed(1708)
```

```{r distribution, echo=FALSE, fig.align='left', fig.cap=capFig("Distributions of each measure."), fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
p1 <- ggplot(df, aes(x=factor(binary_guess))) +
  geom_bar(fill="#F29724") +
  labs(title="Binary", x="Correct (1) vs. Incorrect (0)", y="Count")

p2 <- ggplot(df, aes(x=factor(judged_similarity))) +
  geom_bar(fill="#0E2B29") +
  labs(title="Ordinal rating", x="Similarity score (0–3)", y="Count")

p3 <- ggplot(df, aes(x=cosine_rescaled)) +
  geom_density(fill="#EAE8E0") +
  labs(title="Cosine similarity (rescaled)", x="0–1 scale", y="Density")

p1 + p2 + p3
```

```{r ordinal vs cosine, echo=FALSE, fig.align='left', fig.cap=capFig("Ordinal ratings vs cosine similarity."), fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
ggplot(df, aes(x=cosine_rescaled, y=factor(judged_similarity))) +
  geom_jitter(alpha=0.1, height=0.2, color="#0E2B29") +
  stat_summary(fun=mean, geom="point", color="red", size=3) +
  labs(x="Cosine similarity (rescaled)", y="Ordinal rating (0–3)")
```

```{r binary vs ordinal, echo=FALSE, fig.align='left', fig.cap=capFig("Heatmap: binary × ordinal."), fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
ggplot(df, aes(x=factor(judged_similarity), y=factor(binary_guess))) +
  geom_bin2d() +
  scale_fill_viridis_c() +
  labs(x="Ordinal rating", y="Binary")
```

------------------------------------------------------------------------

```{r ECDF, echo=FALSE, fig.align='left', fig.cap=capFig("Empirical cumulative distributions of normalized ordinal similarity (0–1) and rescaled cosine similarity (0–1). The dashed horizontal line marks the overall proportion of binary successes. Differences between ECDFs indicate distinct partitioning of the response space by human ratings and computational similarity; the binary baseline shows how frequently the strict criterion is met."), fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
# Long form for ECDF overlay
ecdf_dat <- df %>%
  select(ord_norm, cosine_rescaled) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value))

ggplot(ecdf_dat, aes(value, colour = metric)) +
  stat_ecdf(geom = "step") +
  geom_hline(yintercept = mean(df$binary_guess), linetype = 2) +
  scale_colour_manual(values = colorBlindBlack8) +
  labs(x = "Similarity (0–1)", y = "F(value)", colour = NULL) +
  theme(legend.position = "top")

```

If the cosine ECDF lies to the right of the ordinal ECDF, cosine assigns
higher similarity more often; if left, it’s more conservative. The
dashed line (binary mean) highlights how much graded signal exists below
the strict “hit” rate.

------------------------------------------------------------------------

```{r Bland-Altman, echo=FALSE, fig.align='left', fig.cap=capFig("Bland–Altman plot comparing normalized ordinal similarity and rescaled cosine similarity: difference (ordinal − cosine) versus their mean. Solid line shows mean bias; dashed lines show ±1.96 SD limits of agreement. Systematic deviations from zero indicate directional disagreement; funneling suggests heteroscedasticity."), fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
ba <- df %>%
  filter(!is.na(cosine_rescaled)) %>%
  transmute(
    mean_pair = (ord_norm + cosine_rescaled)/2,
    diff_pair = ord_norm - cosine_rescaled
  )

mu  <- mean(ba$diff_pair)
sdv <- sd(ba$diff_pair)

ggplot(ba, aes(mean_pair, diff_pair)) +
  geom_point(alpha = 0.2, shape = 16) +
  geom_hline(yintercept = mu, size = 0.4) +
  geom_hline(yintercept = mu + 1.96*sdv, linetype = 2) +
  geom_hline(yintercept = mu - 1.96*sdv, linetype = 2) +
  labs(x = "Mean of (ordinal, cosine)",
       y = "Difference (ordinal – cosine)")
```

Points above zero mean humans rated higher than cosine; below zero means
cosine outruns human ratings. If spread widens at higher means, the two
measures diverge more for “more similar” responses.

------------------------------------------------------------------------

```{r ordinal composition, echo=FALSE, fig.align='left', fig.cap=capFig("Stacked proportions of ordinal categories (0–3) across deciles of rescaled cosine similarity. Increasing shares of higher ordinal categories with higher cosine deciles indicate convergence between computational and human-graded similarity."), fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
ord_mix <- df_bins %>%
  count(cos_decile, ord_f) %>%
  group_by(cos_decile) %>% mutate(prop = n/sum(n))

ggplot(ord_mix, aes(x = cos_decile, y = prop, fill = ord_f)) +
  geom_col(width = 0.9, alpha = 0.8) +
  #scale_fill_grey(start = 0.8, end = 0.2) +
  scale_fill_manual(values = colorBlindBlack8[4:8]) +
  labs(x = "Cosine decile (low → high)", y = "Proportion", fill = "Ordinal") + 
  theme(legend.position = "top")
```

Watch the bars: if the “2” and “3” slices expand toward the right,
cosine aligns with human-perceived similarity. Flat or noisy profiles
imply weak alignment.

------------------------------------------------------------------------

```{r joint hexbin, echo=FALSE, fig.align='left', fig.cap=capFig("Joint distribution of rescaled cosine similarity and ordinal ratings, with the fitted logistic decision boundary for binary correctness (binary scaled to 0–3 for overlay). Kernel contours (or hexbin density) summarize mass in the graded space; the logistic curve indicates where responses cross into binary “correct.”"), fig.height=5, fig.width=7, message=FALSE, warning=FALSE}

dat <- df %>% dplyr::filter(!is.na(cosine_rescaled))

ggplot(dat, aes(x = cosine_rescaled, y = judged_similarity)) +
  stat_binhex(bins = 35) +
  scale_fill_gradient(low = "grey90", high = "grey20", guide = "none") +
  geom_jitter(aes(y = binary_guess * 3), alpha = .10, height = .15, width = 0, size = 0.4) +
  stat_smooth(
    data = dat,
    mapping = aes(x = cosine_rescaled, y = binary_guess),
    method = "glm",
    method.args = list(family = binomial()),
    se = FALSE, colour = "black",
    inherit.aes = FALSE
  ) +
  scale_y_continuous(
    name = "Ordinal (0–3) & binary (scaled)",
    sec.axis = sec_axis(~ . / 3, name = "Pr(Binary = 1)")  # map left 0–3 to right 0–1
  ) +
  labs(#title = "Joint space (hexbin): Cosine vs Ordinal with binary decision boundary",
       x = "Cosine (0–1)") 
```

Dense regions at mid-to-high cosine and mid ordinal values reveal
near-misses. The logistic curve shows the effective cosine threshold
required for binary success.

------------------------------------------------------------------------

```{r observed p vs cosine, echo=FALSE, fig.align='left', fig.cap=capFig("Observed probability of binary correctness as a function of binned cosine similarity. Points are sized by bin counts and colored by observed probability. The curve indicates how increases in computational similarity translate into strict success."), fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
cal <- df %>%
  filter(!is.na(cosine_rescaled)) %>%
  mutate(bin = cut(cosine_rescaled, breaks = seq(0, 1, by = .05), include.lowest = TRUE)) %>%
  group_by(bin) %>%
  summarise(
    x = mean(cosine_rescaled),
    p = mean(binary_guess),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(is.finite(x), is.finite(p), n > 0)

ggplot(cal, aes(x, p)) +
  geom_line(color = "grey30") +
  geom_point(aes(size = n, fill = p), shape = 21, color = "black") +
  scale_fill_gradient(low = "#c6dbef", high = "#08306b") +  # minimal but expressive
  coord_cartesian(ylim = c(0,1)) +
  labs(x = "Cosine (binned midpoints)", y = "Observed probability", size = "N", fill = "Pr(1)")
```

Monotonic rise = good calibration; plateaus or step-like jumps indicate
thresholds. Large, pale points with low p show regions with many
attempts but low strict success.

------------------------------------------------------------------------

```{r precision-recall, echo=FALSE, fig.align='left', fig.cap=capFig("Precision–Recall curve using rescaled cosine similarity as a decision score for predicting binary correctness under class imbalance. Dashed horizontal line marks the baseline prevalence. Area under the PR curve (AUPRC) quantifies utility beyond prevalence."), fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
sc <- df %>% filter(!is.na(cosine_rescaled))
pr <- pr.curve(
  scores.class0 = sc$cosine_rescaled[sc$binary_guess==1],
  scores.class1 = sc$cosine_rescaled[sc$binary_guess==0],
  curve = TRUE
)

pr_dat <- data.frame(recall = pr$curve[,1], precision = pr$curve[,2])

ggplot(pr_dat, aes(recall, precision)) +
  geom_path() +
  geom_hline(yintercept = mean(sc$binary_guess), linetype = 2) +
  labs(#title = "Precision–Recall using cosine to predict binary",
       subtitle = paste0("AUPRC ≈ ", round(pr$auc.integral, 3),
                         " | baseline = ", round(mean(sc$binary_guess), 3)),
       x = "Recall", y = "Precision") 
```

If the PR curve stays well above the baseline, cosine is informative for
the hard task. A rapid precision drop at higher recall signals many
false positives among “near-similar” cases.

------------------------------------------------------------------------

```{r near-miss quantification, echo=FALSE, fig.align='left', fig.cap=capFig("Proportions of responses that are binary failures (0) yet exceed a high-similarity threshold either in cosine (e.g., ≥ .70) or in ordinal ratings (≥ 2). Bars show the share of “discarded but semantically close” responses."), fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
nm <- df %>% filter(!is.na(cosine_rescaled)) %>%
  summarise(
    n_total = n(),
    n_nm_cos70 = sum(binary_guess==0 & cosine_rescaled >= .70),
    n_nm_ord2  = sum(binary_guess==0 & judged_similarity >= 2),
    frac_nm_cos70 = n_nm_cos70/n_total,
    frac_nm_ord2  = n_nm_ord2/n_total
  )

nm
# For a compact bar:
nm_long <- nm %>%
  pivot_longer(starts_with("frac"), names_to = "near_miss", values_to = "frac")
ggplot(nm_long, aes(near_miss, frac)) +
  geom_col(width = 0.6, fill = "grey20") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(#title = "Share of **near-misses** hidden by binary=0",
       x = NULL, y = "Proportion of all responses")
```

These are the cases the abstract cares about: they are wrong by the
strict criterion but meaningfully similar by graded/computational
measures. Higher bars = more nuance lost by binary.

------------------------------------------------------------------------

```{r per-item slope, echo=FALSE, fig.align='left', fig.cap=capFig("For each stimulus concept, lines connect mean scores on Binary (proportion), Ordinal (normalized), and Cosine (rescaled). Line opacity encodes the within-item spread (max–min). Diverging trajectories indicate disagreement across measures at the item level."), fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
by_item_long <- df %>%
  group_by(stimulus_concept) %>%
  summarise(
    Binary  = mean(binary_guess, na.rm = TRUE),
    Ordinal = mean(ord_norm, na.rm = TRUE),
    Cosine  = mean(cosine_rescaled, na.rm = TRUE)
  ) %>%
  pivot_longer(-stimulus_concept, names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = factor(Metric, levels = c("Binary","Ordinal","Cosine")),
         x = as.numeric(Metric)) %>%
  group_by(stimulus_concept) %>%
  mutate(span = max(Value) - min(Value)) %>%
  ungroup()

ggplot(by_item_long, aes(x, Value, group = stimulus_concept)) +
  geom_line(aes(alpha = scales::rescale(span)), colour = "grey20") +
  geom_point(size = 0.8, colour = "black") +
  scale_x_continuous(breaks = 1:3, labels = levels(by_item_long$Metric)) +
  coord_cartesian(ylim = c(0,1)) +
  guides(alpha = "none") +
  labs(#title = "Per-item shifts across metrics (0–1 scale)",
       x = NULL, y = NULL)
```

Look for items whose lines climb from Binary to Ordinal/Cosine: these
are systematically “rescued” by graded measures. Flat lines indicate
consensus across metrics.

------------------------------------------------------------------------

```{r hexbin binary, echo=FALSE, fig.align='left', fig.cap=capFig("Hexbin density of rescaled cosine similarity versus normalized ordinal similarity, with binary successes overlaid as open circles. Shading reflects observations per hexagon."), fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
base <- ggplot(df %>% filter(!is.na(cosine_rescaled)),
               aes(cosine_rescaled, ord_norm)) +
  stat_binhex(bins = 35) +
  scale_fill_gradient(low = "grey85", high = "grey20", guide = "none") +
  labs(#title = "Cosine vs Ordinal (hexbin density) with binary=1 overlay",
       x = "Cosine (0–1)", y = "Ordinal (0–1)")

base + geom_point(data = subset(df, binary_guess == 1 & !is.na(cosine_rescaled)),
                  aes(cosine_rescaled, ord_norm),
                  shape = 1, size = 3, alpha = 0.6)
```

If successes cluster at the upper-right, binary hits require both high
cosine and high ordinal. Sparse, isolated success markers reveal “lucky
hits” or boundary cases.

------------------------------------------------------------------------

```{r disagreement, echo=FALSE, fig.align='left', fig.cap=capFig("Stimulus concepts ranked by absolute disagreement between mean normalized ordinal similarity and mean rescaled cosine similarity. Positive bars indicate higher human ratings than cosine; negative bars indicate higher cosine than human ratings."), fig.height=9, fig.width=6, message=FALSE, warning=FALSE}
disc <- df %>%
  filter(!is.na(cosine_rescaled)) %>%
  group_by(stimulus_concept) %>%
  summarise(
    ord = mean(ord_norm), cos = mean(cosine_rescaled),
    diff = ord - cos, adiff = abs(diff), n = n()
  ) %>%
  arrange(desc(adiff))


ggplot(disc, aes(x = reorder(stimulus_concept, adiff), y = diff)) +
  geom_col(fill = "grey30") +
  coord_flip() +
  geom_hline(yintercept = 0, colour = "black") +
  labs(#title = "Top stimulus-level disagreements: ordinal − cosine",
       x = NULL, y = "Difference (−=cosine higher, +=ordinal higher)")
```

This is the case-study list. Positive extremes: humans perceive
similarity that cosine underestimates (e.g., pragmatic or contextual
ties). Negative extremes: cosine overestimates semantic closeness
relative to human perception.

# Export df

**Output.** We write a single analysis-ready CSV (`final_df.csv`)
consumed by `2_modeling.Rmd`.

```{r echo=TRUE, message=FALSE, warning=FALSE}
write_csv(df, paste0(data, "final_df.csv"))
```

# Session info

```{r echo=TRUE, message=FALSE, warning=FALSE}
sessionInfo()
```
