---
title: "Experiments in language evolution: inferences from closed- vs open-ended semantic space paradigms"
subtitle: "Data modeling"
author: "analysis by Aleksandra Ćwiek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_document:
    theme: readable
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '3'
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What this document does

**Purpose.** This notebook fits and evaluates **baseline hierarchical
models** for three outcome measures that capture how well participants’
free-text responses match target concepts:

1.  `binary_guess` — whether the response was judged exact/near-exact
    (yes/no),

2.  `judged_similarity` — coder-assigned similarity on a four-point
    scale,

3.  `cosine_rescaled` — automatic cosine similarity (ConceptNet,
    rescaled to [0,1]).

Each outcome is analyzed separately. We summarize posterior
distributions, report Bayesian R², and check convergence/diagnostics.

**Inputs.** One cleaned, analysis-ready dataset from `1_preparation.Rmd`
(`data/final_df.csv`), containing participant IDs, stimulus/concept IDs,
language, and the three outcome measures.

**Outputs.**

-   Cached fitted models in `models/` (RDS files).

-   Summary tables (posterior means/HDIs, random-effect SDs, Bayesian
    R²).

-   Diagnostic plots and posterior predictive checks.

-   Figures saved to `plots/2_modeling/` (auto-saved with stable
    numbering).

# Data preparation

## Source setup

```{r source setup, echo = TRUE, message=FALSE, warning = FALSE}
########## folders ##########
library(here)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

parentfolder <- dirname(getwd())

data          <- paste0(parentfolder, '/data/')
models        <- paste0(parentfolder, '/models/')
plots         <- paste0(parentfolder, '/plots/')
scripts       <- paste0(parentfolder, '/scripts/')

########## source file ##########

#source(paste0(scripts, "adjectives-preparation.R"))

#################### packages ####################
# Core tidyverse for wrangling + plotting; janitor/skimr for quick structure checks; patchwork for combining plots
library(tidyverse)
library(janitor)
library(skimr)
library(knitr)
library(cowplot)
library(ggdist)
library(readxl)
library(lubridate)
library(patchwork)
library(hexbin)
library(PRROC)

# Bayesian
library(brms);
library(loo);packageVersion("loo")
library(HDInterval);packageVersion("HDInterval")
library(cmdstanr);
library(emmeans);packageVersion("emmeans")
library(posterior);
library(bayesplot);
library(BayesFactor);
library(tidybayes);packageVersion("tidybayes")
library(bayestestR);
library(cmdstanr)
options(brms.backend = "cmdstanr") # Speed gain: ~30-50% faster than rstan
options(brms.file_refit = "on_change")  # only refit if model changes

# use all available cores for parallel computing
options(mc.cores = parallel::detectCores())

# Keep plots/readability consistent
theme_set(theme_minimal(base_size = 12) + panel_border())

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r echo = TRUE, message=FALSE, warning = FALSE}
# --- figure & table captions ---
outputFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to"); if (is.null(outputFormat)) outputFormat <- ""
capTabNo <- 1L; capFigNo <- 1L
.last_fig_no <- NA_integer_

capTab <- function(x){
  if (outputFormat == 'html'){
    x <- paste0("**Table ", capTabNo, ".** ", x)
    capTabNo <<- capTabNo + 1L
  }
  x
}

capFig <- function(x, show_R_version=FALSE, show_package_versions=NULL, is_map=FALSE){
  if (outputFormat == 'html'){
    assign(".last_fig_no", capFigNo, envir = .GlobalEnv) # remember current number
    x <- paste0("**Figure ", capFigNo, ".** ", x)
    capFigNo <<- capFigNo + 1L
  }
  x
}

# --- plot hook to auto-save ---
.default_plot_hook <- knitr::knit_hooks$get("plot")
.has_magick <- requireNamespace("magick", quietly = TRUE)

knitr::knit_hooks$set(plot = function(x, options){
  res <- if (is.function(.default_plot_hook)) .default_plot_hook(x, options) else NULL

  if (!is.null(options$fig.cap)) {
    fig_no <- get(".last_fig_no", envir = .GlobalEnv, inherits = FALSE)
    if (!is.na(fig_no)) {
      fig_no_str <- sprintf("%02d", fig_no)

      # subfolder = Rmd filename without extension
      rmd_name <- tools::file_path_sans_ext(basename(knitr::current_input()))
      target_dir <- file.path(plots, rmd_name)
      if (!dir.exists(target_dir)) dir.create(target_dir, recursive = TRUE)

      png_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".png"))
      tif_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".tif"))

      file.copy(x, png_out, overwrite = TRUE)

      if (.has_magick) {
        img <- magick::image_read(png_out)
        magick::image_write(img, path = tif_out, format = "tiff", compression = "lzw")
      } else {
        file.copy(png_out, tif_out, overwrite = TRUE)
      }
    }
  }

  res
})

```

Other helpers

```{r}
bb_corr <- function(x, y, R = 4000, seed = 1708) {
  stopifnot(length(x) == length(y))
  set.seed(seed)
  n <- length(x)
  out <- numeric(R)
  for (r in seq_len(R)) {
    w <- rexp(n)          # Dirichlet via i.i.d. Exponential then normalize
    w <- w / sum(w)
    mx <- sum(w * x); my <- sum(w * y)
    sx <- sqrt(sum(w * (x - mx)^2))
    sy <- sqrt(sum(w * (y - my)^2))
    out[r] <- sum(w * (x - mx) * (y - my)) / (sx * sy)
  }
  out
}

summ_post <- function(draws, name){
  h <- hdi(draws, ci = 0.95)
  tibble(pair = name,
         mean = mean(draws),
         median = median(draws),
         hdi_low = h$CI_low,
         hdi_high = h$CI_high)
}

bf_bic <- function(y, x){
  fit1 <- lm(y ~ x)
  fit0 <- lm(y ~ 1)
  BIC0 <- BIC(fit0); BIC1 <- BIC(fit1)
  # BF10 ≈ exp((BIC0 - BIC1)/2)
  exp((BIC0 - BIC1)/2)
}
```

## Load in data frames

```{r echo=TRUE, message=FALSE, warning=FALSE}
df    <- read_csv(paste0(data, "final_df.csv"))  %>% clean_names()
cats  <- read_csv(paste0(data, "meaning_categories.csv")) %>% clean_names()
```

# Correlations

```{r echo=TRUE, message=FALSE, warning=FALSE}
dat_cor <- df %>%
  select(cosine = cosine_rescaled, judged = ord_norm, binary = bin_f) %>%
  drop_na()
```

Pearson correlations (with CIs)

```{r echo=TRUE, message=FALSE, warning=FALSE}
pearson_mat <- cor(dat_cor, use = "pairwise.complete.obs", method = "pearson")
pearson_mat
```

Pairwise Pearson tests (p-values + CI)

```{r echo=TRUE, message=FALSE, warning=FALSE}
pearson_tests <- list(
  cosine_judged = cor.test(dat_cor$cosine, dat_cor$judged),
  cosine_binary = cor.test(dat_cor$cosine, dat_cor$binary),
  judged_binary = cor.test(dat_cor$judged, dat_cor$binary)
)
pearson_tests

pearson_tbl <- tibble(
  pair = c("cosine ~ judged", "cosine ~ binary", "judged ~ binary"),
  r = c(cor(dat_cor$cosine, dat_cor$judged),
        cor(dat_cor$cosine, dat_cor$binary),
        cor(dat_cor$judged, dat_cor$binary))
)
```

Bayesian correlations via BayesFactor (Point-biserial is just Pearson on
numeric 0/1; we still get Bayesian posterior for rho.)

```{r echo=TRUE, message=FALSE, warning=FALSE}
post_cos_jud <- bb_corr(dat_cor$cosine, dat_cor$judged, R = 8000, seed = 1708)
post_cos_bin <- bb_corr(dat_cor$cosine, dat_cor$binary, R = 8000, seed = 1708)
post_jud_bin <- bb_corr(dat_cor$judged, dat_cor$binary, R = 8000, seed = 1708)

bayes_corr_summary <- bind_rows(
  summ_post(post_cos_jud, "cosine ~ judged"),
  summ_post(post_cos_bin, "cosine ~ binary"),
  summ_post(post_jud_bin, "judged ~ binary")
)
bayes_corr_summary

# Posterior density plots
bind_rows(
  data.frame(rho = post_cos_jud, pair = "cosine~judged"),
  data.frame(rho = post_cos_bin, pair = "cosine~binary"),
  data.frame(rho = post_jud_bin, pair = "judged~binary")
) %>%
  ggplot(aes(rho)) + geom_density() + facet_wrap(~pair, scales = "free") +
  geom_vline(xintercept = 0, linetype = 2)
```

BIC-approx Bayes factors for "correlation != 0" For
continuous-continuous: linear model; for (cont \~ binary): also linear
is fine here.

```{r echo=TRUE, message=FALSE, warning=FALSE}
BF_cos_jud <- bf_bic(dat_cor$cosine, dat_cor$judged)
BF_cos_bin <- bf_bic(dat_cor$cosine, dat_cor$binary)
BF_jud_bin <- bf_bic(dat_cor$judged, dat_cor$binary)

bf_tbl <- tibble(
  pair = c("cosine ~ judged", "cosine ~ binary", "judged ~ binary"),
  BF10 = c(BF_cos_jud, BF_cos_bin, BF_jud_bin)
)
```

One clean summary table

```{r echo=TRUE, message=FALSE, warning=FALSE}
pearson_tbl %>%
  left_join(bayes_corr_summary, by = "pair") %>%
  left_join(bf_tbl, by = "pair")
```

# Variance of random factors

Explore the variance across all possible random factors:

-   Environment

-   Input device

-   Output device

-   Native language

-   Participant

-   Stimulus/concept

-   Hearing disabilities

# Modeling

## What we model

-   **Binary accuracy**: `binary_guess ∈ {0,1}`, modeled with a
    **Bernoulli** distribution and logit link.

-   **Ordinal judgments**: `judged_similarity ∈ {0,1,2,3}`, modeled with
    a **cumulative logit (ordinal)** likelihood.

-   **Cosine similarity**: `cosine_rescaled ∈ [0,1]`, modeled with a
    **zero-one-inflated Beta (ZOIB)** or a simpler **Beta** after
    applying an ε-open transform to handle boundary values.

All models include varying intercepts for participants and stimuli to
account for individual and item-level heterogeneity.

## Full (for Cognitive Science)

We fit baseline models for each of the three measures, with only
random intercepts for participant, native language, and stimulus.

Full solution. (Beware: It might take days.)

```{r eval=FALSE, message=FALSE, warning=FALSE}
{
  set.seed(1708)

  ## ---- ensure models dir exists (you already have `models` as an absolute path) ----
  if (!dir.exists(models)) dir.create(models, recursive = TRUE, showWarnings = FALSE)

  ## ---- data prep (minimal; no z-scoring of cosine) ----
  ## ---------- COMPLETE-CASE DATA (based on cosine)
  # Rationale: use the same rows for all three outcomes to enable direct comparability
  if (!"ord_norm" %in% names(df)) df$ord_norm <- df$judged_similarity / 3
  df_brm <- subset(df, !is.na(cosine_rescaled))

  ## Keep exact-1 spike info for reporting (from original column but CC rows)
  ones_table_stim <- aggregate(I(df_brm$cosine_rescaled == 1) ~ stimulus_concept, data = df_brm, FUN = mean)
  names(ones_table_stim) <- c("stimulus_concept", "one_rate")

  df_brm$participant      <- factor(df_brm$participant)
  df_brm$stimulus_concept <- factor(df_brm$stimulus_concept)
  df_brm$native_language  <- factor(df_brm$native_language) # decided to do without it because it was not variable and caused much longer computation times
  df_brm$judged_ord       <- factor(df_brm$judged_similarity, levels = 0:3, ordered = TRUE)
 # Replace values extremely close to 1 with exactly 1
  df_brm$cosine_rescaled <- ifelse(
    !is.na(df_brm$cosine_rescaled) & df_brm$cosine_rescaled > 1 & df_brm$cosine_rescaled <
      1.001,
    1,
    df_brm$cosine_rescaled
  )
  
  ## ---- cosine preparation (rescaling to make comparable across measures) ----
  
  rescale_to_unit <- function(x) {
    eps <- 1e-10
    tiny <- 1e-12
    
    # Step 1: Epsilon-open to avoid exact boundaries
    x_open <- pmin(pmax(x - eps, tiny), 1 - tiny)
    
    # Step 2: Min-max rescale to [0,1]
    xmin <- min(x_open, na.rm = TRUE)
    xmax <- max(x_open, na.rm = TRUE)
    span <- xmax - xmin
    
    if (!is.finite(span) || span <= 0) {
      stop("Cannot rescale: no variation in x")
    }
    
    x_scaled <- (x_open - xmin) / span
    
    # Step 3: Ensure strictly in (0,1) for Beta
    x_final <- pmin(pmax(x_scaled, tiny), 1 - tiny)
    
    return(x_final)
  }
  
  # Apply to data
  # Nudge exact 1.0s only (keeping original scale)
  df_brm$cosine_original <- df_brm$cosine_rescaled
  df_brm$cosine_original[df_brm$cosine_original == 1.0] <- 0.999999
  
  # Full min-max rescaling (for comparable plot)
  df_brm$cosine_scaled <- rescale_to_unit(df_brm$cosine_rescaled)
  
  ## ---- very wide priors for baselines / couplings (not for ZOIB components) ----
  .pri_bin <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )
  .pri_ord <- c(
    set_prior("normal(0, 5)", class = "Intercept"),  # thresholds
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )
  .pri_wide <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )

  ## ========================= BASELINES =========================
  .bin_rds <- file.path(models, "fit_bin.rds")
  if (file.exists(.bin_rds)) {
    fit_bin <- readRDS(.bin_rds)
  } else {
    fit_bin <- brm(
      binary_guess ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = bernoulli(),
      prior   = .pri_bin,
      backend = "cmdstanr",
      chains  = 4, iter = 8000, warmup = 4000, 
      threads = threading(2), cores = parallel::detectCores(),
      control = list(adapt_delta = 0.97, max_treedepth = 12),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_bin, .bin_rds)
  }

  .ord_rds <- file.path(models, "fit_ord.rds")
  if (file.exists(.ord_rds)) {
    fit_ord <- readRDS(.ord_rds)
  } else {
    fit_ord <- brm(
      judged_ord ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = cumulative(link = "logit"),
      prior   = .pri_ord,
      backend = "cmdstanr",
      chains  = 4, iter = 10000, warmup = 4000, 
      threads = threading(2), cores = parallel::detectCores(),
      control = list(adapt_delta = 0.99, max_treedepth = 13),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_ord, .ord_rds)
  }

  .cos_org_rds <- file.path(models, "fit_cos_org.rds")
  if (file.exists(.cos_org_rds)) {
    fit_cos_org <- readRDS(.cos_org_rds)
  } else {
    fit_cos_org <- brm(
      cosine_original ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = Beta(), 
      prior = c(
        set_prior("normal(0, 2)", class = "Intercept"),
        set_prior("student_t(3, 0, 2.5)", class = "sd"),
        set_prior("gamma(0.01, 0.01)", class = "phi")),
      backend = "cmdstanr",
      chains  = 4, iter = 10000, warmup = 4000, 
      threads = threading(2), cores = parallel::detectCores(),
      control = list(adapt_delta = 0.99, max_treedepth = 13),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_cos_org, .cos_org_rds)
  }
  
  .cos_com_rds <- file.path(models, "fit_cos_com.rds")
  if (file.exists(.cos_com_rds)) {
    fit_cos_com <- readRDS(.cos_com_rds)
  } else {
    fit_cos_com <- brm(
      cosine_scaled ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = Beta(), 
      prior = c(
        set_prior("normal(0, 2)", class = "Intercept"),
        set_prior("student_t(3, 0, 2.5)", class = "sd"),
        set_prior("gamma(0.01, 0.01)", class = "phi")),
      backend = "cmdstanr",
      chains  = 4, iter = 10000, warmup = 4000, 
      threads = threading(2), cores = parallel::detectCores(),
      control = list(adapt_delta = 0.99, max_treedepth = 13),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_cos_com, .cos_com_rds)
  }

  ## ---- compact Bayes R^2 (kept as r2_table) ----
  # Path to saved R2 summary
  .r2_rds <- file.path(models, "r2_table.rds")
  
  if (file.exists(.r2_rds)) {
    # If the summary table already exists, just load it
    r2_table <- readRDS(.r2_rds)
    message("Loaded existing r2_table from ", .r2_rds)
  
  } else {
    # Otherwise, compute it fresh from model fits
    .r2 <- function(fit) {
      r <- bayes_R2(fit)
      c(Estimate = mean(r),
        Q2.5 = quantile(r, .025),
        Q97.5 = quantile(r, .975))
    }
  
    r2_table <- data.frame(
      model = c("Binary (Bernoulli)", "Judged (Ordinal)", "Cosine (Beta)"),
      rbind(.r2(fit_bin), .r2(fit_ord), .r2(fit_cos_org))
    )
  
    # clean rounding
    r2_table$Estimate <- round(r2_table$Estimate, 3)
    r2_table$Q2.5     <- round(r2_table$Q2.5, 3)
    r2_table$Q97.5    <- round(r2_table$Q97.5, 3)
  
    # Save table to reuse later
    saveRDS(r2_table, .r2_rds)
    message("Saved new r2_table to ", .r2_rds)
  }
  

  ## ---- cleanup helpers (keep only model objects + r2_table + path vars) ----
  rm(list = c(".pri_bin",".pri_ord",".pri_wide",
              ".bin_rds",".ord_rds",".cos_org_rds", ".cos_com_rds",
              ".m_ord_cos_rds",".m_bin_cos_rds",".m_bin_jud_rds",".m_cos_jud_rds", ".r2_rds"))
}
```

## Extract results

Compare the cosine similarity models on two scales

```{r}
# Extract stimulus posteriors from both models
stim_original <- fitted(fit_cos_org, 
                       newdata = data.frame(stimulus_concept = levels(df_brm$stimulus_concept)),
                       re_formula = ~(1|stimulus_concept))

stim_scaled <- fitted(fit_cos_com,
                     newdata = data.frame(stimulus_concept = levels(df_brm$stimulus_concept)),
                     re_formula = ~(1|stimulus_concept))

# Perfect rank preservation
cor(stim_original[,"Estimate"], stim_scaled[,"Estimate"], method = "spearman")
# Should be 1.0
```

Verify Rank Concordance

```{r}
# Extract stimulus-level predictions from both
get_stim_preds <- function(fit, label) {
  nd <- data.frame(stimulus_concept = levels(df_brm$stimulus_concept))
  preds <- fitted(fit, newdata = nd, 
                  re_formula = ~(1|stimulus_concept), 
                  summary = TRUE)
  tibble(
    stimulus = nd$stimulus_concept,
    mean = preds[, "Estimate"],
    model = label
  )
}

stim_both <- bind_rows(
  get_stim_preds(fit_cos_org, "Original scale"),
  get_stim_preds(fit_cos_com, "Rescaled")
)

# Rank correlation (should be perfect)
stim_wide <- stim_both %>%
  pivot_wider(names_from = model, values_from = mean)

cor(stim_wide$`Original scale`, stim_wide$`Rescaled`, method = "spearman")
# Should be 1.0 - perfect rank preservation

# Visualize
ggplot(stim_wide, aes(x = `Original scale`, y = `Rescaled`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Stimulus rankings preserved by rescaling")
```

### Compare outcome scales (pasted from claude just like that)

Binary model:   P(Y = 1)              → [0, 1] probability
Cosine model:   E(Y)                  → [0, 1] continuous
Ordinal model:  P(Y = 0,1,2,3)        → How to map to [0,1]?

Ordinal to continuous mapping

```{r}
# Helper function: extract expected value from ordinal model
get_ordinal_expected <- function(fit_ord, newdata) {
  # Get predicted probabilities for each category
  # fitted() for ordinal gives P(Y = k) for each k
  pred <- fitted(fit_ord, newdata = newdata, 
                 re_formula = ~(1|stimulus_concept),
                 summary = FALSE)  # Get all posterior draws
  
  # pred is array: draws × observations × categories
  # Dimensions: [n_draws, n_stimuli, 4 categories]
  
  # Compute expected value for each draw and stimulus
  # E(Y) = 0*P(Y=0) + 1*P(Y=1) + 2*P(Y=2) + 3*P(Y=3)
  expected_vals <- apply(pred, c(1, 2), function(probs) {
    sum(c(0, 1, 2, 3) * probs)
  })
  
   
  # Normalize to [0, 1]
  expected_vals_norm <- expected_vals / 3
  
  # Return summary statistics
  tibble(
    stimulus = newdata$stimulus_concept,
    Estimate = apply(expected_vals_norm, 2, mean),
    Q2.5 = apply(expected_vals_norm, 2, quantile, 0.025),
    Q97.5 = apply(expected_vals_norm, 2, quantile, 0.975)
    )
  }
  
  # Create newdata for all stimuli
  nd_stim <- data.frame(stimulus_concept = levels(df_brm$stimulus_concept))
  
  # Extract predictions from all three models
  stim_binary <- fitted(fit_bin, newdata = nd_stim,
                        re_formula = ~(1|stimulus_concept),
                        summary = TRUE) %>%
    as_tibble() %>%
    mutate(stimulus = nd_stim$stimulus_concept,
           measure = "Binary")
  
  stim_ordinal <- get_ordinal_expected(fit_ord, nd_stim) %>%
    mutate(measure = "Judged")
  
  stim_cosine <- fitted(fit_cos_com, newdata = nd_stim,
                        re_formula = ~(1|stimulus_concept),
                        summary = TRUE) %>%
    as_tibble() %>%
    mutate(stimulus = nd_stim$stimulus_concept,
           measure = "Cosine (rescaled)")
  
  # Combine all three
  all_measures <- bind_rows(
    stim_binary %>% select(stimulus, Estimate, Q2.5, Q97.5, measure),
    stim_ordinal %>% select(stimulus, Estimate, Q2.5, Q97.5, measure),
    stim_cosine %>% select(stimulus, Estimate, Q2.5, Q97.5, measure)
  ) %>%
    left_join(cats, by = c("stimulus" = "meaning"))
```

Visualization 1: Heatmap

```{r}
ggplot(all_measures, aes(x = measure, y = stimulus, fill = Estimate)) +
  geom_tile(color = "white") +
  facet_grid(category ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(limits = c(0, 1), 
                       name = "μ [0,1]",
                       option = "plasma") +
  labs(x = NULL, y = NULL,
       title = "Stimulus success across evaluation lenses",
       subtitle = "All measures normalized to [0,1] scale") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

Visualization 2: Lollipop Plot

```{r}
# Reorder stimuli by cosine performance for readability
stim_order <- all_measures %>%
  filter(measure == "Cosine (rescaled)") %>%
  arrange(Estimate) %>%
  pull(stimulus)

all_measures <- all_measures %>%
  mutate(stimulus = factor(stimulus, levels = stim_order),
         measure = factor(measure, 
                         levels = c("Binary", "Judged", "Cosine (rescaled)")))

ggplot(all_measures, aes(x = Estimate, y = stimulus, color = measure)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  geom_errorbarh(aes(xmin = Q2.5, xmax = Q97.5), 
                 height = 0.2,
                 position = position_dodge(width = 0.5)) +
  facet_wrap(~ category, scales = "free_y", ncol = 2) +
  scale_color_brewer(palette = "Dark2", name = "Evaluation") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Posterior mean μ [0,1]", y = NULL) +
  theme_minimal(base_size = 11)
```

Visualization 3: "Progressive Inclusivity" Plot
This directly shows your research question:

```{r}
# Count how many stimuli "succeed" under each threshold
thresholds <- seq(0.1, 0.9, 0.1)

success_counts <- map_df(thresholds, function(thresh) {
  all_measures %>%
    group_by(measure) %>%
    summarise(
      threshold = thresh,
      n_above = sum(Estimate > thresh),
      prop_above = mean(Estimate > thresh)
    )
})

ggplot(success_counts, aes(x = threshold, y = n_above, color = measure)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Dark2", name = "Evaluation lens") +
  scale_x_continuous(breaks = thresholds) +
  labs(x = "Success threshold (μ)", 
       y = "Number of stimuli above threshold") +
  theme_minimal(base_size = 12)
```

Visualization 4: Direct Comparison Grid

```{r}
# Advantage plot adapted for all three measures
all_measures_wide <- all_measures %>%
  select(stimulus, subcategory, measure, Estimate) %>%
  pivot_wider(names_from = measure, values_from = Estimate)

# Compute gaps
all_measures_wide <- all_measures_wide %>%
  mutate(
    `Judged - Binary` = Judged - Binary,
    `Cosine - Judged` = `Cosine (rescaled)` - Judged,
    `Cosine - Binary` = `Cosine (rescaled)` - Binary
  )

# Aggregate by subcategory
subcat_gaps <- all_measures_wide %>%
  group_by(subcategory) %>%
  summarise(
    n = n(),
    across(c(`Judged - Binary`, `Cosine - Judged`, `Cosine - Binary`), 
           mean, .names = "{.col}_mean")
  )

# Plot
subcat_gaps %>%
  pivot_longer(cols = ends_with("_mean"), 
               names_to = "contrast", 
               values_to = "gap") %>%
  mutate(contrast = str_remove(contrast, "_mean"),
         subcategory = fct_reorder(subcategory, gap, .fun = max)) %>%
  ggplot(aes(x = gap, y = subcategory, color = contrast)) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_point(aes(size = n), position = position_dodge(width = 0.4)) +
  geom_segment(aes(x = 0, xend = gap, y = subcategory, yend = subcategory),
               position = position_dodge(width = 0.4),
               linewidth = 0.8) +
  scale_color_brewer(palette = "Dark2", name = "Contrast") +
  scale_size_continuous(range = c(2, 6), name = "N stimuli") +
  labs(x = "Gap in μ (positive = second measure more permissive)",
       y = NULL)
```

Critical Code Block: Getting Ordinal Expected Values
If the above get_ordinal_expected() function doesn't work with your brms version, here's a more robust alternative:

```{r}
# Alternative method using posterior_predict
get_ordinal_expected_v2 <- function(fit_ord, newdata) {
  # Get posterior predicted categories for each draw
  pred_cats <- posterior_predict(fit_ord, 
                                 newdata = newdata,
                                 re_formula = ~(1|stimulus_concept))
  
  # pred_cats is matrix: draws × observations
  # Values are 0, 1, 2, or 3
  
  # Convert categories to [0,1] scale
  pred_norm <- pred_cats / 3
  
  # Summarize across draws
  tibble(
    stimulus = newdata$stimulus_concept,
    Estimate = apply(pred_norm, 2, mean),
    Q2.5 = apply(pred_norm, 2, quantile, 0.025),
    Q97.5 = apply(pred_norm, 2, quantile, 0.975)
  )
}

# Test both methods and compare
stim_ord_v1 <- get_ordinal_expected(fit_ord, nd_stim)
stim_ord_v2 <- get_ordinal_expected_v2(fit_ord, nd_stim)

# Should be highly correlated (might differ slightly)
cor(stim_ord_v1$Estimate, stim_ord_v2$Estimate)
```

# Session info

```{r echo=TRUE, message=FALSE, warning=FALSE}
sessionInfo()
```
