---
title: "Experiments in language evolution: inferences from closed- vs open-ended semantic space paradigms"
subtitle: "Data modeling"
author: "analysis by Aleksandra Ćwiek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_document:
    theme: readable
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '3'
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What this document does

**Purpose.** This notebook fits and evaluates **baseline hierarchical
models** for three outcome measures that capture how well participants’
free-text responses match target concepts:

1.  `binary_guess` — whether the response was judged exact/near-exact
    (yes/no),

2.  `judged_similarity` — coder-assigned similarity on a four-point
    scale,

3.  `cosine_rescaled` — automatic cosine similarity (ConceptNet,
    rescaled to [0,1]).

Each outcome is analyzed separately. We summarize posterior
distributions, report Bayesian R², and check convergence/diagnostics.

**Inputs.** One cleaned, analysis-ready dataset from `1_preparation.Rmd`
(`data/final_df.csv`), containing participant IDs, stimulus/concept IDs,
language, and the three outcome measures.

**Outputs.**

-   Cached fitted models in `models/` (RDS files).

-   Summary tables (posterior means/HDIs, random-effect SDs, Bayesian
    R²).

-   Diagnostic plots and posterior predictive checks.

-   Figures saved to `plots/2_modeling/` (auto-saved with stable
    numbering).

# Abstract

**Experiments in language evolution: inferences from closed- vs
open-ended semantic space paradigms**

Svetlana Kuleshova^\*1,2,^ Stefan Hartmann^3^, and Michael Pleyer^1^,
Marta Sibierska^1^, Johan Blomberg^4^, Przemysław Żywiczyński^1^,
Aleksandra Ćwiek^5,1^, Sławomir Wacewicz^1^

^\*^Corresponding Author:
[40010189\@parisnanterre.fr](mailto:40010189@parisnanterre.fr)

^1^Nicolaus Copernicus University in Toruń, Center for Language
Evolution Studies, Fosa Staromiejska 3, 87-100 Toruń, Poland.\
^2^ArScAn-Équipe AnTET (UMR 7041), CNRS, Université Paris Nanterre, 21
Allée de l'Université, 92000, Nanterre, France.\
^3^Heinrich Heine University Düsseldorf, Faculty of Arts and Humanities,
Universitätsstraße 1, 40225 Düsseldorf, Germany.\
^4^Stockholm University, Department of Swedish and Multilingualism.
Universitetsvägen 10, 106 91 Stockholm.\
^5^Leibniz-Centre General Linguistics, Berlin, 10719, Germany

We challenge the validity of the multiple-choice paradigm used in many
communication experiments. We replicated two experiments and replaced
the original multiple-choice design with free-text answers. Our results
indicate that participants performed worse compared to the original
studies. The post hoc analysis showed that often relevant semantic
domains were correctly identified, but full congruency with the target
concepts was rare. We discuss how the results of each paradigm can be
mapped onto a larger picture of language evolution.

What inferences about reality can we make from linguistic experiments?
By definition, controlled experiments do not represent reality in all
its complexity, and thus, it is important to be cautious when drawing
conclusions. As one example, some recent experimental studies have made
strong claims concluding that humans can understand improvised or
interspecies forms of communication, specifically, novel non-linguistic
vocalizations (Ćwiek et al., 2021) and gestures in non-human great apes
(Graham and Hobaiter, 2023). Here, we follow up on these studies,
altering one crucial aspect: Both original studies used a
multiple-choice paradigm, whereby participants were constrained to
choose one answer from a small set of alternatives; that is, for each
gesture or vocalisation they were offered a set of potential
interpretations and asked to guess the one best reflecting its intended
meaning. We expected that replacing the closed-ended, multiple-choice
paradigm with an open-ended, free-text paradigm would lead to different
results and thus different conclusions about human capacities to
‘understand’ ape gestures or novel vocalizations.

To test our hypothesis, we replicated the studies by Ćwiek et al. and
Graham & Hobaiter, but with the participants freely typing in their
answer (i.e., their interpretation of the meaning of an ape gesture or
novel vocalisation) rather than selecting it from a set of predefined
choices. To assess the proximity of the participants’ answers with the
target concepts, we used three complementary approaches: (i) binary
assessment of correctness (yes/no), (ii) manual coding on a four-point
scale, (iii) cosine similarity between the participants’ responses and
the target concepts using ConceptNet. For (i) and (ii), participants’
answers were coded by two raters on a four-point Likert scale
(transformed into a binary system for (i)), showing moderate-to-high
inter-rater agreement. All mismatches between coders were discussed in
the team and resolved. For (iii), the proximity between the words in the
responses (heads of phrases in the case of multi-word answers) and the
word denoting the target concept in ConceptNet (Speer et al. 2017) were
assessed automatically. Across all approaches, our results show that
participants performed qualitatively worse compared to the original
studies, with a vast majority of responses showing no resemblance
whatsoever to the target concept according to the manual coding
procedure. In the ConceptNet analysis, the responses were found to be
conceptually fairly close to the target concepts in many cases (e.g.,
the same semantic domain, as in ‘sleep’ vs ‘snore’), but hardly any
response was fully congruent with the target concept.

Apart from revisiting the question to what extent participants
“understand” novel vocalizations or ape gestures, our data give us the
opportunity to evaluate different possibilities of interpreting
participants’ free-text answers: manual coding with different levels of
granularity on the one hand, and automatic similarity detection using
NLP resources on the other. The three measures we used carve out
different semantic landscapes, shaping how we perceive communicative
transparency. A perfect-match approach treats meaning as a binary on/off
switch, limiting “understanding” to exact hits while discarding
near-misses as failures. The Likert-scale approach expands this space,
recognizing perceived shades of similarity. The ConceptNet-based cosine
similarity approach stretches the landscape even further, mapping words
onto a gradient of conceptual distances, where even broad associations
register as meaningful connections. Because these methods define
similarity differently, they potentially reveal different facets of how
novel signals convey meaning.

We thus conclude that different setups and analysis methods give us a
nuanced perspective on the role of semantic space in semiotic
experiments. These various paths to data collection and data treatment
can inform a nuanced discussion of how experimental results can inform
the evolution of today’s languages.

# Data preparation

## Source setup

```{r source setup, echo = TRUE, message=FALSE, warning = FALSE}
########## folders ##########
library(here)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

parentfolder <- dirname(getwd())

data          <- paste0(parentfolder, '/data/')
models        <- paste0(parentfolder, '/models/')
plots         <- paste0(parentfolder, '/plots/')
scripts       <- paste0(parentfolder, '/scripts/')

########## source file ##########

#source(paste0(scripts, "adjectives-preparation.R"))

#################### packages ####################
# Core tidyverse for wrangling + plotting; janitor/skimr for quick structure checks; patchwork for combining plots
library(tidyverse)
library(janitor)
library(skimr)
library(knitr)
library(cowplot)
library(ggdist)
library(readxl)
library(lubridate)
library(patchwork)
library(hexbin)
library(PRROC)

# Bayesian
library(brms);
library(loo);packageVersion("loo")
library(HDInterval);packageVersion("HDInterval")
library(cmdstanr);
library(emmeans);packageVersion("emmeans")
library(posterior);
library(bayesplot);
library(BayesFactor);
library(tidybayes);packageVersion("tidybayes")
library(bayestestR);
library(cmdstanr)
options(brms.backend = "cmdstanr") # Speed gain: ~30-50% faster than rstan
options(brms.file_refit = "on_change")  # only refit if model changes

# use all available cores for parallel computing
options(mc.cores = parallel::detectCores())

# Keep plots/readability consistent
theme_set(theme_minimal(base_size = 12) + panel_border())

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r echo = TRUE, message=FALSE, warning = FALSE}
# --- figure & table captions ---
outputFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to"); if (is.null(outputFormat)) outputFormat <- ""
capTabNo <- 1L; capFigNo <- 1L
.last_fig_no <- NA_integer_

capTab <- function(x){
  if (outputFormat == 'html'){
    x <- paste0("**Table ", capTabNo, ".** ", x)
    capTabNo <<- capTabNo + 1L
  }
  x
}

capFig <- function(x, show_R_version=FALSE, show_package_versions=NULL, is_map=FALSE){
  if (outputFormat == 'html'){
    assign(".last_fig_no", capFigNo, envir = .GlobalEnv) # remember current number
    x <- paste0("**Figure ", capFigNo, ".** ", x)
    capFigNo <<- capFigNo + 1L
  }
  x
}

# --- plot hook to auto-save ---
.default_plot_hook <- knitr::knit_hooks$get("plot")
.has_magick <- requireNamespace("magick", quietly = TRUE)

knitr::knit_hooks$set(plot = function(x, options){
  res <- if (is.function(.default_plot_hook)) .default_plot_hook(x, options) else NULL

  if (!is.null(options$fig.cap)) {
    fig_no <- get(".last_fig_no", envir = .GlobalEnv, inherits = FALSE)
    if (!is.na(fig_no)) {
      fig_no_str <- sprintf("%02d", fig_no)

      # subfolder = Rmd filename without extension
      rmd_name <- tools::file_path_sans_ext(basename(knitr::current_input()))
      target_dir <- file.path(plots, rmd_name)
      if (!dir.exists(target_dir)) dir.create(target_dir, recursive = TRUE)

      png_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".png"))
      tif_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".tif"))

      file.copy(x, png_out, overwrite = TRUE)

      if (.has_magick) {
        img <- magick::image_read(png_out)
        magick::image_write(img, path = tif_out, format = "tiff", compression = "lzw")
      } else {
        file.copy(png_out, tif_out, overwrite = TRUE)
      }
    }
  }

  res
})

```

Other helpers

```{r}
bb_corr <- function(x, y, R = 4000, seed = 1708) {
  stopifnot(length(x) == length(y))
  set.seed(seed)
  n <- length(x)
  out <- numeric(R)
  for (r in seq_len(R)) {
    w <- rexp(n)          # Dirichlet via i.i.d. Exponential then normalize
    w <- w / sum(w)
    mx <- sum(w * x); my <- sum(w * y)
    sx <- sqrt(sum(w * (x - mx)^2))
    sy <- sqrt(sum(w * (y - my)^2))
    out[r] <- sum(w * (x - mx) * (y - my)) / (sx * sy)
  }
  out
}

summ_post <- function(draws, name){
  h <- hdi(draws, ci = 0.95)
  tibble(pair = name,
         mean = mean(draws),
         median = median(draws),
         hdi_low = h$CI_low,
         hdi_high = h$CI_high)
}

bf_bic <- function(y, x){
  fit1 <- lm(y ~ x)
  fit0 <- lm(y ~ 1)
  BIC0 <- BIC(fit0); BIC1 <- BIC(fit1)
  # BF10 ≈ exp((BIC0 - BIC1)/2)
  exp((BIC0 - BIC1)/2)
}
```

## Load in data frames

```{r echo=TRUE, message=FALSE, warning=FALSE}
df    <- read_csv(paste0(data, "final_df.csv"))  %>% clean_names()
cats  <- read_csv(paste0(data, "meaning_categories.csv")) %>% clean_names()
```

# Correlations

```{r}
dat_cor <- df %>%
  select(cosine = cosine_rescaled, judged = ord_norm, binary = bin_f) %>%
  drop_na()
```

Pearson correlations (with CIs)

```{r}
pearson_mat <- cor(dat_cor, use = "pairwise.complete.obs", method = "pearson")
pearson_mat
```

Pairwise Pearson tests (p-values + CI)

```{r}
pearson_tests <- list(
  cosine_judged = cor.test(dat_cor$cosine, dat_cor$judged),
  cosine_binary = cor.test(dat_cor$cosine, dat_cor$binary),
  judged_binary = cor.test(dat_cor$judged, dat_cor$binary)
)
pearson_tests

pearson_tbl <- tibble(
  pair = c("cosine ~ judged", "cosine ~ binary", "judged ~ binary"),
  r = c(cor(dat_cor$cosine, dat_cor$judged),
        cor(dat_cor$cosine, dat_cor$binary),
        cor(dat_cor$judged, dat_cor$binary))
)
```

Bayesian correlations via BayesFactor (Point-biserial is just Pearson on
numeric 0/1; we still get Bayesian posterior for rho.)

```{r}
post_cos_jud <- bb_corr(dat_cor$cosine, dat_cor$judged, R = 8000, seed = 1708)
post_cos_bin <- bb_corr(dat_cor$cosine, dat_cor$binary, R = 8000, seed = 1708)
post_jud_bin <- bb_corr(dat_cor$judged, dat_cor$binary, R = 8000, seed = 1708)

bayes_corr_summary <- bind_rows(
  summ_post(post_cos_jud, "cosine ~ judged"),
  summ_post(post_cos_bin, "cosine ~ binary"),
  summ_post(post_jud_bin, "judged ~ binary")
)
bayes_corr_summary

# Posterior density plots
bind_rows(
  data.frame(rho = post_cos_jud, pair = "cosine~judged"),
  data.frame(rho = post_cos_bin, pair = "cosine~binary"),
  data.frame(rho = post_jud_bin, pair = "judged~binary")
) %>%
  ggplot(aes(rho)) + geom_density() + facet_wrap(~pair, scales = "free") +
  geom_vline(xintercept = 0, linetype = 2)
```

BIC-approx Bayes factors for "correlation != 0" For
continuous-continuous: linear model; for (cont \~ binary): also linear
is fine here.

```{r}
BF_cos_jud <- bf_bic(dat_cor$cosine, dat_cor$judged)
BF_cos_bin <- bf_bic(dat_cor$cosine, dat_cor$binary)
BF_jud_bin <- bf_bic(dat_cor$judged, dat_cor$binary)

bf_tbl <- tibble(
  pair = c("cosine ~ judged", "cosine ~ binary", "judged ~ binary"),
  BF10 = c(BF_cos_jud, BF_cos_bin, BF_jud_bin)
)
```

One clean summary table

```{r}
pearson_tbl %>%
  left_join(bayes_corr_summary, by = "pair") %>%
  left_join(bf_tbl, by = "pair")
```

# Variance of random factors

Explore the variance across all possible random factors:

-   Environment

-   Input device

-   Output device

-   Native language

-   Participant

-   Stimulus/concept

-   Hearing disabilities

# Modeling

## What we model

-   **Binary accuracy**: `binary_guess ∈ {0,1}`, modeled with a
    **Bernoulli** distribution and logit link.

-   **Ordinal judgments**: `judged_similarity ∈ {0,1,2,3}`, modeled with
    a **cumulative logit (ordinal)** likelihood.

-   **Cosine similarity**: `cosine_rescaled ∈ [0,1]`, modeled with a
    **zero-one-inflated Beta (ZOIB)** or a simpler **Beta** after
    applying an ε-open transform to handle boundary values.

All models include varying intercepts for participants and stimuli to
account for individual and item-level heterogeneity.

## Full (for Cognitive Science)

We fit baseline models for each of the three measures, with only
random intercepts for participant, native language, and stimulus.

Full solution. (Beware: It might take days.)

```{r eval=FALSE, message=FALSE, warning=FALSE}
{
  set.seed(1708)

  ## ---- ensure models dir exists (you already have `models` as an absolute path) ----
  if (!dir.exists(models)) dir.create(models, recursive = TRUE, showWarnings = FALSE)

  ## ---- data prep (minimal; no z-scoring of cosine) ----
  ## ---------- COMPLETE-CASE DATA (based on cosine)
  # Rationale: use the same rows for all three outcomes to enable direct comparability
  if (!"ord_norm" %in% names(df)) df$ord_norm <- df$judged_similarity / 3
  df_brm <- subset(df, !is.na(cosine_rescaled))

  ## Keep exact-1 spike info for reporting (from original column but CC rows)
  ones_table_stim <- aggregate(I(df_brm$cosine_rescaled == 1) ~ stimulus_concept, data = df_brm, FUN = mean)
  names(ones_table_stim) <- c("stimulus_concept", "one_rate")

  df_brm$participant      <- factor(df_brm$participant)
  df_brm$stimulus_concept <- factor(df_brm$stimulus_concept)
  df_brm$native_language  <- factor(df_brm$native_language) # decided to do without it because it was not variable and caused much longer computation times
  df_brm$judged_ord       <- factor(df_brm$judged_similarity, levels = 0:3, ordered = TRUE)
 # Replace values extremely close to 1 with exactly 1
  df_brm$cosine_rescaled <- ifelse(
    !is.na(df_brm$cosine_rescaled) & df_brm$cosine_rescaled > 1 & df_brm$cosine_rescaled <
      1.001,
    1,
    df_brm$cosine_rescaled
  )
  
  ## ---- cosine preparation (rescaling to make comparable across measures) ----
  
  rescale_to_unit <- function(x) {
    eps <- 1e-10
    tiny <- 1e-12
    
    # Step 1: Epsilon-open to avoid exact boundaries
    x_open <- pmin(pmax(x - eps, tiny), 1 - tiny)
    
    # Step 2: Min-max rescale to [0,1]
    xmin <- min(x_open, na.rm = TRUE)
    xmax <- max(x_open, na.rm = TRUE)
    span <- xmax - xmin
    
    if (!is.finite(span) || span <= 0) {
      stop("Cannot rescale: no variation in x")
    }
    
    x_scaled <- (x_open - xmin) / span
    
    # Step 3: Ensure strictly in (0,1) for Beta
    x_final <- pmin(pmax(x_scaled, tiny), 1 - tiny)
    
    return(x_final)
  }
  
  # Apply to data
  # Nudge exact 1.0s only (keeping original scale)
  df_brm$cosine_original <- df_brm$cosine_rescaled
  df_brm$cosine_original[df_brm$cosine_original == 1.0] <- 0.999999
  
  # Full min-max rescaling (for comparable plot)
  df_brm$cosine_scaled <- rescale_to_unit(df_brm$cosine_rescaled)
  
  ## ---- very wide priors for baselines / couplings (not for ZOIB components) ----
  .pri_bin <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )
  .pri_ord <- c(
    set_prior("normal(0, 5)", class = "Intercept"),  # thresholds
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )
  .pri_wide <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )

  ## ========================= BASELINES =========================
  .bin_rds <- file.path(models, "fit_bin.rds")
  if (file.exists(.bin_rds)) {
    fit_bin <- readRDS(.bin_rds)
  } else {
    fit_bin <- brm(
      binary_guess ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = bernoulli(),
      prior   = .pri_bin,
      chains  = 4, iter = 8000, warmup = 4000, threads = threading(2),
      control = list(adapt_delta = 0.97, max_treedepth = 12),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_bin, .bin_rds)
  }

  .ord_rds <- file.path(models, "fit_ord.rds")
  if (file.exists(.ord_rds)) {
    fit_ord <- readRDS(.ord_rds)
  } else {
    fit_ord <- brm(
      judged_ord ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = cumulative(link = "logit"),
      prior   = .pri_ord,
      chains  = 4, iter = 10000, warmup = 4000, threads = threading(2),
      control = list(adapt_delta = 0.99, max_treedepth = 13),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_ord, .ord_rds)
  }

  .cos_org_rds <- file.path(models, "fit_cos_org.rds")
  if (file.exists(.cos_org_rds)) {
    fit_cos_org <- readRDS(.cos_org_rds)
  } else {
    fit_cos_org <- brm(
      cosine_original ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = Beta(), 
      prior = c(
        set_prior("normal(0, 2)", class = "Intercept"),
        set_prior("student_t(3, 0, 2.5)", class = "sd"),
        set_prior("gamma(0.01, 0.01)", class = "phi")),
      chains  = 4, iter = 8000, warmup = 4000, threads = threading(2),
      control = list(adapt_delta = 0.99, max_treedepth = 13),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_cos_org, .cos_org_rds)
  }
  
  .cos_com_rds <- file.path(models, "fit_cos_com.rds")
  if (file.exists(.cos_com_rds)) {
    fit_cos_com <- readRDS(.cos_com_rds)
  } else {
    fit_cos_com <- brm(
      cosine_scaled ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = Beta(), 
      prior = c(
        set_prior("normal(0, 2)", class = "Intercept"),
        set_prior("student_t(3, 0, 2.5)", class = "sd"),
        set_prior("gamma(0.01, 0.01)", class = "phi")),
      chains  = 4, iter = 8000, warmup = 4000, threads = threading(2),
      control = list(adapt_delta = 0.99, max_treedepth = 13),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_cos_com, .cos_com_rds)
  }

  ## ---- compact Bayes R^2 (kept as r2_table) ----
  .r2 <- function(fit) { r <- bayes_R2(fit); c(Estimate = mean(r), Q2.5 = quantile(r,.025), Q97.5 = quantile(r,.975)) }
  r2_table <- data.frame(
    model = c("Binary (Bernoulli)", "Judged (Ordinal)", "Cosine (Beta)"),
    rbind(.r2(fit_bin), .r2(fit_ord), .r2(fit_cos_org))
  )
  rownames(r2_table) <- NULL
  r2_table$Estimate <- round(r2_table$Estimate, 3)
  r2_table$Q2.5     <- round(r2_table$Q2.5, 3)
  r2_table$Q97.5    <- round(r2_table$Q97.5, 3)
  print(r2_table)
  

  ## ---- cleanup helpers (keep only model objects + r2_table + path vars) ----
  rm(list = c(".pri_bin",".pri_ord",".pri_wide",
              ".bin_rds",".ord_rds",".cos_org_rds", ".cos_com_rds",
              ".m_ord_cos_rds",".m_bin_cos_rds",".m_bin_jud_rds",".m_cos_jud_rds"))
}
```

## Extract results

Compare the cosine similarity models on two scales

```{r}
# Extract stimulus posteriors from both models
stim_original <- fitted(fit_cos_org, 
                       newdata = data.frame(stimulus_concept = levels(df_brm$stimulus_concept)),
                       re_formula = ~(1|stimulus_concept))

stim_scaled <- fitted(fit_cos_com,
                     newdata = data.frame(stimulus_concept = levels(df_brm$stimulus_concept)),
                     re_formula = ~(1|stimulus_concept))

# Perfect rank preservation
cor(stim_original[,"Estimate"], stim_scaled[,"Estimate"], method = "spearman")
# Should be 1.0
```

Verify Rank Concordance

```{r}
# Extract stimulus-level predictions from both
get_stim_preds <- function(fit, label) {
  nd <- data.frame(stimulus_concept = levels(df_brm$stimulus_concept))
  preds <- fitted(fit, newdata = nd, 
                  re_formula = ~(1|stimulus_concept), 
                  summary = TRUE)
  tibble(
    stimulus = nd$stimulus_concept,
    mean = preds[, "Estimate"],
    model = label
  )
}

stim_both <- bind_rows(
  get_stim_preds(fit_cos_original, "Original scale"),
  get_stim_preds(fit_cos_scaled, "Rescaled")
)

# Rank correlation (should be perfect)
stim_wide <- stim_both %>%
  pivot_wider(names_from = model, values_from = mean)

cor(stim_wide$`Original scale`, stim_wide$`Rescaled`, method = "spearman")
# Should be 1.0 - perfect rank preservation

# Visualize
ggplot(stim_wide, aes(x = `Original scale`, y = `Rescaled`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Stimulus rankings preserved by rescaling")
```

### Compare outcome scales (pasted from claude just like that)

Binary model:   P(Y = 1)              → [0, 1] probability
Cosine model:   E(Y)                  → [0, 1] continuous
Ordinal model:  P(Y = 0,1,2,3)        → How to map to [0,1]?

Ordinal to continuous mapping

```{r}
# Helper function: extract expected value from ordinal model
get_ordinal_expected <- function(fit_ord, newdata) {
  # Get predicted probabilities for each category
  # fitted() for ordinal gives P(Y = k) for each k
  pred <- fitted(fit_ord, newdata = newdata, 
                 re_formula = ~(1|stimulus_concept),
                 summary = FALSE)  # Get all posterior draws
  
  # pred is array: draws × observations × categories
  # Dimensions: [n_draws, n_stimuli, 4 categories]
  
  # Compute expected value for each draw and stimulus
  # E(Y) = 0*P(Y=0) + 1*P(Y=1) + 2*P(Y=2) + 3*P(Y=3)
  expected_vals <- apply(pred, c(1, 2), function(probs) {
    sum(c(0, 1, 2, 3) * probs)
  })
  
   
  # Normalize to [0, 1]
  expected_vals_norm <- expected_vals / 3
  
  # Return summary statistics
  tibble(
    stimulus = newdata$stimulus_concept,
    Estimate = apply(expected_vals_norm, 2, mean),
    Q2.5 = apply(expected_vals_norm, 2, quantile, 0.025),
    Q97.5 = apply(expected_vals_norm, 2, quantile, 0.975)
    )
  }
  
  # Create newdata for all stimuli
  nd_stim <- data.frame(stimulus_concept = levels(df_brm$stimulus_concept))
  
  # Extract predictions from all three models
  stim_binary <- fitted(fit_bin, newdata = nd_stim,
                        re_formula = ~(1|stimulus_concept),
                        summary = TRUE) %>%
    as_tibble() %>%
    mutate(stimulus = nd_stim$stimulus_concept,
           measure = "Binary")
  
  stim_ordinal <- get_ordinal_expected(fit_ord, nd_stim) %>%
    mutate(measure = "Judged")
  
  stim_cosine <- fitted(fit_cos_scaled, newdata = nd_stim,
                        re_formula = ~(1|stimulus_concept),
                        summary = TRUE) %>%
    as_tibble() %>%
    mutate(stimulus = nd_stim$stimulus_concept,
           measure = "Cosine (rescaled)")
  
  # Combine all three
  all_measures <- bind_rows(
    stim_binary %>% select(stimulus, Estimate, Q2.5, Q97.5, measure),
    stim_ordinal %>% select(stimulus, Estimate, Q2.5, Q97.5, measure),
    stim_cosine %>% select(stimulus, Estimate, Q2.5, Q97.5, measure)
  ) %>%
    left_join(cats, by = c("stimulus" = "meaning"))
```

Visualization 1: Heatmap

```{r}
ggplot(all_measures, aes(x = measure, y = stimulus, fill = Estimate)) +
  geom_tile(color = "white") +
  facet_grid(category ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(limits = c(0, 1), 
                       name = "μ [0,1]",
                       option = "plasma") +
  labs(x = NULL, y = NULL,
       title = "Stimulus success across evaluation lenses",
       subtitle = "All measures normalized to [0,1] scale") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

Visualization 2: Lollipop Plot

```{r}
# Reorder stimuli by cosine performance for readability
stim_order <- all_measures %>%
  filter(measure == "Cosine (rescaled)") %>%
  arrange(Estimate) %>%
  pull(stimulus)

all_measures <- all_measures %>%
  mutate(stimulus = factor(stimulus, levels = stim_order),
         measure = factor(measure, 
                         levels = c("Binary", "Judged", "Cosine (rescaled)")))

ggplot(all_measures, aes(x = Estimate, y = stimulus, color = measure)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  geom_errorbarh(aes(xmin = Q2.5, xmax = Q97.5), 
                 height = 0.2,
                 position = position_dodge(width = 0.5)) +
  facet_wrap(~ category, scales = "free_y", ncol = 2) +
  scale_color_brewer(palette = "Dark2", name = "Evaluation lens") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Posterior mean μ [0,1]", y = NULL,
       title = "Three lenses reveal nested semantic landscapes") +
  theme_minimal(base_size = 11)
```

Visualization 3: "Progressive Inclusivity" Plot
This directly shows your research question:

```{r}
# Count how many stimuli "succeed" under each threshold
thresholds <- seq(0.1, 0.9, 0.1)

success_counts <- map_df(thresholds, function(thresh) {
  all_measures %>%
    group_by(measure) %>%
    summarise(
      threshold = thresh,
      n_above = sum(Estimate > thresh),
      prop_above = mean(Estimate > thresh)
    )
})

ggplot(success_counts, aes(x = threshold, y = n_above, color = measure)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Dark2", name = "Evaluation lens") +
  scale_x_continuous(breaks = thresholds) +
  labs(x = "Success threshold (μ)", 
       y = "Number of stimuli above threshold",
       title = "Cosine lens is most permissive, binary most restrictive",
       subtitle = "More stimuli 'succeed' under distributional similarity") +
  theme_minimal(base_size = 12)
```

Visualization 4: Direct Comparison Grid

```{r}
# Advantage plot adapted for all three measures
all_measures_wide <- all_measures %>%
  select(stimulus, subcategory, measure, Estimate) %>%
  pivot_wider(names_from = measure, values_from = Estimate)

# Compute gaps
all_measures_wide <- all_measures_wide %>%
  mutate(
    `Judged - Binary` = Judged - Binary,
    `Cosine - Judged` = `Cosine (rescaled)` - Judged,
    `Cosine - Binary` = `Cosine (rescaled)` - Binary
  )

# Aggregate by subcategory
subcat_gaps <- all_measures_wide %>%
  group_by(subcategory) %>%
  summarise(
    n = n(),
    across(c(`Judged - Binary`, `Cosine - Judged`, `Cosine - Binary`), 
           mean, .names = "{.col}_mean")
  )

# Plot
subcat_gaps %>%
  pivot_longer(cols = ends_with("_mean"), 
               names_to = "contrast", 
               values_to = "gap") %>%
  mutate(contrast = str_remove(contrast, "_mean"),
         subcategory = fct_reorder(subcategory, gap, .fun = max)) %>%
  ggplot(aes(x = gap, y = subcategory, color = contrast)) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_point(aes(size = n), position = position_dodge(width = 0.4)) +
  geom_segment(aes(x = 0, xend = gap, y = subcategory, yend = subcategory),
               position = position_dodge(width = 0.4),
               linewidth = 0.8) +
  scale_color_brewer(palette = "Dark2", name = "Contrast") +
  scale_size_continuous(range = c(2, 6), name = "N stimuli") +
  labs(x = "Gap in μ (positive = second measure more permissive)",
       y = NULL,
       title = "Progressive inclusivity across evaluation lenses")
```

Critical Code Block: Getting Ordinal Expected Values
If the above get_ordinal_expected() function doesn't work with your brms version, here's a more robust alternative:
```{r}
# Alternative method using posterior_predict
get_ordinal_expected_v2 <- function(fit_ord, newdata) {
  # Get posterior predicted categories for each draw
  pred_cats <- posterior_predict(fit_ord, 
                                 newdata = newdata,
                                 re_formula = ~(1|stimulus_concept))
  
  # pred_cats is matrix: draws × observations
  # Values are 0, 1, 2, or 3
  
  # Convert categories to [0,1] scale
  pred_norm <- pred_cats / 3
  
  # Summarize across draws
  tibble(
    stimulus = newdata$stimulus_concept,
    Estimate = apply(pred_norm, 2, mean),
    Q2.5 = apply(pred_norm, 2, quantile, 0.025),
    Q97.5 = apply(pred_norm, 2, quantile, 0.975)
  )
}

# Test both methods and compare
stim_ord_v1 <- get_ordinal_expected(fit_ord, nd_stim)
stim_ord_v2 <- get_ordinal_expected_v2(fit_ord, nd_stim)

# Should be highly correlated (might differ slightly)
cor(stim_ord_v1$Estimate, stim_ord_v2$Estimate)
```


## Compromise (for Protolang)

Compromise to meet computational time.

-   We keep the 1’s from cosine similarity conceptually (they’re just ε
    below 1 for Beta math), preserving order and near-perfect matches.
    Also rescaling to the true 0-1 range (not positively skewed with
    much higher mean due to the nature of thecosine similarity scale).

-   **All three outcomes on comparable 0–1 scales (binary 0/1, judged
    0–1, cosine ε-open (0,1))**, so the per-stimulus posterior means can
    be min–max rescaled to overlay the three “landscapes” directly (same
    axis).

-   Fast & robust: **Bernoulli + Gaussian + Beta** are far lighter than
    cumulative + ZOIB at N=25k, but they still answer the main question:
    the measures are related but not identical, visible in the differing
    shapes/ranks of stim_landscapes_scaled.

-   **Priors:** Weakly informative on intercepts and group-level SDs
    (stabilize estimates, avoid overfitting).

-   **Sampling:** 4 chains, 8000 iterations (4000 warmup), with
    `adapt_delta = 0.98–0.99` and `max_treedepth = 12–13` for
    robustness.

```{r echo=TRUE, message=FALSE, warning=FALSE}
{
  set.seed(1708)

  ## ---------- ensure models dir exists ----------
  if (!dir.exists(models)) dir.create(models, recursive = TRUE, showWarnings = FALSE)

  ## ---------- COMPLETE-CASE DATA (based on cosine) ----------
  # Rationale: use the same rows for all three outcomes to enable direct comparability
  df_cc <- subset(df, !is.na(cosine_rescaled))
  if (!"ord_norm" %in% names(df_cc)) df_cc$ord_norm <- df_cc$judged_similarity / 3

  df_cc$participant      <- factor(df_cc$participant)
  df_cc$stimulus_concept <- factor(df_cc$stimulus_concept)

  ## Keep exact-1 spike info for reporting (from original column but CC rows)
  ones_table_stim <- aggregate(I(df_cc$cosine_rescaled == 1) ~ stimulus_concept, data = df_cc, FUN = mean)
  names(ones_table_stim) <- c("stimulus_concept", "one_rate")

  ## ---------- Cosine transform: epsilon-open + unit-stretch ----------
  eps  <- 1e-10
  tiny <- 1e-12
  x    <- df_cc$cosine_rescaled
  x_open <- pmin(pmax(x - eps, tiny), 1 - tiny)
  xmin <- min(x_open, na.rm = TRUE); xmax <- max(x_open, na.rm = TRUE); span <- xmax - xmin
  if (!is.finite(span) || span <= 0) stop("Cosine span after opening is zero; cannot scale.")
  df_cc$cosine_unit_open <- pmin(pmax((x_open - xmin) / span, tiny), 1 - tiny)

  ## ---------- very wide priors (intercept + RE sd only) ----------
  .pri_intercepts_only <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )

  ## =================================================================
  ##                           BASELINES (fast, CC)
  ## =================================================================
  ctrl_strict <- list(adapt_delta = 0.99, max_treedepth = 13)

  ## 1) Binary (Bernoulli) on df_cc
  .bin_cc_rds <- file.path(models, "fit_bin_cc.rds")
  if (file.exists(.bin_cc_rds)) {
    fit_bin_cc <- readRDS(.bin_cc_rds)
  } else {
    fit_bin_cc <- brm(
      binary_guess ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_cc,
      family  = bernoulli(),
      prior   = .pri_intercepts_only,
      chains  = 4, iter = 8000, warmup = 4000,
      control = ctrl_strict,
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_bin_cc, .bin_cc_rds, compress = "xz")
  }

  ## 2) Judged (Gaussian 0–1) on df_cc
  .ord_cc_rds <- file.path(models, "fit_ord_gauss_cc.rds")
  if (file.exists(.ord_cc_rds)) {
    fit_ord_gauss_cc <- readRDS(.ord_cc_rds)
  } else {
    fit_ord_gauss_cc <- brm(
      ord_norm ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_cc,
      family  = gaussian(),
      prior   = .pri_intercepts_only,
      chains  = 4, iter = 8000, warmup = 4000,
      control = ctrl_strict,
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_ord_gauss_cc, .ord_cc_rds, compress = "xz")
  }

  ## 3) Cosine (Beta on unit-stretched) on df_cc
  .cos_cc_rds <- file.path(models, "fit_cos_beta_unit_cc.rds")
  if (file.exists(.cos_cc_rds)) {
    fit_cos_beta_cc <- readRDS(.cos_cc_rds)
  } else {
    fit_cos_beta_cc <- brm(
      cosine_unit_open ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_cc,
      family  = Beta(),
      prior   = .pri_intercepts_only,
      chains  = 4, iter = 8000, warmup = 4000,
      control = ctrl_strict,
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_cos_beta_cc, .cos_cc_rds, compress = "xz")
  }

  ## ================== ROBUST R² (cache to RDS) ==================
  .recalc_r2 <- FALSE  # set TRUE to force recomputation

  .r2_tbl_rds <- file.path(models, "r2_table_cc.rds")
  .r2_bin_rds <- file.path(models, "r2_draws_bin_cc.rds")
  .r2_ord_rds <- file.path(models, "r2_draws_ord_cc.rds")
  .r2_cos_rds <- file.path(models, "r2_draws_cos_cc.rds")

  if (!.recalc_r2 &&
      file.exists(.r2_tbl_rds) &&
      file.exists(.r2_bin_rds) &&
      file.exists(.r2_ord_rds) &&
      file.exists(.r2_cos_rds)) {

    r2_bin <- readRDS(.r2_bin_rds)
    r2_ord <- readRDS(.r2_ord_rds)
    r2_cos <- readRDS(.r2_cos_rds)
    r2_table_cc <- readRDS(.r2_tbl_rds)

  } else {

    r2_bin <- bayes_R2(fit_bin_cc)       # numeric vector of draws
    r2_ord <- bayes_R2(fit_ord_gauss_cc)
    r2_cos <- bayes_R2(fit_cos_beta_cc)

    saveRDS(r2_bin, .r2_bin_rds)
    saveRDS(r2_ord, .r2_ord_rds)
    saveRDS(r2_cos, .r2_cos_rds)

    .summ <- function(x) c(Estimate = mean(x),
                           `Q2.5`   = as.numeric(quantile(x, .025)),
                           `Q97.5`  = as.numeric(quantile(x, .975)))
    r2_table_cc <- rbind(.summ(r2_bin), .summ(r2_ord), .summ(r2_cos))
    r2_table_cc <- as.data.frame(r2_table_cc)
    r2_table_cc <- tibble::rownames_to_column(r2_table_cc, "Model")
    r2_table_cc <- dplyr::mutate(r2_table_cc, dplyr::across(where(is.numeric), ~round(.x, 3)))
    r2_table_cc$Model <- c("Binary (Bernoulli, CC)",
                           "Judged (Gaussian 0–1, CC)",
                           "Cosine (Beta, unit-stretched, CC)")

    saveRDS(r2_table_cc, .r2_tbl_rds)
  }

  print(r2_table_cc)

  ## ================== PER-STIMULUS POSTERIOR MEANS (CC) ==================
  stim_levels_cc <- levels(df_cc$stimulus_concept)
  new_stim_cc <- data.frame(stimulus_concept = stim_levels_cc)

  # Binary: marginal over participant; keep only stimulus RE
  bin_pred <- as.data.frame(
    fitted(fit_bin_cc, newdata = new_stim_cc, re_formula = ~(1|s|stimulus_concept), summary = TRUE)
  )
  bin_stim <- data.frame(stimulus_concept = stim_levels_cc, bin_mean = bin_pred$Estimate)

  # Judged (Gaussian)
  ord_pred <- as.data.frame(
    fitted(fit_ord_gauss_cc, newdata = new_stim_cc, re_formula = ~(1|s|stimulus_concept), summary = TRUE)
  )
  ord_stim <- data.frame(stimulus_concept = stim_levels_cc, ord_mean = ord_pred$Estimate)

  # Cosine (Beta)
  cos_pred <- as.data.frame(
    fitted(fit_cos_beta_cc, newdata = new_stim_cc, re_formula = ~(1|s|stimulus_concept), summary = TRUE)
  )
  cos_stim <- data.frame(stimulus_concept = stim_levels_cc, cos_mean = cos_pred$Estimate)

  # Merge + add spike info (computed on CC rows)
  suppressWarnings({
    stim_landscapes_cc <- dplyr::left_join(bin_stim, ord_stim, by = "stimulus_concept") |>
      dplyr::left_join(cos_stim, by = "stimulus_concept") |>
      dplyr::left_join(ones_table_stim, by = "stimulus_concept")
  })

  # Min–max rescale summaries (for plotting; does not affect models)
  mm <- function(v) (v - min(v, na.rm = TRUE)) / (max(v, na.rm = TRUE) - min(v, na.rm = TRUE))
  stim_landscapes_cc_scaled <- transform(
    stim_landscapes_cc,
    bin_mm = mm(bin_mean),
    ord_mm = mm(ord_mean),
    cos_mm = mm(cos_mean)
  )

  ## ---------- keep useful objects ----------
  assign("df_cc", df_cc, inherits = FALSE)
  assign("fit_bin_cc", fit_bin_cc, inherits = FALSE)
  assign("fit_ord_gauss_cc", fit_ord_gauss_cc, inherits = FALSE)
  assign("fit_cos_beta_cc", fit_cos_beta_cc, inherits = FALSE)
  assign("r2_table_cc", r2_table_cc, inherits = FALSE)
  assign("r2_draws_cc", list(Binary = r2_bin, Judged = r2_ord, Cosine = r2_cos), inherits = FALSE)
  assign("stim_landscapes_cc", stim_landscapes_cc, inherits = FALSE)
  assign("stim_landscapes_cc_scaled", stim_landscapes_cc_scaled, inherits = FALSE)

  ## ---------- cleanup helpers ----------
  rm(.pri_intercepts_only, .bin_cc_rds, .ord_cc_rds, .cos_cc_rds,
     .r2_tbl_rds, .r2_bin_rds, .r2_ord_rds, .r2_cos_rds,
     .r2_summ, stim_levels_cc, new_stim_cc, bin_pred, ord_pred, cos_pred, mm,
     ones_table_stim, eps, tiny, x, x_open, xmin, xmax, span, ctrl_strict)
}
```

### Extract results

Bayesian R² for each fitted model using bayes_R2(). This is a quick,
comparable goodness-of-fit summary.

```{r echo=TRUE, message=FALSE, warning=FALSE}
r2_table_cc
```

Posterior predictive checks for all three models, shown side-by-side for
quick visual comparison. If patchwork is available we combine them
horizontally; otherwise they will print one after another.

```{r echo=TRUE, message=FALSE, warning=FALSE}
suppressWarnings(has_patchwork <- requireNamespace("patchwork", quietly = TRUE))

p_bin  <- pp_check(fit_bin_cc,       ndraws = 200) + ggtitle("Binary (Bernoulli)")
p_gaus <- pp_check(fit_ord_gauss_cc, ndraws = 200) + ggtitle("Judged (Gaussian 0–1)")
p_beta <- pp_check(fit_cos_beta_cc,  ndraws = 200) + ggtitle("Cosine (Beta)")

if (has_patchwork) {
  patchwork::wrap_plots(p_bin, p_gaus, p_beta, ncol = 3)
} else {
  print(p_bin); print(p_gaus); print(p_beta)
}
```

### Posteriors visualized

Goal of this section:

-   Extract posterior draws of the *population-level* mean (μ) for each
    model, on the OUTCOME scale (probability for Bernoulli, 0–1 scale
    for Gaussian/Beta).

-   Summarize μ for each model (mean/median/95% CrI) for the manuscript
    table.

-   Compute pairwise posterior probabilities that one model’s μ exceeds
    another, plus the posterior on their differences.

Why re_formula = NA?

It drops all random effects so we get the population-level intercept
(μ), i.e., the grand mean across participants and stimuli on the outcome
scale. We pass a single valid row from df_cc as 'newdata' just to anchor
factor levels; with re_formula = NA that row's values don't affect μ.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Descriptive posterior summaries for intercepts (μ) + pairwise probabilities

# --- Helper: posterior draws of μ on outcome scale (drop all REs)
.post_mu_draws <- function(fit, label, ref_row = 1) {
  # We ask brms for the mean of the likelihood (fitted), not raw draws of y,
  # and we want *all draws*, hence summary = FALSE.
  # For Bernoulli this is P(Y=1); for Gaussian(0–1) it is the expected value;
  # for Beta it is the expected mean on the 0–1 scale.
  mu <- fitted(fit, newdata = df_cc[ref_row, , drop = TRUE],
               re_formula = NA, summary = FALSE)
  tibble(model = label, mu = as.numeric(mu))
}

# --- Get/assemble draws (ordered Binary -> Judged -> Cosine)
mu_draws <- bind_rows(
  .post_mu_draws(fit_bin_cc,       "Binary guess"),
  .post_mu_draws(fit_ord_gauss_cc, "Judged similarity"),
  .post_mu_draws(fit_cos_beta_cc,  "Cosine similarity")
) %>%
  mutate(model = factor(model,
                        levels = c("Binary guess", "Judged similarity", "Cosine similarity")))

# --- Manuscript-ready summary of μ (posterior mean, median, 95% CrI)
# We summarize across posterior draws within each model. Numbers are on the
# outcome scale, so the three μ's are directly comparable as "average level"
# of each metric (e.g., average hit rate, average judged similarity 0–1, etc.).
mu_summary <- mu_draws %>%
  group_by(model) %>%
  summarise(
    mean   = mean(mu),
    median = median(mu),
    `Q2.5` = quantile(mu, .025),
    `Q97.5`= quantile(mu, .975),
    .groups = "drop"
  )

# Optional: friendly formatting for tiny numbers
fmt_num <- function(x) {
  out <- ifelse(x < 0.001, formatC(x, format = "e", digits = 2),
                ifelse(x < 0.01,  formatC(x, format = "f", digits = 3),
                                 formatC(x, format = "f", digits = 3)))
  as.character(out)
}

mu_summary_print <- mu_summary %>%
  mutate(across(where(is.numeric), fmt_num)) %>%
  rename(`Model` = model,
         `Mean μ` = mean, `Median μ` = median,
         `2.5%` = `Q2.5`, `97.5%` = `Q97.5`)

cat("\nIntercept (population mean μ on outcome scale) — posterior summaries:\n")
knitr::kable(mu_summary_print, align = "lcccc")

# --- Pairwise comparisons between μ's
# We compare the *same-iteration* scale (posterior draws) of the μ's across models.
# For each pair we compute:
#   - P(Second > First): posterior probability that μ_second exceeds μ_first
#   - Posterior on the difference (Second - First): median and 95% CrI
# Interpretation: This is *not* a null-hypothesis test; it’s a direct Bayesian
# probability statement about contrasts of population means on the outcome scale.
set.seed(1708)

get_draws_for <- function(lbl) filter(mu_draws, model == lbl) %>% pull(mu)

d_bin  <- get_draws_for("Binary guess")
d_judg <- get_draws_for("Judged similarity")
d_cos  <- get_draws_for("Cosine similarity")

# We align lengths by random pairing (if needed) to avoid recycling; this keeps
# contrasts unbiased even if the upstream models had different numbers of draws.
pair_prob <- function(x, y, n = 10000L) {
  n <- min(length(x), length(y), n)
  ix <- sample(seq_along(x), n, replace = length(x) < n)
  iy <- sample(seq_along(y), n, replace = length(y) < n)
  dx <- y[iy] - x[ix]               # positive => y > x
  tibble(
    n = n,
    diff_median = median(dx),
    diff_q2.5   = quantile(dx, .025),
    diff_q97.5  = quantile(dx, .975),
    P_y_gt_x    = mean(dx > 0)
  )
}

cmp_list <- list(
  `Judged − Binary` = pair_prob(d_bin, d_judg),
  `Cosine − Binary` = pair_prob(d_bin, d_cos),
  `Cosine − Judged` = pair_prob(d_judg, d_cos)
)

pairwise_summary <- bind_rows(cmp_list, .id = "Comparison")

# Friendly printout
pairwise_print <- pairwise_summary %>%
  mutate(
    across(c(diff_median, diff_q2.5, diff_q97.5), fmt_num),
    `P(Second > First)` = formatC(P_y_gt_x, format = "f", digits = 3)
  ) %>%
  select(Comparison,
         `P(Second > First)`,
         `Diff median` = diff_median,
         `Diff 2.5%`   = diff_q2.5,
         `Diff 97.5%`  = diff_q97.5,
         n)

cat("\nPairwise comparisons of intercepts μ (posterior on differences and probabilities):\n")
knitr::kable(pairwise_print, align = "lccccr")

# --- Keep tidy objects for later use/exports
assign("mu_draws",    mu_draws,    inherits = FALSE)
assign("mu_summary",  mu_summary,  inherits = FALSE)
assign("pairwise_summary", pairwise_summary, inherits = FALSE)

```

```{r posterior means, echo=FALSE, fig.align='left', fig.cap=capFig("Posterior draws of the global mean (intercept on the outcome scale) for each model, plotted together in one figure."), fig.height=5, fig.width=6, message=FALSE, warning=FALSE}
# Intercept (population mean μ on outcome scale) — ordered plot

.post_mu_draws <- function(fit, label, ref_row = 1) {
  mu <- fitted(fit, newdata = df_cc[ref_row, , drop = TRUE],
               re_formula = NA, summary = FALSE)
  tibble(model = label, mu = as.numeric(mu))
}

mu_draws <- bind_rows(
  .post_mu_draws(fit_bin_cc,       "Binary guess"),
  .post_mu_draws(fit_ord_gauss_cc, "Judged similarity"),
  .post_mu_draws(fit_cos_beta_cc,  "Cosine similarity")
) %>%
  mutate(model = factor(model,
                        levels = c("Cosine similarity",
                                   "Judged similarity",
                                   "Binary guess")))

mu_summ <- mu_draws %>%
  group_by(model) %>%
  summarise(med = median(mu),
            lo  = quantile(mu, .025),
            hi  = quantile(mu, .975),
            .groups = "drop")

ggplot(mu_draws, aes(x = mu, y = model)) +
  geom_violin(fill = "grey85", color = "grey40", width = 0.9, trim = TRUE) +
  geom_point(data = mu_summ, aes(x = med, y = model), size = 2) +
  geom_segment(data = mu_summ, aes(x = lo, xend = hi, y = model, yend = model),
               linewidth = 0.8) +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,.2)) +
  labs(x = "μ on outcome scale [0,1]", y = NULL)
```

#### Stimuli

```{r eval=TRUE, message=FALSE, warning=FALSE, include=FALSE}
# per-stimulus posterior means + join to cats
nd_stim <- data.frame(stimulus_concept = levels(df_cc$stimulus_concept))

.get_stim <- function(fit, label) {
  f <- fitted(fit, newdata = nd_stim,
              re_formula = ~(1|s|stimulus_concept), summary = TRUE)
  tibble::tibble(
    meaning = nd_stim$stimulus_concept,
    mean = f[, "Estimate"],
    lo   = f[, "Q2.5"],
    hi   = f[, "Q97.5"],
    measure = label
  )
}

stim_means <- bind_rows(
  .get_stim(fit_bin_cc,       "Binary guess"),
  .get_stim(fit_ord_gauss_cc, "Judged similarity"),
  .get_stim(fit_cos_beta_cc,  "Cosine similarity")
) %>%
  left_join(cats, by = "meaning") %>%
  mutate(
    measure = factor(measure, levels = c("Binary guess", "Judged similarity", "Cosine similarity")),
    category = factor(category, levels = c("noun","property","demonstrative","action"))
  )
```

```{r heatmap stimuli, echo=FALSE, fig.align='left', fig.cap=capFig("Posterior means per stimulus on the raw outcome scale (μ). Rows = stimuli, columns = outcome measure. Stimuli grouped by semantic category."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
ggplot(stim_means, aes(x = measure, y = meaning, fill = mean)) +
  geom_tile(color = "white") +
  facet_grid(category ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(limits = c(0,1)) +
  labs(x = NULL, y = NULL, fill = "μ") +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank())
```

```{r heatmap stimuli min-max, echo=FALSE, fig.align='left', fig.cap=capFig("Per-stimulus posterior means, min–max scaled within each outcome measure to emphasize relative patterns across stimuli."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
stim_means_mm <- stim_means %>%
  group_by(measure) %>%
  mutate(mean_mm = (mean - min(mean))/(max(mean)-min(mean))) %>%
  ungroup()

ggplot(stim_means_mm, aes(x = measure, y = meaning, fill = mean_mm)) +
  geom_tile(color = "white") +
  facet_grid(category ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(name = "μ (min–max)") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank())
```

```{r lolipops stimuli, echo=FALSE, fig.align='left', fig.cap=capFig("Posterior means and 95% credible intervals per stimulus, separated by semantic category. Colors = outcome measures."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
ggplot(stim_means, aes(x = mean, y = meaning, color = measure)) +
  geom_point(position = position_dodge(width = 0.4)) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.2,
                 position = position_dodge(width = 0.4)) +
  facet_wrap(~ category, scales = "free_y") +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Posterior mean μ [0–1]", y = NULL)
```

```{r subcategory boxplots stimuli, echo=FALSE, fig.align='left', fig.cap=capFig("Posterior means aggregated at the semantic subcategory level. Each box = distribution of stimulus means within a subcategory; colors = outcome measures."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
ggplot(stim_means, aes(x = subcategory, y = mean, fill = measure)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, position = position_dodge(width = 0.8)) +
  facet_wrap(~ category, scales = "free_x") +
  labs(x = NULL, y = "Posterior mean μ [0–1]") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#### Participant

```{r eval=TRUE, message=FALSE, warning=FALSE, include=FALSE}
nd_part <- data.frame(participant = levels(df_cc$participant))

.get_part <- function(fit, label) {
  f <- fitted(fit, newdata = nd_part,
              re_formula = ~(1|p|participant), summary = TRUE)
  tibble::tibble(
    participant = nd_part$participant,
    mean = f[, "Estimate"],
    measure = label
  )
}

part_means <- bind_rows(
  .get_part(fit_bin_cc,       "Binary guess"),
  .get_part(fit_ord_gauss_cc, "Judged similarity"),
  .get_part(fit_cos_beta_cc,  "Cosine similarity")
) %>%
  mutate(measure = factor(measure, levels = c("Binary guess", "Judged similarity", "Cosine similarity")))
```

```{r density participant, echo=FALSE, fig.align='left', fig.cap=capFig("Distributions of posterior means per participant across the three outcome measures."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
ggplot(part_means, aes(x = mean)) +
  geom_density(fill = "grey85", color = "grey40") +
  facet_wrap(~ measure, nrow = 1, scales = "free_y") +
  labs(x = "Posterior mean per participant (μ)", y = "Density")
```

```{r boxplot participant, echo=FALSE, fig.align='left', fig.cap=capFig("Distribution of posterior means per participant, shown as boxplots for each outcome measure."), fig.height=5, fig.width=6, message=FALSE, warning=FALSE}
ggplot(part_means, aes(x = measure, y = mean, fill = measure)) +
  geom_boxplot(alpha = 0.6) +
  labs(x = NULL, y = "Posterior mean μ [0–1]")
```

### Semantic landscape

```{r}
stopifnot(exists("cats"), exists("df_cc"),
          exists("fit_bin_cc"), exists("fit_ord_gauss_cc"), exists("fit_cos_beta_cc"))

# helper: per-stimulus posterior mean on OUTCOME scale, stimulus RE only
.get_stim <- function(fit, label) {
  nd <- data.frame(stimulus_concept = levels(df_cc$stimulus_concept))
  f  <- fitted(fit, newdata = nd, re_formula = ~(1|s|stimulus_concept), summary = TRUE)
  tibble(
    meaning = nd$stimulus_concept,
    mean    = f[, "Estimate"],
    lo      = f[, "Q2.5"],
    hi      = f[, "Q97.5"],
    measure = label
  )
}

# long → wide with subcategory
stim_means_long <- bind_rows(
  .get_stim(fit_bin_cc,       "Binary guess"),
  .get_stim(fit_ord_gauss_cc, "Judged similarity"),
  .get_stim(fit_cos_beta_cc,  "Cosine similarity")
) %>%
  left_join(cats, by = "meaning")

stim_means_wide <- stim_means_long %>%
  select(meaning, subcategory, measure, mean) %>%
  pivot_wider(names_from = measure, values_from = mean) %>%
  rename(binary = `Binary guess`,
         judged = `Judged similarity`,
         cosine = `Cosine similarity`)

# subcategory centroids (means across stimuli in each subcategory)
subcat_wide <- stim_means_wide %>%
  group_by(subcategory) %>%
  summarise(
    n_stim = dplyr::n(),
    binary = mean(binary, na.rm = TRUE),
    judged = mean(judged, na.rm = TRUE),
    cosine = mean(cosine, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # contrasts (heights) that capture the "gaps" between lenses
    cos_minus_jud = cosine - judged,   # distributional advantage over human-graded
    jud_minus_bin = judged - binary,   # graded advantage over strict hits
    dom = c("Binary","Judged","Cosine")[max.col(cbind(binary, judged, cosine), ties.method = "first")],
    dom = factor(dom, levels = c("Binary","Judged","Cosine"))
  )

assign("stim_means_wide", stim_means_wide, inherits = FALSE)
assign("subcat_wide", subcat_wide, inherits = FALSE)
```

For each subcategory, show two horizontal contrasts on a common zero
line: Judged − Binary (how much graded ratings exceed strict hits) and
Cosine − Judged (how much distributional proximity exceeds human
ratings). This makes the “landscape differences” explicit and
comparable. Subcategories are sorted by Cosine − Judged so the eye
immediately catches where distributional semantics adds the most.

Judged − Binary \> 0 → allowing graded human judgments uncovers
structure ignored by exact hits.

Cosine − Judged \> 0 → distributional proximity recognises
neighborhood-level meaning beyond human “similarity” ratings.

Both \> 0 → a clear gradient: Binary \< Judged \< Cosine for that
subcategory.

Negative values (no such cases) → the earlier lens explains more than
the later one.

```{r map A, echo=FALSE, fig.align='left', fig.cap=capFig("Subcategory Advantage Grid. For each subcategory, the right-facing bar shows Cosine − Judged (distributional advantage) and the left/right bar shows Judged − Binary (graded advantage over exact hits). Bars to the right of zero indicate the second measure exceeds the first. Point size encodes the number of stimuli in the subcategory. Interpretation: where both bars are large and positive, free-text responses are captured poorly by binary hits but progressively better by human grading and best by distributional proximity."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
adv_long <- subcat_wide %>%
  transmute(
    subcategory, n_stim,
    `Judged − Binary` = jud_minus_bin,
    `Cosine − Judged` = cos_minus_jud
  ) %>%
  pivot_longer(cols = c(`Judged − Binary`, `Cosine − Judged`),
               names_to = "contrast", values_to = "value")

# order subcategories by Cosine − Judged (largest to smallest)
ord <- subcat_wide %>% arrange(desc(cos_minus_jud)) %>% pull(subcategory)
adv_long <- adv_long %>% mutate(subcategory = factor(subcategory, levels = ord))

ggplot(adv_long, aes(x = value, y = subcategory, color = contrast)) +
  geom_vline(xintercept = 0, linetype = 3, color = "grey60") +
  geom_point(aes(size = n_stim), alpha = 0.9, position = position_dodge(width = 0.3)) +
  geom_segment(aes(x = 0, xend = value, y = subcategory, yend = subcategory),
               position = position_dodge(width = 0.3), linewidth = 0.9, alpha = 0.7) +
  scale_color_brewer(palette = "Dark2", name = "Contrast") +
  scale_size_continuous(range = c(2.5, 7), name = "Stimuli\nper subcategory") +
  labs(x = "Difference on μ (outcome scale)", y = NULL)
```

```{r map B, echo=FALSE, fig.align='left', fig.cap=capFig("Semantic topography of similarity measures. The plot renders the difference between distributional (cosine) and human-judged similarity as a topographic surface in Judged (x) × Cosine (y) space. Blue “elevations” indicate regions where cosine similarity assigns higher values than human ratings, while red “depressions” show the reverse. The dashed line marks equality (Cos = Jud). Black-rimmed points are subcategory centroids, scaled by their Binary μ (proportion of exact matches). This visualization highlights how different measures carve out distinct semantic landscapes: binary exact hits are rare (small capitals), judged similarity introduces a gradient, and cosine expands the landscape further by capturing semantic neighborhoods beyond human ratings."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
# --- ROBUST "Semantic Topography" (Cos − Jud) -------------------------------
suppressPackageStartupMessages({
  library(dplyr); library(ggplot2)
  has_repel <- requireNamespace("ggrepel", quietly = TRUE)
})

stopifnot(exists("subcat_wide"))

# 1) Clean data for surface
surf_df <- subcat_wide %>%
  transmute(judged = as.numeric(judged),
            cosine = as.numeric(cosine),
            elev   = as.numeric(cos_minus_jud)) %>%
  tidyr::drop_na()

# Safety: need at least 5 points with variation
enough_pts <- nrow(distinct(surf_df, judged, cosine)) >= 5
vary_elev  <- (is.finite(var(surf_df$elev)) && var(surf_df$elev) > 1e-8)

# Axis ranges with padding, even if range is tiny
rx <- range(surf_df$judged, na.rm = TRUE); dx <- diff(rx); if (!is.finite(dx) || dx <= 1e-6) { rx <- c(rx[1]-0.05, rx[1]+0.05) }
ry <- range(surf_df$cosine, na.rm = TRUE); dy <- diff(ry); if (!is.finite(dy) || dy <= 1e-6) { ry <- c(ry[1]-0.05, ry[1]+0.05) }
pad_x <- max(0.05, 0.08 * diff(rx)); pad_y <- max(0.05, 0.08 * diff(ry))

# 2) Build grid and elevation by LOESS (fallback path)
make_surface <- function() {
  grid_x <- seq(rx[1] - pad_x, rx[2] + pad_x, length.out = 160)
  grid_y <- seq(ry[1] - pad_y, ry[2] + pad_y, length.out = 160)
  grid   <- expand.grid(judged = grid_x, cosine = grid_y)

  if (enough_pts && vary_elev) {
    fit <- loess(elev ~ judged * cosine, data = surf_df, span = 0.75, degree = 2,
                 control = loess.control(surface = "interpolate", statistics = "approximate"))
    grid$elev <- as.numeric(predict(fit, newdata = grid))
  } else {
    # Degenerate: constant (or nearly) — just use the mean as a flat surface
    grid$elev <- mean(surf_df$elev, na.rm = TRUE)
  }

  grid %>% tidyr::drop_na()
}

elev_df <- make_surface()

# 3) Color scale that always shows something: symmetric around 0
rng_elev <- range(elev_df$elev[is.finite(elev_df$elev)], na.rm = TRUE)
M <- max(abs(rng_elev), 0.05)  # ensure non-zero span
fill_scale <- scale_fill_gradient2(
  low = "#B2182B", mid = "white", high = "#2166AC",
  midpoint = 0, limits = c(-M, M), name = "Cos − Jud"
)

# 4) Build the plot
base <- ggplot() +
  geom_raster(data = elev_df, aes(x = judged, y = cosine, fill = elev), alpha = 0.95) +
  fill_scale +
  geom_abline(slope = 1, intercept = 0, linetype = 3, color = "grey30") +
  coord_equal(
    xlim = c(rx[1] - pad_x, rx[2] + pad_x),
    ylim = c(ry[1] - pad_y, ry[2] + pad_y),
    expand = FALSE
  ) +
  labs(
    x = "Judged similarity (μ)",
    y = "Cosine similarity (μ)"
  ) +
  theme(panel.grid = element_blank())

# 5) Add 0-ridge & contour bands only if there’s meaningful variation
has_both_signs <- (min(elev_df$elev, na.rm = TRUE) < 0) && (max(elev_df$elev, na.rm = TRUE) > 0)
if (has_both_signs) {
  base <- base +
    geom_contour(data = elev_df, aes(x = judged, y = cosine, z = elev),
                 breaks = 0, color = "grey15", linewidth = 0.5, alpha = 0.8) +
    geom_contour(data = elev_df, aes(x = judged, y = cosine, z = elev),
                 bins = 8, color = "white", linewidth = 0.25, alpha = 0.35)
}

# 6) Capitals (subcategory centroids), size = Binary; labels
capitals <- subcat_wide %>%
  transmute(subcategory, judged, cosine, binary)

p_topo <- base +
  geom_point(data = capitals,
             aes(x = judged, y = cosine, size = binary),
             shape = 21, fill = "white", color = "black", stroke = 0.5, alpha = 0.96) +
  scale_size_continuous(range = c(3, 8), name = "Binary (μ)")

if (has_repel) {
  p_topo <- p_topo +
    ggrepel::geom_text_repel(
      data = capitals, aes(x = judged, y = cosine, label = subcategory),
      size = 3.1, show.legend = FALSE, max.overlaps = 100
    )
} else {
  p_topo <- p_topo +
    geom_text(
      data = capitals, aes(x = judged, y = cosine, label = subcategory),
      size = 3.1, vjust = -0.8, show.legend = FALSE
    )
}

print(p_topo)
```

```{r map C, echo=FALSE, fig.align='left', fig.cap=capFig("Each subcategory’s centroid in Judged (x) × Cosine (y) becomes a “capital.” We tessellate the plane into Voronoi regions, coloring each “country” by Cosine − Judged (blue = distributional advantage; red = judged advantage). Point size shows Binary μ (strict hits), and labels name the subcategories. The dashed 45° line marks equality. This gives a true map where borders reflect which centroid is closest; visually, you get semantic “territories” defined by graded measures, with color/topography conveying which lens is more permissive."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
# --- MAP C: Voronoi "countries" in Judged × Cosine space --------------------
suppressPackageStartupMessages({
  library(ggplot2); library(dplyr)
  has_force <- requireNamespace("ggforce", quietly = TRUE)
  has_repel <- requireNamespace("ggrepel", quietly = TRUE)
})

if (has_force) {
  library(ggforce)

  # Compute a loose bounding box for tiles
  rng_x <- range(subcat_wide$judged, na.rm = TRUE)
  rng_y <- range(subcat_wide$cosine, na.rm = TRUE)
  pad_x <- diff(rng_x) * 0.08
  pad_y <- diff(rng_y) * 0.08
  xmin <- rng_x[1] - pad_x; xmax <- rng_x[2] + pad_x
  ymin <- rng_y[1] - pad_y; ymax <- rng_y[2] + pad_y

  p_voronoi <- ggplot(subcat_wide, aes(x = judged, y = cosine)) +
    # Voronoi tiles colored by Cos−Jud
    ggforce::geom_voronoi_tile(
      aes(fill = cos_minus_jud),
      bound = c(xmin, ymin, xmax, ymax),
      alpha = 0.8, color = "white", linewidth = 0.4
    ) +
    # equality reference
    geom_abline(slope = 1, intercept = 0, linetype = 3, color = "grey60") +
    # Capitals: subcategory centroids
    geom_point(aes(size = binary), shape = 21, fill = "white", color = "black", stroke = 0.5) +
    { if (has_repel)
        ggrepel::geom_text_repel(aes(label = subcategory), size = 3.1,
                                 max.overlaps = 100, seed = 1708, box.padding = 0.3)
      else
        geom_text(aes(label = subcategory), size = 3.1, vjust = -0.8)
    } +
    scale_fill_distiller(
      palette = "RdBu", direction = -1,
      name = "Cos − Jud",
      guide = guide_colorbar(title.position = "top")
    ) +
    scale_size_continuous(range = c(3, 8), name = "Binary (μ)") +
    coord_equal(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
    labs(
      x = "Judged similarity (μ)",
      y = "Cosine similarity (μ)"
    ) +
    theme(panel.grid = element_blank())

  print(p_voronoi)

} else {
  message("Package 'ggforce' not installed. Install with install.packages('ggforce') to see the Voronoi map.")
  # Fallback: simple scatter with colored background tiles omitted
  ggplot(subcat_wide, aes(x = judged, y = cosine)) +
    geom_abline(slope = 1, intercept = 0, linetype = 3, color = "grey60") +
    geom_point(aes(color = cos_minus_jud, size = binary)) +
    { if (has_repel) ggrepel::geom_text_repel(aes(label = subcategory), size = 3.1)
      else geom_text(aes(label = subcategory), size = 3.1, vjust = -0.8) } +
    scale_color_distiller(palette = "RdBu", direction = -1, name = "Cos − Jud") +
    scale_size_continuous(range = c(3,8), name = "Binary (μ)") +
    coord_equal() +
    labs(x = "Judged similarity (μ)", y = "Cosine similarity (μ)",
         title = "Subcategory map (fallback): points colored by Cos−Jud; size = Binary")
}

```

# Session info

```{r echo=TRUE, message=FALSE, warning=FALSE}
sessionInfo()
```
