---
title: "Experiments in language evolution: inferences from closed- vs open-ended semantic space paradigms"
subtitle: "Data modeling"
author: "analysis by Aleksandra Ćwiek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_document:
    theme: readable
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '3'
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

**Experiments in language evolution: inferences from closed- vs open-ended semantic space paradigms**

Svetlana Kuleshova\*1,2, Stefan Hartmann3, and Michael Pleyer1, Marta
Sibierska1, Johan Blomberg4, Przemysław Żywiczyński1, Aleksandra
Ćwiek5,1, Sławomir Wacewicz1

\*Corresponding Author: 40010189\@parisnanterre.fr

1Nicolaus Copernicus University in Toruń, Center for Language Evolution
Studies, Fosa Staromiejska 3, 87-100 Toruń, Poland.

2ArScAn-Équipe AnTET (UMR 7041), CNRS, Université Paris Nanterre, 21
Allée de l'Université, 92000, Nanterre, France.

3Heinrich Heine University Düsseldorf, Faculty of Arts and Humanities,
Universitätsstraße 1, 40225 Düsseldorf, Germany.

4Stockholm University, Department of Swedish and Multilingualism.
Universitetsvägen 10, 106 91 Stockholm.

5Leibniz-Centre General Linguistics, Berlin, 10719, Germany

We challenge the validity of the multiple-choice paradigm used in many
communication experiments. We replicated two experiments and replaced
the original multiple-choice design with free-text answers. Our results
indicate that participants performed worse compared to the original
studies. The post hoc analysis showed that often relevant semantic
domains were correctly identified, but full congruency with the target
concepts was rare. We discuss how the results of each paradigm can be
mapped onto a larger picture of language evolution.

What inferences about reality can we make from linguistic experiments?
By definition, controlled experiments do not represent reality in all
its complexity, and thus, it is important to be cautious when drawing
conclusions. As one example, some recent experimental studies have made
strong claims concluding that humans can understand improvised or
interspecies forms of communication, specifically, novel non-linguistic
vocalizations (Ćwiek et al., 2021) and gestures in non-human great apes
(Graham and Hobaiter, 2023). Here, we follow up on these studies,
altering one crucial aspect: Both original studies used a
multiple-choice paradigm, whereby participants were constrained to
choose one answer from a small set of alternatives; that is, for each
gesture or vocalisation they were offered a set of potential
interpretations and asked to guess the one best reflecting its intended
meaning. We expected that replacing the closed-ended, multiple-choice
paradigm with an open-ended, free-text paradigm would lead to different
results and thus different conclusions about human capacities to
‘understand’ ape gestures or novel vocalizations.

To test our hypothesis, we replicated the studies by Ćwiek et al. and
Graham & Hobaiter, but with the participants freely typing in their
answer (i.e., their interpretation of the meaning of an ape gesture or
novel vocalisation) rather than selecting it from a set of predefined
choices. To assess the proximity of the participants’ answers with the
target concepts, we used three complementary approaches: (i) binary
assessment of correctness (yes/no), (ii) manual coding on a four-point
scale, (iii) cosine similarity between the participants’ responses and
the target concepts using ConceptNet. For (i) and (ii), participants’
answers were coded by two raters on a four-point Likert scale
(transformed into a binary system for (i)), showing moderate-to-high
inter-rater agreement. All mismatches between coders were discussed in
the team and resolved. For (iii), the proximity between the words in the
responses (heads of phrases in the case of multi-word answers) and the
word denoting the target concept in ConceptNet (Speer et al. 2017) were
assessed automatically. Across all approaches, our results show that
participants performed qualitatively worse compared to the original
studies, with a vast majority of responses showing no resemblance
whatsoever to the target concept according to the manual coding
procedure. In the ConceptNet analysis, the responses were found to be
conceptually fairly close to the target concepts in many cases (e.g.,
the same semantic domain, as in ‘sleep’ vs ‘snore’), but hardly any
response was fully congruent with the target concept.

Apart from revisiting the question to what extent participants
“understand” novel vocalizations or ape gestures, our data give us the
opportunity to evaluate different possibilities of interpreting
participants’ free-text answers: manual coding with different levels of
granularity on the one hand, and automatic similarity detection using
NLP resources on the other. The three measures we used carve out
different semantic landscapes, shaping how we perceive communicative
transparency. A perfect-match approach treats meaning as a binary on/off
switch, limiting “understanding” to exact hits while discarding
near-misses as failures. The Likert-scale approach expands this space,
recognizing perceived shades of similarity. The ConceptNet-based cosine
similarity approach stretches the landscape even further, mapping words
onto a gradient of conceptual distances, where even broad associations
register as meaningful connections. Because these methods define
similarity differently, they potentially reveal different facets of how
novel signals convey meaning.

We thus conclude that different setups and analysis methods give us a
nuanced perspective on the role of semantic space in semiotic
experiments. These various paths to data collection and data treatment
can inform a nuanced discussion of how experimental results can inform
the evolution of today’s languages.

# Data preparation

## Source setup

```{r source setup, echo = TRUE, message=FALSE, warning = FALSE}
########## folders ##########
library(here)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

parentfolder <- dirname(getwd())

data          <- paste0(parentfolder, '/data/')
models        <- paste0(parentfolder, '/models/')
plots         <- paste0(parentfolder, '/plots/')
scripts       <- paste0(parentfolder, '/scripts/')

########## source file ##########

#source(paste0(scripts, "adjectives-preparation.R"))

#################### packages ####################
# Core tidyverse for wrangling + plotting; janitor/skimr for quick structure checks; patchwork for combining plots
library(tidyverse)
library(janitor)
library(skimr)
library(knitr)
library(cowplot)
library(ggdist)
library(readxl)
library(lubridate)
library(patchwork)
library(hexbin)
library(PRROC)

# Bayesian
library(brms);
library(loo);packageVersion("loo")
library(HDInterval);packageVersion("HDInterval")
library(cmdstanr);
library(emmeans);packageVersion("emmeans")
library(posterior);
library(bayesplot);
library(BayesFactor);
library(tidybayes);packageVersion("tidybayes")
library(bayestestR);

# use all available cores for parallel computing
options(mc.cores = parallel::detectCores())

# Keep plots/readability consistent
theme_set(theme_minimal(base_size = 12) + panel_border())

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r echo = TRUE, message=FALSE, warning = FALSE}
# --- figure & table captions ---
outputFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to"); if (is.null(outputFormat)) outputFormat <- ""
capTabNo <- 1L; capFigNo <- 1L
.last_fig_no <- NA_integer_

capTab <- function(x){
  if (outputFormat == 'html'){
    x <- paste0("**Table ", capTabNo, ".** ", x)
    capTabNo <<- capTabNo + 1L
  }
  x
}

capFig <- function(x, show_R_version=FALSE, show_package_versions=NULL, is_map=FALSE){
  if (outputFormat == 'html'){
    assign(".last_fig_no", capFigNo, envir = .GlobalEnv) # remember current number
    x <- paste0("**Figure ", capFigNo, ".** ", x)
    capFigNo <<- capFigNo + 1L
  }
  x
}

# --- plot hook to auto-save ---
.default_plot_hook <- knitr::knit_hooks$get("plot")
.has_magick <- requireNamespace("magick", quietly = TRUE)

knitr::knit_hooks$set(plot = function(x, options){
  res <- if (is.function(.default_plot_hook)) .default_plot_hook(x, options) else NULL

  if (!is.null(options$fig.cap)) {
    fig_no <- get(".last_fig_no", envir = .GlobalEnv, inherits = FALSE)
    if (!is.na(fig_no)) {
      fig_no_str <- sprintf("%02d", fig_no)

      # subfolder = Rmd filename without extension
      rmd_name <- tools::file_path_sans_ext(basename(knitr::current_input()))
      target_dir <- file.path(plots, rmd_name)
      if (!dir.exists(target_dir)) dir.create(target_dir, recursive = TRUE)

      png_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".png"))
      tif_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".tif"))

      file.copy(x, png_out, overwrite = TRUE)

      if (.has_magick) {
        img <- magick::image_read(png_out)
        magick::image_write(img, path = tif_out, format = "tiff", compression = "lzw")
      } else {
        file.copy(png_out, tif_out, overwrite = TRUE)
      }
    }
  }

  res
})

```

Other helpers

```{r}
bb_corr <- function(x, y, R = 4000, seed = 1708) {
  stopifnot(length(x) == length(y))
  set.seed(seed)
  n <- length(x)
  out <- numeric(R)
  for (r in seq_len(R)) {
    w <- rexp(n)          # Dirichlet via i.i.d. Exponential then normalize
    w <- w / sum(w)
    mx <- sum(w * x); my <- sum(w * y)
    sx <- sqrt(sum(w * (x - mx)^2))
    sy <- sqrt(sum(w * (y - my)^2))
    out[r] <- sum(w * (x - mx) * (y - my)) / (sx * sy)
  }
  out
}

summ_post <- function(draws, name){
  h <- hdi(draws, ci = 0.95)
  tibble(pair = name,
         mean = mean(draws),
         median = median(draws),
         hdi_low = h$CI_low,
         hdi_high = h$CI_high)
}

bf_bic <- function(y, x){
  fit1 <- lm(y ~ x)
  fit0 <- lm(y ~ 1)
  BIC0 <- BIC(fit0); BIC1 <- BIC(fit1)
  # BF10 ≈ exp((BIC0 - BIC1)/2)
  exp((BIC0 - BIC1)/2)
}
```


## Load in data frames

```{r echo=TRUE, message=FALSE, warning=FALSE}
df <- read_csv(paste0(data, "final_df.csv"))  %>% clean_names()
```

# Correlations

```{r}
dat_cor <- df %>%
  select(cosine = cosine_rescaled, judged = ord_norm, binary = bin_f) %>%
  drop_na()
```

Pearson correlations (with CIs)

```{r}
pearson_mat <- cor(dat_cor, use = "pairwise.complete.obs", method = "pearson")
pearson_mat
```

Pairwise Pearson tests (p-values + CI)

```{r}
pearson_tests <- list(
  cosine_judged = cor.test(dat_cor$cosine, dat_cor$judged),
  cosine_binary = cor.test(dat_cor$cosine, dat_cor$binary),
  judged_binary = cor.test(dat_cor$judged, dat_cor$binary)
)
pearson_tests

pearson_tbl <- tibble(
  pair = c("cosine ~ judged", "cosine ~ binary", "judged ~ binary"),
  r = c(cor(dat_cor$cosine, dat_cor$judged),
        cor(dat_cor$cosine, dat_cor$binary),
        cor(dat_cor$judged, dat_cor$binary))
)
```

Bayesian correlations via BayesFactor
(Point-biserial is just Pearson on numeric 0/1; we still get Bayesian posterior for rho.)

```{r}
post_cos_jud <- bb_corr(dat_cor$cosine, dat_cor$judged, R = 8000, seed = 1708)
post_cos_bin <- bb_corr(dat_cor$cosine, dat_cor$binary, R = 8000, seed = 1708)
post_jud_bin <- bb_corr(dat_cor$judged, dat_cor$binary, R = 8000, seed = 1708)

bayes_corr_summary <- bind_rows(
  summ_post(post_cos_jud, "cosine ~ judged"),
  summ_post(post_cos_bin, "cosine ~ binary"),
  summ_post(post_jud_bin, "judged ~ binary")
)
bayes_corr_summary

# Posterior density plots
bind_rows(
  data.frame(rho = post_cos_jud, pair = "cosine~judged"),
  data.frame(rho = post_cos_bin, pair = "cosine~binary"),
  data.frame(rho = post_jud_bin, pair = "judged~binary")
) %>%
  ggplot(aes(rho)) + geom_density() + facet_wrap(~pair, scales = "free") +
  geom_vline(xintercept = 0, linetype = 2)
```

BIC-approx Bayes factors for "correlation != 0"
For continuous-continuous: linear model; for (cont ~ binary): also linear is fine here.

```{r}
BF_cos_jud <- bf_bic(dat_cor$cosine, dat_cor$judged)
BF_cos_bin <- bf_bic(dat_cor$cosine, dat_cor$binary)
BF_jud_bin <- bf_bic(dat_cor$judged, dat_cor$binary)

bf_tbl <- tibble(
  pair = c("cosine ~ judged", "cosine ~ binary", "judged ~ binary"),
  BF10 = c(BF_cos_jud, BF_cos_bin, BF_jud_bin)
)
```

One clean summary table

```{r}
pearson_tbl %>%
  left_join(bayes_corr_summary, by = "pair") %>%
  left_join(bf_tbl, by = "pair")
```

# Modeling

We first fit baseline models for each of the three measures, with only random intercepts for participant, native language, and stimulus. After baselines we’ll add cross-measure “coupling” models to test how measures relate (but aren’t identical), compare by LOO (and optionally Bayes factors via bridgesampling), and extract posteriors that demonstrate distinct yet overlapping semantic landscapes.

Full solution. (It will take days.)

```{r}
{
  set.seed(1708)

  ## ---- ensure models dir exists (you already have `models` as an absolute path) ----
  if (!dir.exists(models)) dir.create(models, recursive = TRUE, showWarnings = FALSE)

  ## ---- data prep (minimal; no z-scoring of cosine) ----
  if (!"ord_norm" %in% names(df)) df$ord_norm <- df$judged_similarity / 3
  df_brm <- df
  df_brm$participant      <- factor(df_brm$participant)
  df_brm$stimulus_concept <- factor(df_brm$stimulus_concept)
  df_brm$native_language  <- factor(df_brm$native_language) # decided to do without it because it was not variable and caused much longer computation times
  df_brm$judged_ord       <- factor(df_brm$judged_similarity, levels = 0:3, ordered = TRUE)
 # Replace values extremely close to 1 with exactly 1
  df_brm$cosine_rescaled <- ifelse(
    !is.na(df_brm$cosine_rescaled) & df_brm$cosine_rescaled > 1 & df_brm$cosine_rescaled <
      1.001,
    1,
    df_brm$cosine_rescaled
  )
  
  ## ---- very wide priors for baselines / couplings (not for ZOIB components) ----
  .pri_bin <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )
  .pri_ord <- c(
    set_prior("normal(0, 5)", class = "Intercept"),  # thresholds
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )
  .pri_wide <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )

  ## ========================= BASELINES =========================
  .bin_rds <- file.path(models, "fit_bin.rds")
  if (file.exists(.bin_rds)) {
    fit_bin <- readRDS(.bin_rds)
  } else {
    fit_bin <- brm(
      binary_guess ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = bernoulli(),
      prior   = .pri_bin,
      chains  = 4, iter = 8000, warmup = 4000,
      control = list(adapt_delta = 0.97, max_treedepth = 12),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_bin, .bin_rds)
  }

  .ord_rds <- file.path(models, "fit_ord.rds")
  if (file.exists(.ord_rds)) {
    fit_ord <- readRDS(.ord_rds)
  } else {
    fit_ord <- brm(
      judged_ord ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = cumulative(link = "logit"),
      prior   = .pri_ord,
      chains  = 4, iter = 8000, warmup = 4000,
      control = list(adapt_delta = 0.97, max_treedepth = 12),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_ord, .ord_rds)
  }

  .cos_rds <- file.path(models, "fit_cos.rds")
  if (file.exists(.cos_rds)) {
    fit_cos <- readRDS(.cos_rds)
  } else {
    fit_cos <- brm(
      cosine_rescaled ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = zero_one_inflated_beta(),  # includes exact 0s and 1s
      # per your rule: no explicit priors for ZOIB subcomponents
      chains  = 4, iter = 8000, warmup = 4000,
      control = list(adapt_delta = 0.97, max_treedepth = 12),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_cos, .cos_rds)
  }

  ## ---- compact Bayes R^2 (kept as r2_table) ----
  .r2 <- function(fit) { r <- bayes_R2(fit); c(Estimate = mean(r), Q2.5 = quantile(r,.025), Q97.5 = quantile(r,.975)) }
  r2_table <- data.frame(
    model = c("Binary (Bernoulli)", "Judged (Ordinal)", "Cosine (ZOIB)"),
    t(rbind(.r2(fit_bin), .r2(fit_ord), .r2(fit_cos)))
  )
  rownames(r2_table) <- NULL
  r2_table$Estimate <- round(r2_table$Estimate, 3)
  r2_table$Q2.5     <- round(r2_table$Q2.5, 3)
  r2_table$Q97.5    <- round(r2_table$Q97.5, 3)
  print(r2_table)

  ## ========================= COUPLING MODELS =========================
  ## Judged <- Cosine (no z-scoring; cosine_rescaled is 0–1 and interpretable)
  .m_ord_cos_rds <- file.path(models, "m_ord_cos.rds")
  if (file.exists(.m_ord_cos_rds)) {
    m_ord_cos <- readRDS(.m_ord_cos_rds)
  } else {
    m_ord_cos <- brm(
      judged_ord ~ 1 + cosine_rescaled +
        (1 + cosine_rescaled | s | stimulus_concept) +
        (1 | p | participant) + (1 | l | native_language),
      data    = df_brm, family = cumulative("logit"),
      prior   = .pri_wide,
      chains  = 4, iter = 8000, warmup = 4000,
      control = list(adapt_delta = 0.98, max_treedepth = 13),
      seed    = 1708, refresh = 50,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(m_ord_cos, .m_ord_cos_rds)
  }

  ## Binary <- Cosine
  .m_bin_cos_rds <- file.path(models, "m_bin_cos.rds")
  if (file.exists(.m_bin_cos_rds)) {
    m_bin_cos <- readRDS(.m_bin_cos_rds)
  } else {
    m_bin_cos <- brm(
      binary_guess ~ 1 + cosine_rescaled +
        (1 + cosine_rescaled | s | stimulus_concept) +
        (1 | p | participant) + (1 | l | native_language),
      data    = df_brm, family = bernoulli(),
      prior   = .pri_wide,
      chains  = 4, iter = 8000, warmup = 4000,
      control = list(adapt_delta = 0.98, max_treedepth = 13),
      seed    = 1708, refresh = 50,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(m_bin_cos, .m_bin_cos_rds)
  }

  ## Binary <- Judged (uses your 0–1 'ord_norm')
  .m_bin_jud_rds <- file.path(models, "m_bin_jud.rds")
  if (file.exists(.m_bin_jud_rds)) {
    m_bin_jud <- readRDS(.m_bin_jud_rds)
  } else {
    m_bin_jud <- brm(
      binary_guess ~ 1 + ord_norm +
        (1 + ord_norm | s | stimulus_concept) +
        (1 | p | participant) + (1 | l | native_language),
      data    = df_brm, family = bernoulli(),
      prior   = .pri_wide,
      chains  = 4, iter = 8000, warmup = 4000,
      control = list(adapt_delta = 0.98, max_treedepth = 13),
      seed    = 1708, refresh = 50,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(m_bin_jud, .m_bin_jud_rds)
  }

  ## Cosine (ZOIB, μ) <- Judged (keep zoi/oi intercept-only)
  .m_cos_jud_rds <- file.path(models, "m_cos_jud.rds")
  if (file.exists(.m_cos_jud_rds)) {
    m_cos_jud <- readRDS(.m_cos_jud_rds)
  } else {
    m_cos_jud <- brm(
      cosine_rescaled ~ 1 + ord_norm +
        (1 + ord_norm | s | stimulus_concept) +
        (1 | p | participant) + (1 | l | native_language),
      data    = df_brm, family = zero_one_inflated_beta(),
      # no explicit priors for ZOIB components per your instruction
      chains  = 4, iter = 8000, warmup = 4000,
      control = list(adapt_delta = 0.98, max_treedepth = 13),
      seed    = 1708, refresh = 50,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(m_cos_jud, .m_cos_jud_rds)
  }

  ## ---- LOO comparisons (vs respective baselines) ----
  cat("\n# LOO: Binary baseline vs Binary <- Cosine\n")
  print(loo_compare(loo(fit_bin), loo(m_bin_cos)))
  cat("\n# LOO: Binary baseline vs Binary <- Judged\n")
  print(loo_compare(loo(fit_bin), loo(m_bin_jud)))
  cat("\n# LOO: Ordinal baseline vs Judged <- Cosine\n")
  print(loo_compare(loo(fit_ord), loo(m_ord_cos)))
  cat("\n# LOO: Cosine baseline vs Cosine <- Judged (ZOIB)\n")
  print(loo_compare(loo(fit_cos), loo(m_cos_jud)))

  ## ---- cleanup helpers (keep only model objects + r2_table + path vars) ----
  rm(list = c(".pri_bin",".pri_ord",".pri_wide",
              ".bin_rds",".ord_rds",".cos_rds",
              ".m_ord_cos_rds",".m_bin_cos_rds",".m_bin_jud_rds",".m_cos_jud_rds"))
}
```

Compromise to meet computational time.

    -     We keep the 1’s from cosine similarity conceptually (they’re just ε below 1 for Beta math), preserving order and near-perfect matches. Also rescaling to true 0-1 range (not positively skewed with much higher mean due to the nature of cosine similarity scale).

    -     All three outcomes are on comparable 0–1 scales (binary 0/1, judged 0–1, cosine ε-open (0,1)), so the per-stimulus posterior means can be min–max rescaled to overlay the three “landscapes” directly (same axis).
      
    -     Fast & robust: Bernoulli + Gaussian + Beta are far lighter than cumulative + ZOIB at N=27k, but they still answer the main question: the measures are related but not identical, visible in the differing shapes/ranks of stim_landscapes_scaled.
    
```{r}
{
  set.seed(1708)

  ## ---------- ensure models dir exists ----------
  if (!dir.exists(models)) dir.create(models, recursive = TRUE, showWarnings = FALSE)

  ## ---------- data prep ----------
  if (!"ord_norm" %in% names(df)) df$ord_norm <- df$judged_similarity / 3
  df_brm <- df
  df_brm$participant      <- factor(df_brm$participant)
  df_brm$stimulus_concept <- factor(df_brm$stimulus_concept)

  ## Keep info about exact 1s (for reporting the spike)
  ones_table_stim <- aggregate(I(df_brm$cosine_rescaled == 1) ~ stimulus_concept, data = df_brm, FUN = mean)
  names(ones_table_stim) <- c("stimulus_concept", "one_rate")

  ## Cosine for modeling:
  ## 1) tiny nudge to open interval (preserve order)
  eps  <- 1e-10
  tiny <- 1e-12
  x    <- df_brm$cosine_rescaled
  x_open <- pmin(pmax(x - eps, tiny), 1 - tiny)

  ## 2) min–max stretch to [0,1] (use observed min/max), then open again
  xmin <- min(x_open, na.rm = TRUE)
  xmax <- max(x_open, na.rm = TRUE)
  span <- xmax - xmin
  if (!is.finite(span) || span <= 0) stop("Cosine has zero span after opening; cannot min–max scale.")
  x_unit <- (x_open - xmin) / span
  df_brm$cosine_unit_open <- pmin(pmax(x_unit, tiny), 1 - tiny)

  ## ---------- very wide priors (intercepts + RE sd only) ----------
  .pri_intercepts_only <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )

  ## =================================================================
  ##                           BASELINES
  ## =================================================================

  ## 1) BINARY (Bernoulli) — reuse if already run
  .bin_rds <- file.path(models, "fit_bin.rds")
  if (file.exists(.bin_rds)) {
    fit_bin <- readRDS(.bin_rds)
  } else {
    fit_bin <- brm(
      binary_guess ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = bernoulli(),
      prior   = .pri_intercepts_only,
      chains  = 4, iter = 4000, warmup = 2000,
      control = list(adapt_delta = 0.97, max_treedepth = 12),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_bin, .bin_rds)
  }

  ## 2) JUDGED (fast): treat 0..3 as continuous 0–1 (Gaussian)
  .ord_rds <- file.path(models, "fit_ord_gauss.rds")
  if (file.exists(.ord_rds)) {
    fit_ord_gauss <- readRDS(.ord_rds)
  } else {
    fit_ord_gauss <- brm(
      ord_norm ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = gaussian(),
      prior   = .pri_intercepts_only,
      chains  = 4, iter = 4000, warmup = 2000,
      control = list(adapt_delta = 0.95, max_treedepth = 12),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_ord_gauss, .ord_rds)
  }

  ## 3) COSINE (fast): Beta mixed model on unit-stretched (0,1) cosine
  .cos_rds <- file.path(models, "fit_cos_beta_unit.rds")
  if (file.exists(.cos_rds)) {
    fit_cos_beta <- readRDS(.cos_rds)
  } else {
    fit_cos_beta <- brm(
      cosine_unit_open ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = Beta(),
      prior   = .pri_intercepts_only,
      chains  = 4, iter = 4000, warmup = 2000,
      control = list(adapt_delta = 0.95, max_treedepth = 12),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_cos_beta, .cos_rds)
  }

  ## ---------- Compact Bayes R^2 ----------
  .r2 <- function(fit) { r <- bayes_R2(fit); c(Estimate = mean(r), Q2.5 = quantile(r,.025), Q97.5 = quantile(r,.975)) }
  r2_table_fast <- data.frame(
    model = c("Binary (Bernoulli)", "Judged (Gaussian 0–1)", "Cosine (Beta, unit-stretched)"),
    t(rbind(.r2(fit_bin), .r2(fit_ord_gauss), .r2(fit_cos_beta)))
  )
  rownames(r2_table_fast) <- NULL
  r2_table_fast$Estimate <- round(r2_table_fast$Estimate, 3)
  r2_table_fast$Q2.5     <- round(r2_table_fast$Q2.5, 3)
  r2_table_fast$Q97.5    <- round(r2_table_fast$Q97.5, 3)
  print(r2_table_fast)

  ## =================================================================
  ##  Per-stimulus posterior means + min–max for visual comparability
  ## =================================================================
  stim_levels <- levels(df_brm$stimulus_concept)

  # Binary: posterior mean per stimulus
  bin_stim <- tibble::tibble(stimulus_concept = stim_levels) |>
    dplyr::bind_cols(as.data.frame(fitted(fit_bin, re_formula = ~(1|s|stimulus_concept), summary = TRUE))) |>
    dplyr::transmute(stimulus_concept, bin_mean = Estimate)

  # Judged (Gaussian): posterior mean per stimulus
  ord_stim <- tibble::tibble(stimulus_concept = stim_levels) |>
    dplyr::bind_cols(as.data.frame(fitted(fit_ord_gauss, re_formula = ~(1|s|stimulus_concept), summary = TRUE))) |>
    dplyr::transmute(stimulus_concept, ord_mean = Estimate)

  # Cosine (Beta): posterior mean per stimulus
  cos_stim <- tibble::tibble(stimulus_concept = stim_levels) |>
    dplyr::bind_cols(as.data.frame(fitted(fit_cos_beta, re_formula = ~(1|s|stimulus_concept), summary = TRUE))) |>
    dplyr::transmute(stimulus_concept, cos_mean = Estimate)

  stim_landscapes <- bin_stim |>
    dplyr::left_join(ord_stim, by = "stimulus_concept") |>
    dplyr::left_join(cos_stim, by = "stimulus_concept") |>
    dplyr::left_join(ones_table_stim, by = "stimulus_concept")  # keep spike info

  # Min–max rescale summaries for plotting on same axis (does not affect models)
  mm <- function(v) (v - min(v, na.rm = TRUE)) / (max(v, na.rm = TRUE) - min(v, na.rm = TRUE))
  stim_landscapes_scaled <- stim_landscapes |>
    dplyr::mutate(
      bin_mm = mm(bin_mean),
      ord_mm = mm(ord_mean),
      cos_mm = mm(cos_mean)
    )

  # keep the useful things
  assign("fit_bin", fit_bin, inherits = FALSE)
  assign("fit_ord_gauss", fit_ord_gauss, inherits = FALSE)
  assign("fit_cos_beta", fit_cos_beta, inherits = FALSE)
  assign("r2_table_fast", r2_table_fast, inherits = FALSE)
  assign("stim_landscapes", stim_landscapes, inherits = FALSE)
  assign("stim_landscapes_scaled", stim_landscapes_scaled, inherits = FALSE)

  ## ---------- cleanup helpers ----------
  rm(.pri_intercepts_only, .bin_rds, .ord_rds, .cos_rds, .r2,
     ones_table_stim, x, x_open, x_unit, eps, tiny, xmin, xmax, span, stim_levels, mm)
}
```


# Session info

```{r echo=TRUE, message=FALSE, warning=FALSE}
sessionInfo()
```
