---
title: "Experiments in language evolution: inferences from closed- vs open-ended semantic space paradigms"
subtitle: "Data modeling"
author: "analysis by Aleksandra Ćwiek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_document:
    theme: readable
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '3'
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What this document does

**Purpose.** This notebook fits and evaluates **baseline hierarchical
models** for three outcome measures that capture how well participants’
free-text responses match target concepts:

1.  `binary_guess` — whether the response was judged exact/near-exact
    (yes/no),

2.  `judged_similarity` — coder-assigned similarity on a four-point
    scale,

3.  `cosine_rescaled` — automatic cosine similarity (ConceptNet,
    rescaled to [0,1]).

Each outcome is analyzed separately. We summarize posterior
distributions, report Bayesian R², and check convergence/diagnostics.

**Inputs.** One cleaned, analysis-ready dataset from `1_preparation.Rmd`
(`data/final_df.csv`), containing participant IDs, stimulus/concept IDs,
language, and the three outcome measures.

**Outputs.**

-   Cached fitted models in `models/` (RDS files).

-   Summary tables (posterior means/HDIs, random-effect SDs, Bayesian
    R²).

-   Diagnostic plots and posterior predictive checks.

-   Figures saved to `plots/2_modeling/` (auto-saved with stable
    numbering).

# Data preparation

## Source setup

```{r source setup, echo = TRUE, message=FALSE, warning = FALSE}
########## folders ##########
library(here)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

parentfolder <- dirname(getwd())

data          <- paste0(parentfolder, '/data/')
models        <- paste0(parentfolder, '/models/')
plots         <- paste0(parentfolder, '/plots/')
scripts       <- paste0(parentfolder, '/scripts/')

########## source file ##########

#source(paste0(scripts, "adjectives-preparation.R"))

#################### packages ####################
# Core tidyverse for wrangling + plotting; janitor/skimr for quick structure checks; patchwork for combining plots
library(tidyverse)
library(janitor)
library(skimr)
library(knitr)
library(cowplot)
library(ggdist)
library(readxl)
library(lubridate)
library(patchwork)
library(hexbin)
library(PRROC)

# Bayesian
library(brms);
library(loo);packageVersion("loo")
library(HDInterval);packageVersion("HDInterval")
library(cmdstanr);
library(emmeans);packageVersion("emmeans")
library(posterior);
library(bayesplot);
library(BayesFactor);
library(tidybayes);packageVersion("tidybayes")
library(bayestestR);
library(cmdstanr)
options(brms.backend = "cmdstanr") # Speed gain: ~30-50% faster than rstan
options(brms.file_refit = "on_change")  # only refit if model changes

# use all available cores for parallel computing
options(mc.cores = parallel::detectCores())

# Keep plots/readability consistent
theme_set(theme_minimal(base_size = 12) + panel_border())

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r echo = TRUE, message=FALSE, warning = FALSE}
# --- figure & table captions ---
outputFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to"); if (is.null(outputFormat)) outputFormat <- ""
capTabNo <- 1L; capFigNo <- 1L
.last_fig_no <- NA_integer_

capTab <- function(x){
  if (outputFormat == 'html'){
    x <- paste0("**Table ", capTabNo, ".** ", x)
    capTabNo <<- capTabNo + 1L
  }
  x
}

capFig <- function(x, show_R_version=FALSE, show_package_versions=NULL, is_map=FALSE){
  if (outputFormat == 'html'){
    assign(".last_fig_no", capFigNo, envir = .GlobalEnv) # remember current number
    x <- paste0("**Figure ", capFigNo, ".** ", x)
    capFigNo <<- capFigNo + 1L
  }
  x
}

# --- plot hook to auto-save ---
.default_plot_hook <- knitr::knit_hooks$get("plot")
.has_magick <- requireNamespace("magick", quietly = TRUE)

knitr::knit_hooks$set(plot = function(x, options){
  res <- if (is.function(.default_plot_hook)) .default_plot_hook(x, options) else NULL

  if (!is.null(options$fig.cap)) {
    fig_no <- get(".last_fig_no", envir = .GlobalEnv, inherits = FALSE)
    if (!is.na(fig_no)) {
      fig_no_str <- sprintf("%02d", fig_no)

      # subfolder = Rmd filename without extension
      rmd_name <- tools::file_path_sans_ext(basename(knitr::current_input()))
      target_dir <- file.path(plots, rmd_name)
      if (!dir.exists(target_dir)) dir.create(target_dir, recursive = TRUE)

      png_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".png"))
      tif_out <- file.path(target_dir, paste0("Figure_", fig_no_str, ".tif"))

      file.copy(x, png_out, overwrite = TRUE)

      if (.has_magick) {
        img <- magick::image_read(png_out)
        magick::image_write(img, path = tif_out, format = "tiff", compression = "lzw")
      } else {
        file.copy(png_out, tif_out, overwrite = TRUE)
      }
    }
  }

  res
})

```

Other helpers

```{r}
bb_corr <- function(x, y, R = 4000, seed = 1708) {
  stopifnot(length(x) == length(y))
  set.seed(seed)
  n <- length(x)
  out <- numeric(R)
  for (r in seq_len(R)) {
    w <- rexp(n)          # Dirichlet via i.i.d. Exponential then normalize
    w <- w / sum(w)
    mx <- sum(w * x); my <- sum(w * y)
    sx <- sqrt(sum(w * (x - mx)^2))
    sy <- sqrt(sum(w * (y - my)^2))
    out[r] <- sum(w * (x - mx) * (y - my)) / (sx * sy)
  }
  out
}

summ_post <- function(draws, name){
  h <- hdi(draws, ci = 0.95)
  tibble(pair = name,
         mean = mean(draws),
         median = median(draws),
         hdi_low = h$CI_low,
         hdi_high = h$CI_high)
}

bf_bic <- function(y, x){
  fit1 <- lm(y ~ x)
  fit0 <- lm(y ~ 1)
  BIC0 <- BIC(fit0); BIC1 <- BIC(fit1)
  # BF10 ≈ exp((BIC0 - BIC1)/2)
  exp((BIC0 - BIC1)/2)
}
```

## Load in data frames

```{r echo=TRUE, message=FALSE, warning=FALSE}
df    <- read_csv(paste0(data, "final_df.csv"))  %>% clean_names()
cats  <- read_csv(paste0(data, "meaning_categories.csv")) %>% clean_names()
```

# Correlations

```{r echo=TRUE, message=FALSE, warning=FALSE}
dat_cor <- df %>%
  select(cosine = cosine_rescaled, judged = ord_norm, binary = bin_f) %>%
  drop_na()
```

Pearson correlations (with CIs)

```{r echo=TRUE, message=FALSE, warning=FALSE}
pearson_mat <- cor(dat_cor, use = "pairwise.complete.obs", method = "pearson")
pearson_mat
```

Pairwise Pearson tests (p-values + CI)

```{r echo=TRUE, message=FALSE, warning=FALSE}
pearson_tests <- list(
  cosine_judged = cor.test(dat_cor$cosine, dat_cor$judged),
  cosine_binary = cor.test(dat_cor$cosine, dat_cor$binary),
  judged_binary = cor.test(dat_cor$judged, dat_cor$binary)
)
pearson_tests

pearson_tbl <- tibble(
  pair = c("cosine ~ judged", "cosine ~ binary", "judged ~ binary"),
  r = c(cor(dat_cor$cosine, dat_cor$judged),
        cor(dat_cor$cosine, dat_cor$binary),
        cor(dat_cor$judged, dat_cor$binary))
)

pearson_tbl
```

Bayesian correlations via BayesFactor (Point-biserial is just Pearson on
numeric 0/1; we still get Bayesian posterior for rho.)

```{r echo=TRUE, message=FALSE, warning=FALSE}
post_cos_jud <- bb_corr(dat_cor$cosine, dat_cor$judged, R = 8000, seed = 1708)
post_cos_bin <- bb_corr(dat_cor$cosine, dat_cor$binary, R = 8000, seed = 1708)
post_jud_bin <- bb_corr(dat_cor$judged, dat_cor$binary, R = 8000, seed = 1708)

bayes_corr_summary <- bind_rows(
  summ_post(post_cos_jud, "cosine ~ judged"),
  summ_post(post_cos_bin, "cosine ~ binary"),
  summ_post(post_jud_bin, "judged ~ binary")
)
bayes_corr_summary

# Posterior density plots
bind_rows(
  data.frame(rho = post_cos_jud, pair = "cosine~judged"),
  data.frame(rho = post_cos_bin, pair = "cosine~binary"),
  data.frame(rho = post_jud_bin, pair = "judged~binary")
) %>%
  ggplot(aes(rho)) + geom_density() + facet_wrap(~pair, scales = "free") +
  geom_vline(xintercept = 0, linetype = 2)
```

BIC-approx Bayes factors for "correlation != 0" For
continuous-continuous: linear model; for (cont \~ binary): also linear
is fine here.

```{r echo=TRUE, message=FALSE, warning=FALSE}
BF_cos_jud <- bf_bic(dat_cor$cosine, dat_cor$judged)
BF_cos_bin <- bf_bic(dat_cor$cosine, dat_cor$binary)
BF_jud_bin <- bf_bic(dat_cor$judged, dat_cor$binary)

bf_tbl <- tibble(
  pair = c("cosine ~ judged", "cosine ~ binary", "judged ~ binary"),
  BF10 = c(BF_cos_jud, BF_cos_bin, BF_jud_bin)
)
```

One clean summary table

```{r echo=TRUE, message=FALSE, warning=FALSE}
pearson_tbl %>%
  left_join(bayes_corr_summary, by = "pair") %>%
  left_join(bf_tbl, by = "pair")
```


# Modeling

## What we model

-   **Binary accuracy**: `binary_guess ∈ {0,1}`, modeled with a
    **Bernoulli** distribution and logit link.

-   **Ordinal judgments**: `judged_similarity ∈ {0,1,2,3}`, modeled with
    a **cumulative logit (ordinal)** likelihood.

-   **Cosine similarity**: `cosine_rescaled ∈ [0,1]`, modeled with a
    **zero-one-inflated Beta (ZOIB)** or a simpler **Beta** after
    applying an ε-open transform to handle boundary values.

All models include varying intercepts for participants and stimuli to
account for individual and item-level heterogeneity.

## Full (for Cognitive Science)

We fit baseline models for each of the three measures, with only
random intercepts for participant, native language, and stimulus.

```{r eval=TRUE, message=FALSE, warning=FALSE}
{
  set.seed(1708)

  ## ---- ensure models dir exists (you already have `models` as an absolute path) ----
  if (!dir.exists(models)) dir.create(models, recursive = TRUE, showWarnings = FALSE)

  ## ---- data prep (minimal; no z-scoring of cosine) ----
  ## ---------- COMPLETE-CASE DATA (based on cosine)
  # Rationale: use the same rows for all three outcomes to enable direct comparability
  if (!"ord_norm" %in% names(df)) df$ord_norm <- df$judged_similarity / 3
  df_brm <- subset(df, !is.na(cosine_rescaled))

  ## Keep exact-1 spike info for reporting (from original column but CC rows)
  ones_table_stim <- aggregate(I(df_brm$cosine_rescaled == 1) ~ stimulus_concept, data = df_brm, FUN = mean)
  names(ones_table_stim) <- c("stimulus_concept", "one_rate")

  df_brm$participant      <- factor(df_brm$participant)
  df_brm$stimulus_concept <- factor(df_brm$stimulus_concept)
  df_brm$native_language  <- factor(df_brm$native_language) # decided to do without it because it was not variable and caused much longer computation times
  df_brm$judged_ord       <- factor(df_brm$judged_similarity, levels = 0:3, ordered = TRUE)
 # Replace values extremely close to 1 with exactly 1
  df_brm$cosine_rescaled <- ifelse(
    !is.na(df_brm$cosine_rescaled) & df_brm$cosine_rescaled > 1 & df_brm$cosine_rescaled <
      1.001,
    1,
    df_brm$cosine_rescaled
  )
  
  ## ---- cosine preparation (rescaling to make comparable across measures) ----
  
  rescale_to_unit <- function(x) {
    eps <- 1e-10
    tiny <- 1e-12
    
    # Step 1: Epsilon-open to avoid exact boundaries
    x_open <- pmin(pmax(x - eps, tiny), 1 - tiny)
    
    # Step 2: Min-max rescale to [0,1]
    xmin <- min(x_open, na.rm = TRUE)
    xmax <- max(x_open, na.rm = TRUE)
    span <- xmax - xmin
    
    if (!is.finite(span) || span <= 0) {
      stop("Cannot rescale: no variation in x")
    }
    
    x_scaled <- (x_open - xmin) / span
    
    # Step 3: Ensure strictly in (0,1) for Beta
    x_final <- pmin(pmax(x_scaled, tiny), 1 - tiny)
    
    return(x_final)
  }
  
  # Apply to data
  # Nudge exact 1.0s only (keeping original scale)
  df_brm$cosine_original <- df_brm$cosine_rescaled
  df_brm$cosine_original[df_brm$cosine_original == 1.0] <- 0.999999
  
  # Full min-max rescaling (for comparable plot)
  df_brm$cosine_scaled <- rescale_to_unit(df_brm$cosine_rescaled)
  
  ## ---- very wide priors for baselines / couplings (not for ZOIB components) ----
  .pri_bin <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )
  .pri_ord <- c(
    set_prior("normal(0, 5)", class = "Intercept"),  # thresholds
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )
  .pri_wide <- c(
    set_prior("normal(0, 5)", class = "Intercept"),
    #set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 5)", class = "sd")
  )

  ## ========================= BASELINES =========================
  .bin_rds <- file.path(models, "fit_bin.rds")
  if (file.exists(.bin_rds)) {
    fit_bin <- readRDS(.bin_rds)
  } else {
    fit_bin <- brm(
      binary_guess ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = bernoulli(),
      prior   = .pri_bin,
      backend = "cmdstanr",
      chains  = 4, iter = 8000, warmup = 4000, 
      threads = threading(2), cores = parallel::detectCores(),
      control = list(adapt_delta = 0.97, max_treedepth = 12),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_bin, .bin_rds)
  }

  .ord_rds <- file.path(models, "fit_ord.rds")
  if (file.exists(.ord_rds)) {
    fit_ord <- readRDS(.ord_rds)
  } else {
    fit_ord <- brm(
      judged_ord ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = cumulative(link = "logit"),
      prior   = .pri_ord,
      backend = "cmdstanr",
      chains  = 4, iter = 10000, warmup = 4000, 
      threads = threading(2), cores = parallel::detectCores(),
      control = list(adapt_delta = 0.99, max_treedepth = 13),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_ord, .ord_rds)
  }

  .cos_org_rds <- file.path(models, "fit_cos_org.rds")
  if (file.exists(.cos_org_rds)) {
    fit_cos_org <- readRDS(.cos_org_rds)
  } else {
    fit_cos_org <- brm(
      cosine_original ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = Beta(), 
      prior = c(
        set_prior("normal(0, 2)", class = "Intercept"),
        set_prior("student_t(3, 0, 2.5)", class = "sd"),
        set_prior("gamma(0.01, 0.01)", class = "phi")),
      backend = "cmdstanr",
      chains  = 4, iter = 10000, warmup = 4000, 
      threads = threading(2), cores = parallel::detectCores(),
      control = list(adapt_delta = 0.99, max_treedepth = 13),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_cos_org, .cos_org_rds)
  }
  
  .cos_com_rds <- file.path(models, "fit_cos_com.rds")
  if (file.exists(.cos_com_rds)) {
    fit_cos_com <- readRDS(.cos_com_rds)
  } else {
    fit_cos_com <- brm(
      cosine_scaled ~ 1 + (1|p|participant) + (1|s|stimulus_concept),
      data    = df_brm,
      family  = Beta(), 
      prior = c(
        set_prior("normal(0, 2)", class = "Intercept"),
        set_prior("student_t(3, 0, 2.5)", class = "sd"),
        set_prior("gamma(0.01, 0.01)", class = "phi")),
      backend = "cmdstanr",
      chains  = 4, iter = 10000, warmup = 4000, 
      threads = threading(2), cores = parallel::detectCores(),
      control = list(adapt_delta = 0.99, max_treedepth = 13),
      seed    = 1708, refresh = 100,
      save_pars = save_pars(all = TRUE)
    )
    saveRDS(fit_cos_com, .cos_com_rds)
  }

  ## ---- compact Bayes R^2 (kept as r2_table) ----
  # Path to saved R2 summary
  .r2_rds <- file.path(models, "r2_table.rds")
  
  if (file.exists(.r2_rds)) {
    # If the summary table already exists, just load it
    r2_table <- readRDS(.r2_rds)
    message("Loaded existing r2_table from ", .r2_rds)
  
  } else {
    # Otherwise, compute it fresh from model fits
    .r2 <- function(fit) {
      r <- bayes_R2(fit)
      c(Estimate = mean(r),
        Q2.5 = quantile(r, .025),
        Q97.5 = quantile(r, .975))
    }
  
    r2_table <- data.frame(
      model = c("Binary (Bernoulli)", "Judged (Ordinal)", "Cosine (Beta)"),
      rbind(.r2(fit_bin), .r2(fit_ord), .r2(fit_cos_org))
    )
  
    # clean rounding
    r2_table$Estimate <- round(r2_table$Estimate, 3)
    r2_table$Q2.5     <- round(r2_table$Q2.5, 3)
    r2_table$Q97.5    <- round(r2_table$Q97.5, 3)
  
    # Save table to reuse later
    saveRDS(r2_table, .r2_rds)
    message("Saved new r2_table to ", .r2_rds)
  }
  

  ## ---- cleanup helpers (keep only model objects + r2_table + path vars) ----
  rm(list = c(".pri_bin",".pri_ord",".pri_wide",
              ".bin_rds",".ord_rds",".cos_org_rds", ".cos_com_rds",
              ".m_ord_cos_rds",".m_bin_cos_rds",".m_bin_jud_rds",".m_cos_jud_rds", ".r2_rds"))
}
```

## Extract results

### Convergence

Rhat and Bulk ESS:

```{r}
# ===== CONVERGENCE DIAGNOSTICS =====
# Extract Rhat and ESS for checking model quality

binary_diagnostics <- tibble(
  model = "binary",
  max_rhat = max(brms::rhat(fit_bin)),
  min_ess_bulk = min(brms::neff_ratio(fit_bin))
)

ordinal_diagnostics <- tibble(
  model = "ordinal",
  max_rhat = max(brms::rhat(fit_ord)),
  min_ess_bulk = min(brms::neff_ratio(fit_ord))
)

cosine_diagnostics <- tibble(
  model = "cosine",
  max_rhat = max(brms::rhat(fit_cos_com)),
  min_ess_bulk = min(brms::neff_ratio(fit_cos_com))
)

bind_rows(binary_diagnostics, ordinal_diagnostics, cosine_diagnostics)
```

Posterior predictive checks:

```{r}
pp_check(fit_bin, type = "bars", ndraws = 500)
pp_check(fit_ord, type = "bars", ndraws = 500)
pp_check(fit_cos_com, type = "dens_overlay", ndraws = 500)
```


R^2 for the models:

```{r}
r2_table
```


#### Cosine scale verification

```{r}
# Extract stimulus posteriors from both models
stim_original <- fitted(fit_cos_org, 
                       newdata = data.frame(stimulus_concept = levels(df_brm$stimulus_concept)),
                       re_formula = ~(1|stimulus_concept))

stim_scaled <- fitted(fit_cos_com,
                     newdata = data.frame(stimulus_concept = levels(df_brm$stimulus_concept)),
                     re_formula = ~(1|stimulus_concept))

# Perfect rank preservation
cor(stim_original[,"Estimate"], stim_scaled[,"Estimate"], method = "spearman")
# Should be as close to 1.0 as possible
```

Verify Rank Concordance

```{r}
# Extract stimulus-level predictions from both
get_stim_preds <- function(fit, label) {
  nd <- data.frame(stimulus_concept = levels(df_brm$stimulus_concept))
  preds <- fitted(fit, newdata = nd, 
                  re_formula = ~(1|stimulus_concept), 
                  summary = TRUE)
  tibble(
    stimulus = nd$stimulus_concept,
    mean = preds[, "Estimate"],
    model = label
  )
}

stim_both <- bind_rows(
  get_stim_preds(fit_cos_org, "Original scale"),
  get_stim_preds(fit_cos_com, "Rescaled")
)

# Rank correlation (should be perfect)
stim_wide <- stim_both %>%
  pivot_wider(names_from = model, values_from = mean)

cor(stim_wide$`Original scale`, stim_wide$`Rescaled`, method = "spearman")
# Should be 1.0 - perfect rank preservation

# Visualize
ggplot(stim_wide, aes(x = `Original scale`, y = `Rescaled`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Stimulus rankings preserved by rescaling")
```

### Model summaries

```{r}
# Binary model summary
summary(fit_bin)

binary_results <- tibble(
  parameter = c("Intercept", "SD_participant", "SD_stimulus"),
  mean = c(
    fixef(fit_bin)[1, "Estimate"],
    VarCorr(fit_bin)$participant$sd[1, "Estimate"],
    VarCorr(fit_bin)$stimulus_concept$sd[1, "Estimate"]
  ),
  ci_lower = c(
    fixef(fit_bin)[1, "Q2.5"],
    VarCorr(fit_bin)$participant$sd[1, "Q2.5"],
    VarCorr(fit_bin)$stimulus_concept$sd[1, "Q2.5"]
  ),
  ci_upper = c(
    fixef(fit_bin)[1, "Q97.5"],
    VarCorr(fit_bin)$participant$sd[1, "Q97.5"],
    VarCorr(fit_bin)$stimulus_concept$sd[1, "Q97.5"]
  )
)

binary_results
```

The Bayesian logistic regression model for binary accuracy (exact/near-exact matches) revealed that guessing success was very low overall. The population-level intercept of β = -7.11 (95% HDI [-8.65, -5.75]) on the logit scale corresponds to a baseline probability of approximately 0.08% (less than 1 in 1000) of producing an exact match when not accounting for specific stimuli or participants. This extremely negative intercept confirms that exact semantic matches were rare events. The model revealed substantial heterogeneity at both levels: participant-level variation (SD = 1.50, 95% HDI [1.29, 1.73]) indicated that some individuals were consistently better at guessing across all stimuli, while stimulus-level variation (SD = 3.48, 95% HDI [2.43, 5.02]) was even larger, demonstrating that certain concepts were dramatically more guessable than others. The large stimulus SD relative to the participant SD suggests that item characteristics (e.g., iconicity, semantic domain) played a more important role than individual differences in determining success.

```{r}
# Ordinal model summary
summary(fit_ord)

ordinal_results <- tibble(
  parameter = c(paste0("Threshold_", 0:2), "SD_participant", "SD_stimulus"),
  mean = c(
    fixef(fit_ord)[, "Estimate"],
    VarCorr(fit_ord)$participant$sd[1, "Estimate"],
    VarCorr(fit_ord)$stimulus_concept$sd[1, "Estimate"]
  ),
  ci_lower = c(
    fixef(fit_ord)[, "Q2.5"],
    VarCorr(fit_ord)$participant$sd[1, "Q2.5"],
    VarCorr(fit_ord)$stimulus_concept$sd[1, "Q2.5"]
  ),
  ci_upper = c(
    fixef(fit_ord)[, "Q97.5"],
    VarCorr(fit_ord)$participant$sd[1, "Q97.5"],
    VarCorr(fit_ord)$stimulus_concept$sd[1, "Q97.5"]
  )
)

ordinal_results
```

The cumulative logit model for ordinal judgments (0 = completely different, 1 = distantly related, 2 = similar, 3 = identical/near-identical) estimated three threshold parameters that define boundaries between adjacent response categories. The first threshold (between 0 and 1) was β = 3.90 (95% HDI [2.80, 5.03]), indicating a high barrier for responses to be considered even distantly related to targets. The second threshold (between 1 and 2) was β = 4.88 (95% HDI [3.78, 6.01]), and the third threshold (between 2 and 3) was β = 6.19 (95% HDI [5.09, 7.32]). These monotonically increasing thresholds show that each step up in similarity rating required increasingly strong evidence, with the gap between categories 2 and 3 (similar vs. identical) being particularly large. Participant-level variation was moderate (SD = 0.665, 95% HDI [0.580, 0.762]), suggesting relatively consistent rating behavior across individuals. However, stimulus-level variation was substantial (SD = 3.08, 95% HDI [2.28, 4.23]), confirming that some stimuli consistently elicited higher similarity ratings than others. The pattern mirrors the binary model: item characteristics mattered more than individual differences in determining semantic overlap between responses and targets.

```{r}
# Cosine model summary
summary(fit_cos_com)

cosine_results <- tibble(
  parameter = c("Intercept_mu", "Phi", "SD_participant", "SD_stimulus"),
  mean = c(
    fixef(fit_cos_com)[1, "Estimate"],
    summary(fit_cos_com)$spec_pars[1, "Estimate"], # phi parameter
    VarCorr(fit_cos_com)$participant$sd[1, "Estimate"],
    VarCorr(fit_cos_com)$stimulus_concept$sd[1, "Estimate"]
  ),
  ci_lower = c(
    fixef(fit_cos_com)[1, "Q2.5"],
    NA,
    VarCorr(fit_cos_com)$participant$sd[1, "Q2.5"],
    VarCorr(fit_cos_com)$stimulus_concept$sd[1, "Q2.5"]
  ),
  ci_upper = c(
    fixef(fit_cos_com)[1, "Q97.5"],
    NA,
    VarCorr(fit_cos_com)$participant$sd[1, "Q97.5"],
    VarCorr(fit_cos_com)$stimulus_concept$sd[1, "Q97.5"]
  )
)

cosine_results
```

The Beta regression model for cosine similarity scores (computational semantic overlap based on ConceptNet embeddings, scaled 0-1) estimated a population-level intercept of β = -0.534 (95% HDI [-0.863, -0.217]) on the logit scale. When back-transformed to the probability scale, this corresponds to a mean cosine similarity of approximately 0.37, indicating that participant responses were, on average, semantically closer to unrelated concepts than to the target meanings. The precision parameter φ = 2.75 describes the concentration of the Beta distribution, with higher values indicating less variance around the mean; this moderate value suggests considerable spread in cosine scores. Between-participant variation was relatively small (SD = 0.346, 95% HDI [0.311, 0.386]), indicating that participants showed similar levels of semantic distance from targets on average. In contrast, between-stimulus variation was substantially larger (SD = 0.878, 95% HDI [0.676, 1.16]), confirming that the semantic guessability of stimuli varied dramatically. As with the other two measures, this pattern demonstrates that the success or failure of semantic guessing was primarily driven by properties of the stimuli (e.g., which semantic domain they belonged to, their acoustic iconicity) rather than by individual participant abilities.

### Compare outcome scales (pasted from claude just like that)

Binary model:   P(Y = 1)              → [0, 1] probability
Cosine model:   E(Y)                  → [0, 1] continuous
Ordinal model:  P(Y = 0,1,2,3)        → How to map to [0,1]?

Ordinal to continuous mapping

```{r}
# Helper function: extract expected value from ordinal model
get_ordinal_expected <- function(fit_ord, newdata) {
  # Get predicted probabilities for each category
  # fitted() for ordinal gives P(Y = k) for each k
  pred <- fitted(fit_ord, newdata = newdata, 
                 re_formula = ~(1|stimulus_concept),
                 summary = FALSE)  # Get all posterior draws
  
  # pred is array: draws × observations × categories
  # Dimensions: [n_draws, n_stimuli, 4 categories]
  
  # Compute expected value for each draw and stimulus
  # E(Y) = 0*P(Y=0) + 1*P(Y=1) + 2*P(Y=2) + 3*P(Y=3)
  expected_vals <- apply(pred, c(1, 2), function(probs) {
    sum(c(0, 1, 2, 3) * probs)
  })
  
   
  # Normalize to [0, 1]
  expected_vals_norm <- expected_vals / 3
  
  # Return summary statistics
  tibble(
    stimulus = newdata$stimulus_concept,
    Estimate = apply(expected_vals_norm, 2, mean),
    Q2.5 = apply(expected_vals_norm, 2, quantile, 0.025),
    Q97.5 = apply(expected_vals_norm, 2, quantile, 0.975)
    )
  }
  
  # Create newdata for all stimuli
  nd_stim <- data.frame(stimulus_concept = levels(df_brm$stimulus_concept))
  
  # Extract predictions from all three models
  stim_binary <- fitted(fit_bin, newdata = nd_stim,
                        re_formula = ~(1|stimulus_concept),
                        summary = TRUE) %>%
    as_tibble() %>%
    mutate(stimulus = nd_stim$stimulus_concept,
           measure = "Binary")
  
  stim_ordinal <- get_ordinal_expected(fit_ord, nd_stim) %>%
    mutate(measure = "Judged")
  
  stim_cosine <- fitted(fit_cos_com, newdata = nd_stim,
                        re_formula = ~(1|stimulus_concept),
                        summary = TRUE) %>%
    as_tibble() %>%
    mutate(stimulus = nd_stim$stimulus_concept,
           measure = "Cosine (rescaled)")
  
  # Combine all three
  all_measures <- bind_rows(
    stim_binary %>% select(stimulus, Estimate, Q2.5, Q97.5, measure),
    stim_ordinal %>% select(stimulus, Estimate, Q2.5, Q97.5, measure),
    stim_cosine %>% select(stimulus, Estimate, Q2.5, Q97.5, measure)
  ) %>%
    left_join(cats, by = c("stimulus" = "meaning"))
```

Across all three evaluation measures—binary accuracy, ordinal similarity judgments, and computational cosine similarity—a consistent pattern emerged: stimulus-level variation substantially exceeded participant-level variation. This indicates that the guessability of novel vocalizations and ape gestures was primarily determined by properties of the signals themselves (their iconicity, semantic domain, or other stimulus characteristics) rather than by individual cognitive differences among participants. The extremely low intercepts for binary accuracy and the below-midpoint intercept for cosine similarity both confirm that, in absolute terms, semantic guessing success was limited when the response space was unrestricted, even though relative differences between stimuli were substantial and systematic.

### Visualizing data

Visualization 1: Heatmap

```{r heatmap, echo=FALSE, fig.align='left', fig.cap=capFig("Heatmap plot."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
ggplot(all_measures, aes(x = measure, y = stimulus, fill = Estimate)) +
  geom_tile(color = "white") +
  facet_grid(category ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(limits = c(0, 1), 
                       name = "μ [0,1]",
                       option = "plasma") +
  labs(x = NULL, y = NULL,
       title = "Stimulus success across evaluation lenses",
       subtitle = "All measures normalized to [0,1] scale") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

Visualization 2: Lollipop Plot

```{r lollipop, echo=FALSE, fig.align='left', fig.cap=capFig("Lolipop plot."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
# Reorder stimuli by cosine performance for readability
stim_order <- all_measures %>%
  filter(measure == "Cosine (rescaled)") %>%
  arrange(Estimate) %>%
  pull(stimulus)

all_measures <- all_measures %>%
  mutate(stimulus = factor(stimulus, levels = stim_order),
         measure = factor(measure, 
                         levels = c("Binary", "Judged", "Cosine (rescaled)")))

ggplot(all_measures, aes(x = Estimate, y = stimulus, color = measure)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  geom_errorbarh(aes(xmin = Q2.5, xmax = Q97.5), 
                 height = 0.2,
                 position = position_dodge(width = 0.5)) +
  facet_wrap(~ category, scales = "free_y", ncol = 2) +
  scale_color_brewer(palette = "Dark2", name = "Evaluation") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Posterior mean μ [0,1]", y = NULL) +
  theme_minimal(base_size = 16)

# Save as PNG
ggsave(filename = paste0(plots,"lollipop.png"), 
       plot = last_plot(), 
       width = 8, 
       height = 5, 
       dpi = 300)

# Save as TIF
ggsave(filename =  paste0(plots,"lollipop.tif"), 
       plot = last_plot(), 
       width = 8, 
       height = 5, 
       dpi = 300)
```

Visualization 3: "Progressive Inclusivity" Plot
This directly shows your research question:

```{r inclusivity, echo=FALSE, fig.align='left', fig.cap=capFig("Progressive inclusivity plot."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
# Count how many stimuli "succeed" under each threshold
thresholds <- seq(0.1, 0.9, 0.1)

success_counts <- map_df(thresholds, function(thresh) {
  all_measures %>%
    group_by(measure) %>%
    summarise(
      threshold = thresh,
      n_above = sum(Estimate > thresh),
      prop_above = mean(Estimate > thresh)
    )
})

ggplot(success_counts, aes(x = threshold, y = n_above, color = measure)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Dark2", name = "Evaluation") +
  scale_x_continuous(breaks = thresholds) +
  labs(x = "Success threshold (μ)", 
       y = "Number of stimuli above threshold") +
  theme_minimal(base_size = 16)

# Save as PNG
ggsave(filename = paste0(plots,"inclusivity.png"), 
       plot = last_plot(), 
       width = 6, 
       height = 4, 
       dpi = 300)

# Save as TIF
ggsave(filename =  paste0(plots,"inclusivity.tif"), 
       plot = last_plot(), 
       width = 6, 
       height = 4, 
       dpi = 300)
```

Visualization 4: Direct Comparison Grid

```{r comparison_grid, echo=FALSE, fig.align='left', fig.cap=capFig("Direct comparison grid."), fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
# Advantage plot adapted for all three measures
all_measures_wide <- all_measures %>%
  select(stimulus, subcategory, measure, Estimate) %>%
  pivot_wider(names_from = measure, values_from = Estimate)

# Compute gaps
all_measures_wide <- all_measures_wide %>%
  mutate(
    `Judged - Binary` = Judged - Binary,
    `Cosine - Judged` = `Cosine (rescaled)` - Judged,
    `Cosine - Binary` = `Cosine (rescaled)` - Binary
  )

# Aggregate by subcategory
subcat_gaps <- all_measures_wide %>%
  group_by(subcategory) %>%
  summarise(
    n = n(),
    across(c(`Judged - Binary`, `Cosine - Judged`, `Cosine - Binary`), 
           mean, .names = "{.col}_mean")
  )

# Plot
subcat_gaps %>%
  pivot_longer(cols = ends_with("_mean"), 
               names_to = "contrast", 
               values_to = "gap") %>%
  mutate(contrast = str_remove(contrast, "_mean"),
         subcategory = fct_reorder(subcategory, gap, .fun = max)) %>%
  ggplot(aes(x = gap, y = subcategory, color = contrast)) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_point(aes(size = n), position = position_dodge(width = 0.4)) +
  geom_segment(aes(x = 0, xend = gap, y = subcategory, yend = subcategory),
               position = position_dodge(width = 0.4),
               linewidth = 0.8) +
  scale_color_brewer(palette = "Dark2", name = "Contrast") +
  scale_size_continuous(range = c(2, 6), name = "N stimuli") +
  labs(x = "Gap in μ (positive = second measure more permissive)",
       y = NULL)
```

Critical Code Block: Getting Ordinal Expected Values
If the above get_ordinal_expected() function doesn't work with your brms version, here's a more robust alternative:

```{r}
# Alternative method using posterior_predict
get_ordinal_expected_v2 <- function(fit_ord, newdata) {
  # Get posterior predicted categories for each draw
  pred_cats <- posterior_predict(fit_ord, 
                                 newdata = newdata,
                                 re_formula = ~(1|stimulus_concept))
  
  # pred_cats is matrix: draws × observations
  # Values are 0, 1, 2, or 3
  
  # Convert categories to [0,1] scale
  pred_norm <- pred_cats / 3
  
  # Summarize across draws
  tibble(
    stimulus = newdata$stimulus_concept,
    Estimate = apply(pred_norm, 2, mean),
    Q2.5 = apply(pred_norm, 2, quantile, 0.025),
    Q97.5 = apply(pred_norm, 2, quantile, 0.975)
  )
}

# Test both methods and compare
stim_ord_v1 <- get_ordinal_expected(fit_ord, nd_stim)
stim_ord_v2 <- get_ordinal_expected_v2(fit_ord, nd_stim)

# Should be highly correlated (might differ slightly)
cor(stim_ord_v1$Estimate, stim_ord_v2$Estimate)
```

# Descriptive statistics

```{r}
# Overall descriptive statistics
desc_stats_overall <- df_brm %>%
  summarise(
    # Binary
    binary_mean = mean(binary_guess, na.rm = TRUE),
    binary_sd = sd(binary_guess, na.rm = TRUE),
    binary_n = sum(!is.na(binary_guess)),
    
    # Ordinal (judged similarity)
    judged_mean = mean(judged_similarity, na.rm = TRUE),
    judged_sd = sd(judged_similarity, na.rm = TRUE),
    judged_median = median(judged_similarity, na.rm = TRUE),
    judged_n = sum(!is.na(judged_similarity)),
    
    # Ordinal normalized
    judged_norm_mean = mean(ord_norm, na.rm = TRUE),
    judged_norm_sd = sd(ord_norm, na.rm = TRUE),
    judged_norm_median = median(ord_norm, na.rm = TRUE),
    judged_norm_n = sum(!is.na(ord_norm)),
    
    # Cosine
    cosine_mean = mean(cosine_rescaled, na.rm = TRUE),
    cosine_sd = sd(cosine_rescaled, na.rm = TRUE),
    cosine_median = median(cosine_rescaled, na.rm = TRUE),
    cosine_n = sum(!is.na(cosine_rescaled)),
    
    # Cosine scaled
    cosine_sc_mean = mean(cosine_scaled, na.rm = TRUE),
    cosine_sc_sd = sd(cosine_scaled, na.rm = TRUE),
    cosine_sc_median = median(cosine_scaled, na.rm = TRUE),
    cosine_sc_n = sum(!is.na(cosine_scaled))
  )

desc_stats_overall

# Save for manuscript
#write_csv(desc_stats_overall, paste0(data, "descriptive_overall.csv"))
```


```{r}
# Descriptive statistics by semantic category
desc_stats_category <- df_brm %>%
  left_join(cats, by = c("stimulus_concept" = "meaning")) %>%
  group_by(category, subcategory) %>%
  summarise(
    n_stimuli = n_distinct(stimulus_concept),
    n_responses = n(),
    
    # Binary
    binary_mean = mean(binary_guess, na.rm = TRUE),
    binary_sd = sd(binary_guess, na.rm = TRUE),
    binary_prop_correct = mean(binary_guess, na.rm = TRUE), # proportion correct
    
    # Ordinal
    judged_mean = mean(judged_similarity, na.rm = TRUE),
    judged_sd = sd(judged_similarity, na.rm = TRUE),
    judged_median = median(judged_similarity, na.rm = TRUE),
    judged_prop_2plus = mean(judged_similarity >= 2, na.rm = TRUE), # proportion rated similar or exact
    
    # Ordinal normalized
    judged_norm_mean = mean(ord_norm, na.rm = TRUE),
    judged_norm_sd = sd(ord_norm, na.rm = TRUE),
    judged_norm_median = median(ord_norm, na.rm = TRUE),
    
    # Cosine
    cosine_mean = mean(cosine_rescaled, na.rm = TRUE),
    cosine_sd = sd(cosine_rescaled, na.rm = TRUE),
    cosine_median = median(cosine_rescaled, na.rm = TRUE),
    
    # Cosine scaled
    cosine_sc_mean = mean(cosine_scaled, na.rm = TRUE),
    cosine_sc_sd = sd(cosine_scaled, na.rm = TRUE),
    cosine_sc_median = median(cosine_scaled, na.rm = TRUE),
    
    .groups = "drop"
  )

desc_stats_category %>% print(n = Inf)

# write_csv(desc_stats_category, paste0(data, "descriptive_by_category.csv"))
```


```{r}
# Descriptive statistics by individual stimulus
desc_stats_stimulus <- df_brm %>%
  left_join(cats, by = c("stimulus_concept" = "meaning")) %>%
  group_by(stimulus_concept, category, subcategory) %>%
  summarise(
    n_responses = n(),
    
    # Binary
    binary_mean = mean(binary_guess, na.rm = TRUE),
    binary_n_correct = sum(binary_guess == 1, na.rm = TRUE),
    
    # Ordinal
    judged_mean = mean(judged_similarity, na.rm = TRUE),
    judged_sd = sd(judged_similarity, na.rm = TRUE),
    judged_median = median(judged_similarity, na.rm = TRUE),
    
    # Ordinal normalized
    judged_norm_mean = mean(ord_norm, na.rm = TRUE),
    judged_norm_sd = sd(ord_norm, na.rm = TRUE),
    judged_norm_median = median(ord_norm, na.rm = TRUE),
    
    # Distribution across ordinal categories
    judged_prop_0 = mean(judged_similarity == 0, na.rm = TRUE),
    judged_prop_1 = mean(judged_similarity == 1, na.rm = TRUE),
    judged_prop_2 = mean(judged_similarity == 2, na.rm = TRUE),
    judged_prop_3 = mean(judged_similarity == 3, na.rm = TRUE),
    
    # Cosine
    cosine_mean = mean(cosine_rescaled, na.rm = TRUE),
    cosine_sd = sd(cosine_rescaled, na.rm = TRUE),
    cosine_median = median(cosine_rescaled, na.rm = TRUE),
    cosine_q25 = quantile(cosine_rescaled, 0.25, na.rm = TRUE),
    cosine_q75 = quantile(cosine_rescaled, 0.75, na.rm = TRUE),
    
    # Cosine scaled
    cosine_sc_mean = mean(cosine_scaled, na.rm = TRUE),
    cosine_sc_sd = sd(cosine_scaled, na.rm = TRUE),
    cosine_sc_median = median(cosine_scaled, na.rm = TRUE),
    cosine_sc_q25 = quantile(cosine_scaled, 0.25, na.rm = TRUE),
    cosine_sc_q75 = quantile(cosine_scaled, 0.75, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  arrange(desc(cosine_mean))

desc_stats_stimulus %>% print(n = Inf)

#write_csv(desc_stats_stimulus, paste0(data, "descriptive_by_stimulus.csv"))
```


```{r}
# Create publication-ready table combining all measures
desc_table_combined <- desc_stats_stimulus %>%
  select(stimulus_concept, category, n_responses,
         binary_mean, judged_mean, judged_median, 
         judged_norm_mean, judged_norm_median,
         cosine_mean, cosine_median,
         cosine_sc_mean, cosine_sc_median) %>%
  mutate(across(where(is.numeric) & !n_responses, ~round(., 3)))

desc_table_combined %>% print(n = Inf)

#write_csv(desc_table_combined, paste0(data, "table_descriptive_combined.csv"))
```

Table 1. Correlations between evaluation measures. Pearson correlations with Bayesian posterior estimates and 95% highest density intervals. BF₁₀ indicates Bayes factor for correlation ≠ 0. Ordinal judgments were normalized to 0-1 scale for comparability.

```{r}
# Generate Table 1: Correlation table for manuscript
correlation_table_manuscript <- tibble(
  Measure_Pair = c("Cosine ~ Ordinal ", 
                   "Cosine ~ Binary", 
                   "Ordinal ~ Binary"),
  r = c(0.833, 0.682, 0.796),
  HDI_Lower = c(0.826, 0.670, 0.786),
  HDI_Upper = c(0.839, 0.693, 0.806),
  BF10 = c(">10³⁰", ">10³⁰", ">10³⁰")
)

write_csv(correlation_table_manuscript, 
          paste0(data, "table1_correlations.csv"))

# For inline display:
kable(correlation_table_manuscript, 
      digits = 3,
      col.names = c("Measure Pair", "r", "95% HDI Lower", "95% HDI Upper", "BF₁₀"),
      caption = "Table 1. Correlations between evaluation measures")
```

# Session info

```{r echo=TRUE, message=FALSE, warning=FALSE}
sessionInfo()
```
